---
title: "Bayesian Small Area Estimation of Under-5 Malaria Prevalence in West Africa"
subtitle: "A Space-Time BYM2 Fay-Herriot Model Across Ghana, Burkina Faso, and Côte d'Ivoire"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
bibliography: references.bib
link-citations: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6)
# Core libraries
library(tidyverse)
library(haven)
library(sf)
library(geodata)
library(survey)
library(spdep)
library(Matrix)
library(cmdstanr)
library(bayesplot)
library(tidybayes)
library(patchwork)
library(ggcorrplot)
library(posterior)
```

# About the Data

## Demographic and Health Surveys (DHS)

The data for this project is sourced from the Demographic and Health Surveys (DHS) program [@croft2018], which provides nationally representative datasets on health and population indicators. The specific datasets used in this analysis include two DHS rounds for each of the three selected countries: Ghana, Burkina Faso, and Côte d'Ivoire. For each survey round, two files are used: the **household member recode (PR) file**, containing individual-level data including the malaria biomarker, and the **GPS shapefile**, providing geospatial coordinates for the survey clusters.

```
Ghana:           DHS 2014 (GHPR72FL) and DHS 2022 (GHPR8CFL)
Burkina Faso:    DHS 2010 (BFPR62FL) and DHS 2021 (BFPR81FL)
Côte d'Ivoire:   DHS 2012 (CIPR62FL) and DHS 2021 (CIPR81FL)
```

## Outcome Variable: Microscopy-Confirmed Malaria

We use microscopy-confirmed malaria infection (`hml32`) in children aged 6-59 months. While DHS also reports RDT results, microscopy is the recommended gold standard for prevalence estimation in survey-based studies [@florey2014]. Our direct estimates were validated against the published DHS report (26.7% in 2014 and 8.6% in 2022 for Ghana national prevalence), confirming correct variable identification and weighting procedure.

# Data Loading and Cleaning

```{r survey-index}
survey_index <- tribble(
  ~country,        ~iso, ~year, ~round_id, ~pr_path, ~gps_path,
  "Ghana",         "GHA", 2014, 1L,
  "dhs_gh_14/GHPR72DT/GHPR72FL.DTA", "dhs_gh_14/GHGE71FL/GHGE71FL.shp",
  "Ghana",         "GHA", 2022, 2L,
  "dhs_gh_22/GHPR8CDT/GHPR8CFL.DTA", "dhs_gh_22/GHGE8AFL/GHGE8AFL.shp",
  "Burkina",       "BFA", 2010, 1L,
  "dhs_bf_10/BFPR62DT/BFPR62FL.DTA", "dhs_bf_10/BFGE61FL/BFGE61FL.shp",
  "Burkina",       "BFA", 2021, 2L,
  "dhs_bf_21/BFPR81DT/BFPR81FL.DTA", "dhs_bf_21/BFGE81FL/BFGE81FL.shp",
  "Cote d'Ivoire", "CIV", 2012, 1L,
  "dhs_ci_12/CIPR62DT/CIPR62FL.DTA", "dhs_ci_12/CIGE61FL/CIGE61FL.shp",
  "Cote d'Ivoire", "CIV", 2021, 2L,
  "dhs_ci_21/CIPR81DT/CIPR81FL.DTA", "dhs_ci_21/CIGE81FL/CIGE81FL.shp"
)
```

## District Boundaries

District boundaries are obtained from the GADM database. We use Level 2 for Ghana (260 districts) and Level 3 for Burkina Faso (351 departments) and Côte d'Ivoire (113 departments). Boundaries are transformed to WGS 84 (EPSG:4326) to match the DHS GPS data.

```{r district-boundaries}
load_districts <- function(iso) {
  if (iso == "GHA") {
    gadm(country = iso, level = 2, path = "boundaries/") %>%
      st_as_sf() %>% st_transform(crs = 4326) %>%
      rename(district_name = NAME_2, district_code = GID_2) %>%
      mutate(iso = iso)
  } else {
    gadm(country = iso, level = 3, path = "boundaries/") %>%
      st_as_sf() %>% st_transform(crs = 4326) %>%
      rename(district_name = NAME_3, district_code = GID_3) %>%
      mutate(iso = iso)
  }
}

districts_raw <- list(
  GHA = load_districts("GHA"),
  BFA = load_districts("BFA"),
  CIV = load_districts("CIV")
)

# Create numeric district lookups (Stan requires integer indices)
district_lookups <- map(districts_raw, function(d) {
  d %>% st_drop_geometry() %>%
    select(district_name, district_code, iso) %>%
    mutate(district_id = row_number())
})

# Districts ordered by district_id for adjacency matrix construction
districts_ordered <- map2(districts_raw, district_lookups, function(d, lk) {
  d %>% left_join(lk, by = c("district_name", "district_code", "iso")) %>%
    arrange(district_id)
})

map_dfr(district_lookups, ~tibble(districts = nrow(.x)), .id = "iso")
```

## PR Files and Spatial Join

```{r pr-files}
load_pr <- function(pr_path, gps_path, iso, year, round_id,
                    districts_ordered, district_lookup) {
  gps <- read_sf(gps_path) %>%
    rename(cluster_id = DHSCLUST) %>%
    st_transform(crs = 4326)
  
  gps_joined <- st_join(gps, districts_ordered, join = st_nearest_feature) %>%
    select(cluster_id, district_name, district_code) %>%
    st_drop_geometry() %>%
    left_join(district_lookup, by = c("district_name", "district_code")) %>%
    select(cluster_id, district_id, district_name)
  
  pr <- read_dta(pr_path) %>%
    rename_with(tolower) %>%
    select(cluster_id = hv001, hh_id = hv002, line_id = hvidx,
           weight_raw = hv005, stratum = hv023,
           age_months = hc1, malaria = hml32) %>%
    filter(age_months < 60, malaria %in% c(0, 1)) %>%
    mutate(weight = weight_raw / 1e6,
           malaria = as.integer(malaria),
           cluster_id = as.integer(cluster_id),
           stratum = as.integer(stratum),
           country = iso, year = year, round_id = round_id) %>%
    left_join(gps_joined, by = "cluster_id")
  
  pr
}

pr_list <- pmap(survey_index, function(country, iso, year, round_id, pr_path, gps_path) {
  load_pr(pr_path, gps_path, iso, year, round_id,
          districts_ordered[[iso]], district_lookups[[iso]])
})

```

# Direct Estimation

For each observed district-year, we compute weighted prevalence using the Hájek ratio estimator [@hajek1971]:

$$\hat{p}_{it}^w = \frac{\sum_k d_k y_k}{\sum_k d_k}$$

where $d_k$ is the DHS sampling weight and $y_k$ is the binary malaria indicator. Design-based standard errors are obtained via Taylor-series linearisation [@binder1983] accounting for the two-stage stratified cluster design, implemented via the R `survey` package [@lumley2010].

To use these estimates in a Fay-Herriot framework [@wakefield2020] with a Gaussian likelihood, we transform to the logit scale and apply the delta method for the standard error:

$$y_{it} = \text{logit}(\hat{p}_{it}^w), \quad \widehat{\text{SE}}(y_{it}) = \frac{\widehat{\text{SE}}(\hat{p}_{it}^w)}{\hat{p}_{it}^w(1 - \hat{p}_{it}^w)}$$

For numerical stability, prevalences are bounded at $[0.02, 0.98]$ before logit transformation, and logit-scale standard errors $\psi_{it}$ are bounded at $[0.1, 2.0]$ to prevent degenerate likelihood contributions in extreme cases [@wakefield2020].

```{r direct-estimates}
get_direct_estimates <- function(pr) {
  design <- svydesign(
    ids = ~cluster_id, strata = ~stratum, weights = ~weight,
    data = pr, nest = TRUE
  )
  svyby(formula = ~malaria, by = ~district_id, design = design,
        FUN = svymean, na.rm = TRUE, keep.var = TRUE) %>%
    rename(p_hat = malaria, se_p = se) %>%
    filter(!is.na(p_hat), !is.na(se_p),
           se_p > 0, p_hat > 0, p_hat < 1) %>%
    mutate(
      country   = pr$country[1],
      iso       = pr$country[1],
      year      = pr$year[1],
      round_id  = pr$round_id[1],
      p_bounded = pmax(0.02, pmin(0.98, p_hat)),
      y_i       = log(p_bounded / (1 - p_bounded)),
      logit_var = se_p^2 / (pmax(0.02, p_hat)^2 * pmax(0.02, 1 - p_hat)^2),
      psi_i     = pmax(0.1, pmin(sqrt(logit_var), 2.0))
    ) %>%
    filter(is.finite(y_i), is.finite(psi_i)) %>%
    select(country, iso, year, round_id, district_id, y_i, psi_i)
}

direct_est_list <- map(pr_list, get_direct_estimates)
direct_est_all  <- bind_rows(direct_est_list)

direct_est_all %>% count(country, year, name = "observed_districts")

```

# Panel Construction

A balanced panel is constructed by taking the Cartesian product of districts and survey rounds for each country. Observed district-years receive their direct estimates; unobserved cells remain `NA` with an explicit `observed = 0` flag. This structure enables the Bayesian model to predict prevalence for unobserved districts through spatial borrowing, while applying the likelihood only to observed cells.

```{r build-panel}
build_panel <- function(country_iso, district_lookup, direct_est_all) {
  rounds <- direct_est_all %>%
    filter(country == country_iso) %>%
    distinct(year) %>% arrange(year) %>%
    mutate(round_id = row_number())
  
  est_clean <- direct_est_all %>%
    filter(country == country_iso) %>%
    select(district_id, year, y_i, psi_i)
  
  expand_grid(
    district_id = district_lookup$district_id,
    year        = rounds$year
  ) %>%
    left_join(district_lookup %>% select(district_id, district_name, district_code),
              by = "district_id") %>%
    left_join(est_clean, by = c("district_id", "year")) %>%
    left_join(rounds, by = "year") %>%
    arrange(year, district_id) %>%
    mutate(country = country_iso,
           observed = as.integer(!is.na(y_i)))
}

panels <- map(c("GHA", "BFA", "CIV"),
              ~build_panel(.x, district_lookups[[.x]], direct_est_all))
names(panels) <- c("GHA", "BFA", "CIV")

map_dfr(panels, function(p) {
  p %>% group_by(year) %>%
    summarise(total = n(), observed = sum(observed == 1),
              pct = round(100 * observed / total, 1), .groups = "drop")
}, .id = "country")
```

# Spatial Adjacency and BYM2 Scaling Factor

District adjacency is defined using queen contiguity [@bivand2013]. Islands are reconnected using a snap tolerance of 0.1 degrees to ensure a fully connected graph, a necessary condition for a proper ICAR prior [@freni2018].

From the binary adjacency matrix $\mathbf{W}$, we construct the ICAR precision matrix $\mathbf{Q}_{\text{ICAR}} = \mathbf{D} - \mathbf{W}$, where $\mathbf{D}$ is the diagonal matrix of neighbour counts [@besag1991]. Because $\mathbf{Q}_{\text{ICAR}}$ is rank-deficient by one, its marginal variances depend on graph structure. Following @riebler2016, we compute the BYM2 scaling factor $s$ as the geometric mean of the marginal variances:

$$s = \exp\!\left(\frac{1}{N}\sum_{i=1}^{N} \log [\mathbf{Q}_{\text{ICAR}}^{-}]_{ii}\right)$$

Dividing the ICAR component by $\sqrt{s}$ in the BYM2 formula ensures its geometric-mean marginal variance equals one, making $\phi$ interpretable as the proportion of district-level variance that is spatially structured.

```{r adjacency-scaling}
compute_scaling_factor <- function(nb, D) {
  adj_mat <- nb2mat(nb, style = "B", zero.policy = TRUE)
  Q_icar  <- Diagonal(x = rowSums(adj_mat)) - as(adj_mat, "sparseMatrix")
  eps <- 0.001
  Q_reg <- Q_icar + eps * Diagonal(D)
  Q_inv_diag <- diag(solve(Q_reg))
  exp(mean(log(Q_inv_diag)))
}

build_spatial <- function(iso, districts_ordered) {
  D <- nrow(districts_ordered)
  nb <- poly2nb(districts_ordered, queen = TRUE)
  if (sum(sapply(nb, length) == 0) > 0) {
    nb <- poly2nb(districts_ordered, queen = TRUE, snap = 0.1)
  }
  
  adj_mat <- nb2mat(nb, style = "B", zero.policy = TRUE)
  edges <- which(adj_mat == 1, arr.ind = TRUE) %>%
    as.data.frame() %>%
    filter(row < col)
  
  scaling_factor <- compute_scaling_factor(nb, D)
  
  list(D = D, nb = nb,
       node1 = edges$row, node2 = edges$col,
       N_edges = nrow(edges),
       scaling_factor = scaling_factor)
}

spatial <- map(c("GHA", "BFA", "CIV"),
               ~build_spatial(.x, districts_ordered[[.x]]))
names(spatial) <- c("GHA", "BFA", "CIV")

map_dfr(spatial, function(s) {
  tibble(N = s$D, N_edges = s$N_edges, scaling_factor = round(s$scaling_factor, 3))
}, .id = "country")
```

# Covariates: Harmonised Multi-Country Panel

We use a harmonised covariate panel covering all three countries and both survey rounds, with seven standardised covariates derived from public geospatial sources. All covariates were standardized (mean 0, SD 1) to make coefficients directly comparable as per-standard-deviation effects, to allow a single weakly-informative prior $\beta_k \sim \mathcal{N}(0, 1)$ to apply to all coefficients, and to improve Hamiltonian Monte Carlo sampling efficiency [@stan2024].

**Static covariates** (one value per district, applied to both rounds):

- `elev_std` Mean elevation (SRTM)
- `travel_std`  Travel time to nearest healthcare facility [@weiss2020]
- `pop_dens_std`  Log-transformed population density (WorldPop)
- `poverty_std`  Multidimensional poverty headcount

**Time-varying covariates** (per district, per round):

- `rainfall_std` Mean annual precipitation (CHIRPS)
- `temp_std`  Mean annual temperature
- `ntl_std`  Nighttime lights composite (VIIRS) as a proxy for urbanicity and economic activity

We include both nighttime lights and population density. While these capture overlapping urbanicity signals, they index distinct dimensions of human settlement: population density reflects residential intensity, while nighttime lights also capture economic activity, infrastructure, and electrification [@elvidge2014; @bhatt2015]. Both have featured in malaria SAE work, and we retain both subject to the multicollinearity check below.

```{r load-covariates}
covariates_panel <- read_csv("three_country_model_panel.csv") %>%
  select(country, district_id, district_name, time, year, label,
         elev_std, rainfall_std, ntl_std, temp_std,
         travel_std, pop_dens_std, poverty_std)

covariates_panel %>% count(country, time, name = "rows")
```

## Joining Covariates to the Panel

The covariate panel's `time` column (1, 2) maps directly to the survey panel's `round_id` (1, 2), so the join is by `district_id` AND `time` — time-varying covariates correctly broadcast to the appropriate round, while static covariates carry the same value to both rounds.

```{r join-covariates}
join_covariates <- function(iso) {
  panels[[iso]] %>%
    left_join(
      covariates_panel %>%
        filter(country == iso) %>%
        select(district_id, time,
               elev_std, rainfall_std, ntl_std, temp_std,
               travel_std, pop_dens_std, poverty_std),
      by = c("district_id", "round_id" = "time")
    )
}

panels <- list(
  GHA = join_covariates("GHA"),
  BFA = join_covariates("BFA"),
  CIV = join_covariates("CIV")
)

# Verify covariates are present
panels[["GHA"]] %>% select(district_id, year, ends_with("_std")) %>% head()
```

# Exploratory Data Analysis

## Correlation Heatmap

A visual check for multicollinearity among the seven covariates across all observations (three countries, two rounds each).

```{r correlation-heatmap, fig.height=7, fig.width=8}
covariate_cols <- c("elev_std", "rainfall_std", "ntl_std", "temp_std",
                    "travel_std", "pop_dens_std", "poverty_std")

cor_matrix <- covariates_panel %>%
  select(all_of(covariate_cols)) %>%
  cor(use = "complete.obs")

ggcorrplot(
  cor_matrix,
  type = "lower",
  lab = TRUE, lab_size = 3.5,
  colors = c("#2166ac", "white", "#b2182b"),
  title = "Pairwise correlations among standardised covariates",
  ggtheme = ggplot2::theme_minimal()
) +
  theme(plot.title = element_text(hjust = 0.5))
```

All retained pairwise correlations remain below approximately 0.7, consistent with the threshold recommended by @dormann2013 for multicollinearity concern in ecological regression. Notably, while `pop_dens_std` and `ntl_std` are moderately correlated (both index human concentration), they remain below the 0.7 threshold and are retained for the conceptual reasons noted above [@elvidge2014; @bhatt2015].

## Distribution of Direct Estimates

```{r direct-est-dist}
direct_est_all %>%
  mutate(prevalence = plogis(y_i)) %>%
  ggplot(aes(x = factor(year), y = prevalence)) +
  geom_violin(fill = "steelblue", alpha = 0.4) +
  geom_jitter(width = 0.1, alpha = 0.4, size = 0.5) +
  facet_wrap(~country, scales = "free_x") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = NULL, y = "Direct prevalence estimate (district)",
       title = "Distribution of district-level direct estimates by country and round",
       subtitle = "Each point is one district; violin shows the distribution") +
  theme_minimal(base_size = 11)
```

## Adjacency Graph (Ghana)

```{r adjacency-graph-plot, fig.height=8}
gha_with_id <- districts_raw[["GHA"]] %>%
  left_join(district_lookups[["GHA"]] %>% select(district_id, district_code),
            by = "district_code") %>%
  arrange(district_id)

centroid_coords <- suppressWarnings(st_coordinates(st_centroid(gha_with_id)))
centroids <- tibble(
  district_id = gha_with_id$district_id,
  x = centroid_coords[, "X"], y = centroid_coords[, "Y"]
)

sp <- spatial[["GHA"]]
edges_df <- tibble(from_id = sp$node1, to_id = sp$node2) %>%
  left_join(centroids %>% rename(x_from = x, y_from = y),
            by = c("from_id" = "district_id")) %>%
  left_join(centroids %>% rename(x_to = x, y_to = y),
            by = c("to_id" = "district_id"))

ggplot() +
  geom_sf(data = districts_raw[["GHA"]],
          fill = "grey95", colour = "grey60", linewidth = 0.2) +
  geom_segment(data = edges_df,
               aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
               colour = "#b2182b", alpha = 0.55, linewidth = 0.3) +
  geom_point(data = centroids, aes(x = x, y = y),
             colour = "black", size = 0.6) +
  labs(title = "District adjacency graph used for the ICAR prior, Ghana",
       subtitle = paste(sp$D, "districts,", sp$N_edges, "edges (queen contiguity)")) +
  theme_void(base_size = 11)
```

# Stan Data Builders

Two builders mirror the two model specifications. The cross-sectional builder filters to a single year; the space-time builder uses both years.

```{r build-stan-data-cs}
build_stan_data_cs <- function(iso, panels, spatial, year, covar_cols = NULL) {
  panel <- panels[[iso]] %>% filter(year == !!year)
  sp <- spatial[[iso]]
  obs_idx <- which(panel$observed == 1)
  
  if (!is.null(covar_cols)) {
    available <- intersect(covar_cols, names(panel))
    X <- if (length(available) > 0) as.matrix(panel[, available]) else matrix(0, nrow(panel), 0)
    K <- ncol(X)
  } else {
    X <- matrix(0, nrow(panel), 0)
    K <- 0L
  }
  
  list(N = sp$D, N_obs = length(obs_idx),
       obs_idx = obs_idx,
       y = panel$y_i[obs_idx], psi = panel$psi_i[obs_idx],
       N_edges = sp$N_edges, node1 = sp$node1, node2 = sp$node2,
       scaling_factor = sp$scaling_factor,
       K = K, X = X)
}
```

```{r build-stan-data-st}
build_stan_data_st <- function(iso, panels, spatial, covar_cols = NULL) {
  panel <- panels[[iso]]
  sp <- spatial[[iso]]
  obs_idx <- which(panel$observed == 1)
  
  if (!is.null(covar_cols)) {
    available <- intersect(covar_cols, names(panel))
    X <- if (length(available) > 0) as.matrix(panel[, available]) else matrix(0, nrow(panel), 0)
    K <- ncol(X)
  } else {
    X <- matrix(0, nrow(panel), 0)
    K <- 0L
  }
  
  list(N = sp$D, T = 2L, NT = nrow(panel),
       N_obs = length(obs_idx), obs_idx = obs_idx,
       district = panel$district_id, time = panel$round_id,
       y = panel$y_i[obs_idx], psi = panel$psi_i[obs_idx],
       N_edges = sp$N_edges, node1 = sp$node1, node2 = sp$node2,
       scaling_factor = sp$scaling_factor,
       K = K, X = X)
}
```

## Assemble Stan Data for All Models

```{r assemble-stan-data}
covar_cols <- c("elev_std", "rainfall_std", "ntl_std", "temp_std",
                "travel_std", "pop_dens_std", "poverty_std")

stan_data <- list()

# Cross-sectional reference: Ghana 2022 only
stan_data[["GHA_CS_2022"]] <- build_stan_data_cs(
  "GHA", panels, spatial, year = 2022, covar_cols = covar_cols
)

# Space-time models for all three countries
stan_data[["GHA_ST"]] <- build_stan_data_st("GHA", panels, spatial, covar_cols = covar_cols)
stan_data[["BFA_ST"]] <- build_stan_data_st("BFA", panels, spatial, covar_cols = covar_cols)
stan_data[["CIV_ST"]] <- build_stan_data_st("CIV", panels, spatial, covar_cols = covar_cols)

map_dfr(stan_data, function(d) {
  tibble(
    N = d$N,
    NT = ifelse(is.null(d$NT), d$N, d$NT),
    N_obs = d$N_obs, K = d$K,
    N_edges = d$N_edges,
    scaling_factor = round(d$scaling_factor, 3)
  )
}, .id = "model")
```

# Model Specification

## Cross-Sectional Reference Model (Ghana 2022)

Following @wakefield2020, we begin with a cross-sectional Bayesian Fay-Herriot specification fitted to the most recent biomarker round in Ghana (2022). The latent logit prevalence for district $i$ decomposes as

$$\theta_i = \mu + b_i + \mathbf{x}_i^\top \boldsymbol{\beta},$$

where $\mu$ is the national-level logit intercept, $b_i$ is the BYM2 district random effect, and $\mathbf{x}_i$ contains the seven standardised covariates. This specification serves two purposes: it provides a baseline estimate of district-level prevalence at the most policy-relevant time point, and it establishes a reference for the spatial hyperparameters $(\sigma_b, \phi)$ against which the space-time extension is benchmarked. Under correct specification, these should remain broadly stable when the temporal component is added, since the single round identifies the spatial structure independently of temporal confounding.

## Space-Time Extension

The space-time model jointly fits both DHS rounds, with the latent logit prevalence for district $i$ at round $t \in \{1, 2\}$ decomposing as

$$\theta_{it} = \mu + b_i + \delta_t + \gamma_{it} + \mathbf{x}_{it}^\top \boldsymbol{\beta}.$$

With $T = 2$ snapshots from current-status biomarker data, a random walk prior over time (as used for birth-history-derived child mortality in @mercer2015 and @wu2021) is rank-deficient and collapses to a single increment. We therefore use a **fixed year effect**, setting $\delta_1 = 0$ and treating $\delta_2$ as a free regression coefficient. This distinction follows the consensus in the DHS small area estimation literature regarding outcomes derived from biomarkers (current-status snapshots) versus full birth histories (multi-year time series from a single survey).

The space-time interaction $\gamma_{it}$ follows the Type I (IID) specification of @knorrheld2000. Following the identifiability analysis of @goicoa2018, we impose:

1. The constraint $\gamma_{i,2} = -\gamma_{i,1}$, which reduces $\gamma$ to one free parameter per district (we denote this $\gamma_i^*$).
2. The sum-to-zero constraint $\sum_i \gamma_{i,t} = 0$ for each $t$, which prevents the interaction from absorbing the national-level intercept and trend.

## BYM2 Spatial Random Effect

The district effect uses the BYM2 reparameterisation of @riebler2016:

$$b_i = \sigma_b\!\left(\sqrt{1-\phi}\, v_i + \sqrt{\phi/s}\, u_i\right),$$

which combines an IID component ($v_i \sim \mathcal{N}(0, 1)$) with a scaled intrinsic conditional autoregressive (ICAR) component ($u_i$ with sum-to-zero constraint) through the mixing parameter $\phi \in (0, 1)$. The scaling factor $s$ ensures $\phi$ is interpretable as the fraction of district-level variance that is spatially structured. This formulation has substantive advantages over the original @besag1991 BYM model, which conflated the two component variances.

## Priors

Following PC-style recommendations [@simpson2017; @riebler2016]:

| Parameter | Prior | Justification |
|---|---|---|
| $\mu$ | $\mathcal{N}(-1.85, 1)$ | Centred near logit of moderate-burden prevalence (~14%) |
| $\sigma_b$ | $\text{Exp}(4.6)$ | PC-style prior; approximates $P(\sigma_b > 1) = 0.01$ |
| $\phi$ | $\text{Beta}(3, 2)$ | Mild prior preference for spatial > IID variation |
| $\delta_2$ | $\mathcal{N}(0, 5)$ | Weakly informative on national trend; data-driven |
| $\sigma_\gamma$ | $\text{Exp}(2)$ | Allows substantial district trend heterogeneity |
| $\beta_k$ | $\mathcal{N}(0, 1)$ | Weakly informative on standardised covariates |
| $v_i$ | $\mathcal{N}(0, 1)$ | BYM2 IID component, standard normal |
| $u_i$ | ICAR with $\sum_i u_i = 0$ | BYM2 spatial component |

# Model Fitting

## Cross-Sectional Model: Ghana 2022

## Cross-Sectional Model Comparison: IID vs ICAR vs BYM2

We fit three spatial specifications to Ghana 2022 to justify the BYM2 choice via LOO-CV.

### IID Model (no spatial structure)

```{r fit-iid, cache=TRUE}
iid_code <- '
data {
  int<lower=1> N;
  int<lower=1> N_obs;
  array[N_obs] int<lower=1, upper=N> obs_idx;
  vector[N_obs] y;
  vector<lower=0>[N_obs] psi;
  int<lower=1> K;
  matrix[N, K] X;
}
parameters {
  real mu;
  vector[N] v;
  real<lower=0> sigma_b;
  vector[K] beta;
}
transformed parameters {
  vector[N] theta;
  theta = mu + sigma_b * v + X * beta;
}
model {
  mu ~ normal(-1.85, 1.0);
  sigma_b ~ exponential(4.6);
  beta ~ normal(0, 1);
  v ~ std_normal();
  y ~ normal(theta[obs_idx], psi);
}
generated quantities {
  vector[N_obs] log_lik;
  vector[N] prevalence;
  for (i in 1:N_obs)
    log_lik[i] = normal_lpdf(y[i] | theta[obs_idx[i]], psi[i]);
  for (i in 1:N)
    prevalence[i] = inv_logit(theta[i]);
}
'
writeLines(iid_code, "fh_iid.stan")
mod_iid <- cmdstan_model("fh_iid.stan")

sd_cs <- stan_data[["GHA_CS_2022"]]

fit_iid <- mod_iid$sample(
  data = sd_cs, chains = 4, parallel_chains = 4,
  iter_warmup = 1000, iter_sampling = 1500,
  adapt_delta = 0.95, max_treedepth = 12, refresh = 500
)
fit_iid$diagnostic_summary()
```

### ICAR Model (pure spatial, no IID)

```{r fit-icar, cache=TRUE}
icar_code <- '
data {
  int<lower=1> N;
  int<lower=1> N_obs;
  array[N_obs] int<lower=1, upper=N> obs_idx;
  vector[N_obs] y;
  vector<lower=0>[N_obs] psi;
  int<lower=1> N_edges;
  array[N_edges] int<lower=1, upper=N> node1;
  array[N_edges] int<lower=1, upper=N> node2;
  real<lower=0> scaling_factor;
  int<lower=1> K;
  matrix[N, K] X;
}
parameters {
  real mu;
  vector[N] u_raw;
  real<lower=0> sigma_b;
  vector[K] beta;
}
transformed parameters {
  vector[N] theta;
  vector[N] u = u_raw - mean(u_raw);
  theta = mu + sigma_b * u / sqrt(scaling_factor) + X * beta;
}
model {
  mu ~ normal(-1.85, 1.0);
  sigma_b ~ exponential(4.6);
  beta ~ normal(0, 1);
  target += -0.5 * dot_self(u_raw[node1] - u_raw[node2]);
  sum(u_raw) ~ normal(0, 0.001 * N);
  y ~ normal(theta[obs_idx], psi);
}
generated quantities {
  vector[N_obs] log_lik;
  vector[N] prevalence;
  for (i in 1:N_obs)
    log_lik[i] = normal_lpdf(y[i] | theta[obs_idx[i]], psi[i]);
  for (i in 1:N)
    prevalence[i] = inv_logit(theta[i]);
}
'
writeLines(icar_code, "fh_icar.stan")
mod_icar <- cmdstan_model("fh_icar.stan")

fit_icar <- mod_icar$sample(
  data = sd_cs, chains = 4, parallel_chains = 4,
  iter_warmup = 1000, iter_sampling = 1500,
  adapt_delta = 0.95, max_treedepth = 12, refresh = 500
)
fit_icar$diagnostic_summary()
```

### BYM2 Cross-Sectional (with log_lik for LOO)

```{r fit-cross-sectional}
mod_cs <- cmdstan_model("fh_bym2_cross_sectional.stan")

fit_gha_cs <- mod_cs$sample(
  data            = stan_data[["GHA_CS_2022"]],
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 1500,
  iter_sampling   = 2000,
  adapt_delta     = 0.99,
  max_treedepth   = 14,
  seed            = 123,
  refresh         = 500
)

fit_gha_cs$diagnostic_summary()

fit_gha_cs$print(c(
  "mu", "sigma_b", "phi", "p_national",
  "beta[1]", "beta[2]", "beta[3]", "beta[4]",
  "beta[5]", "beta[6]", "beta[7]"
))
```

### LOO-CV Comparison

```{r loo-cv, fig.height=5, fig.width=9}
library(loo)

loo_iid  <- fit_iid$loo(variables = "log_lik")
loo_icar <- fit_icar$loo(variables = "log_lik")

# BYM2 needs log_lik - check if your stan file has it
# If not, use the BYM2 LOO stan file
if ("log_lik" %in% fit_gha_cs$metadata()$stan_variables) {
  loo_bym2 <- fit_gha_cs$loo(variables = "log_lik")
} else {
  # Refit with log_lik version
  bym2_loo_code <- '
data {
  int<lower=1> N;
  int<lower=1> N_obs;
  array[N_obs] int<lower=1, upper=N> obs_idx;
  vector[N_obs] y;
  vector<lower=0>[N_obs] psi;
  int<lower=1> N_edges;
  array[N_edges] int<lower=1, upper=N> node1;
  array[N_edges] int<lower=1, upper=N> node2;
  real<lower=0> scaling_factor;
  int<lower=1> K;
  matrix[N, K] X;
}
parameters {
  real mu;
  vector[N] v;
  vector[N] u_raw;
  real<lower=0> sigma_b;
  real<lower=0, upper=1> phi;
  vector[K] beta;
}
transformed parameters {
  vector[N] theta;
  vector[N] b;
  vector[N] u = u_raw - mean(u_raw);
  b = sigma_b * (sqrt(1 - phi) * v + sqrt(phi / scaling_factor) * u);
  theta = mu + b + X * beta;
}
model {
  mu ~ normal(-1.85, 1.0);
  sigma_b ~ exponential(4.6);
  phi ~ beta(3, 2);
  beta ~ normal(0, 1);
  v ~ std_normal();
  target += -0.5 * dot_self(u_raw[node1] - u_raw[node2]);
  sum(u_raw) ~ normal(0, 0.001 * N);
  y ~ normal(theta[obs_idx], psi);
}
generated quantities {
  vector[N_obs] log_lik;
  vector[N] prevalence;
  for (i in 1:N_obs)
    log_lik[i] = normal_lpdf(y[i] | theta[obs_idx[i]], psi[i]);
  for (i in 1:N)
    prevalence[i] = inv_logit(theta[i]);
}
'
  writeLines(bym2_loo_code, "fh_bym2_loo.stan")
  mod_bym2_loo <- cmdstan_model("fh_bym2_loo.stan")
  fit_bym2_loo <- mod_bym2_loo$sample(
    data = sd_cs, chains = 4, parallel_chains = 4,
    iter_warmup = 1000, iter_sampling = 1500,
    adapt_delta = 0.95, max_treedepth = 12, refresh = 500
  )
  loo_bym2 <- fit_bym2_loo$loo(variables = "log_lik")
}

loo_comp <- loo_compare(list(IID = loo_iid, ICAR = loo_icar, BYM2 = loo_bym2))
print(loo_comp, simplify = FALSE)

# Plot
loo_df <- as_tibble(loo_comp, rownames = "Model")

p_loo <- loo_df %>%
  mutate(Model = factor(Model, levels = rev(Model))) %>%
  ggplot(aes(x = Model, y = elpd_diff)) +
  geom_col(aes(fill = Model), width = 0.6, alpha = 0.8) +
  geom_errorbar(aes(ymin = elpd_diff - 2 * se_diff,
                     ymax = elpd_diff + 2 * se_diff), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = c("BYM2" = "#a80050", "ICAR" = "#2563eb", "IID" = "#f5a623")) +
  coord_flip() +
  labs(title = "LOO-CV: IID vs ICAR vs BYM2 (Ghana 2022)",
       subtitle = "Delta ELPD relative to best model (higher = better)",
       x = NULL, y = "Delta ELPD") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

ggsave("output/figures/loo_cv_comparison.png", p_loo, width = 9, height = 5, dpi = 300, bg = "white")
p_loo
```

### Model Comparison Maps: Direct vs IID vs ICAR vs BYM2

```{r model-comparison-maps, fig.height=6, fig.width=14}
# Extract prevalence from each model
get_prev <- function(fit) {
  s <- fit$summary("prevalence")
  s$median
}

prev_iid_v  <- get_prev(fit_iid)
prev_icar_v <- get_prev(fit_icar)

# BYM2: use fit_gha_cs or fit_bym2_loo (whichever has prevalence)
if ("prevalence" %in% fit_gha_cs$metadata()$stan_variables) {
  prev_bym2_v <- get_prev(fit_gha_cs)
} else {
  prev_bym2_v <- get_prev(fit_bym2_loo)
}

D_gha_cs <- stan_data[["GHA_CS_2022"]]$N

model_prev <- bind_rows(
  tibble(district_id = 1:D_gha_cs, prev = prev_iid_v[1:D_gha_cs], model = "IID"),
  tibble(district_id = 1:D_gha_cs, prev = prev_icar_v[1:D_gha_cs], model = "ICAR"),
  tibble(district_id = 1:D_gha_cs, prev = prev_bym2_v[1:D_gha_cs], model = "BYM2"),
  panels[["GHA"]] %>% filter(round_id == 2) %>%
    mutate(prev = ifelse(observed == 1, plogis(y_i), NA_real_), model = "Direct") %>%
    select(district_id, prev, model)
) %>%
  mutate(model = factor(model, levels = c("Direct", "IID", "ICAR", "BYM2")))

# Join to shapefile via district_id
gha_shp <- districts_raw[["GHA"]] %>%
  mutate(district_id = row_number())

maps_comp <- map_dfr(levels(model_prev$model), function(m) {
  gha_shp %>%
    left_join(model_prev %>% filter(model == m), by = "district_id") %>%
    mutate(model = m)
}) %>%
  mutate(model = factor(model, levels = c("Direct", "IID", "ICAR", "BYM2")))

p_comp <- ggplot(maps_comp) +
  geom_sf(aes(fill = prev), colour = "grey50", linewidth = 0.1) +
  scale_fill_viridis_c(name = "Prevalence",
                       labels = scales::percent_format(accuracy = 1),
                       option = "magma", direction = -1,
                       na.value = "grey80") +
  facet_wrap(~model, nrow = 1) +
  labs(title = "Model comparison: Direct vs IID vs ICAR vs BYM2 (Ghana 2022)",
       subtitle = "Grey = unobserved in direct estimates. Spatial models fill all 260 districts.") +
  theme_void(base_size = 11) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 12, face = "bold"))

ggsave("output/figures/model_comparison_maps_ghana.png", p_comp, width = 14, height = 6, dpi = 300, bg = "white")
p_comp
```

### Posterior Predictive Checks (Cross-Sectional)

```{r ppc-cs}
y_obs_cs <- stan_data[["GHA_CS_2022"]]$y
y_rep_cs <- fit_gha_cs$draws("y_rep", format = "matrix")

ppc_dens_overlay(y = y_obs_cs, yrep = y_rep_cs[1:100, ]) +
  labs(title = "PPC: Density overlay (Ghana 2022 cross-sectional)")

ppc_stat(y = y_obs_cs, yrep = y_rep_cs, stat = "mean") +
  labs(title = "PPC: Mean of y")

ppc_stat(y = y_obs_cs, yrep = y_rep_cs, stat = "sd") +
  labs(title = "PPC: Standard deviation of y")
```

## Space-Time Model: All Three Countries

We fit the space-time model separately for each country. Country-specific fitting (rather than a single pooled model) follows the established convention in DHS-based SAE work [@wakefield2020; @wu2021] and is justified by structural differences in graph topology, transmission dynamics, and survey design.

```{r fit-spacetime-all}
mod_st <- cmdstan_model("fh_bym2_typeI.stan")

fit_st <- list()

for (iso in c("GHA", "BFA", "CIV")) {
  cat("\n========== Fitting space-time model for", iso, "==========\n")
  
  fit_st[[iso]] <- mod_st$sample(
    data            = stan_data[[paste0(iso, "_ST")]],
    chains          = 4,
    parallel_chains = 4,
    iter_warmup     = 1500,
    iter_sampling   = 2000,
    adapt_delta     = 0.99,
    max_treedepth   = 14,
    seed            = 1729,
    refresh         = 500
  )
  
  cat("\nDiagnostics for", iso, ":\n")
  print(fit_st[[iso]]$diagnostic_summary())
}
```

## Summary Across Countries

```{r summary-across-countries}
summary_all <- map_dfr(c("GHA", "BFA", "CIV"), function(iso) {
  draws <- fit_st[[iso]]$draws(
    variables = c("mu", "delta_2", "sigma_b", "phi", "sigma_gamma",
                  "p_round1", "p_round2"),
    format = "draws_matrix"
  )
  
  tibble(
    country  = iso,
    variable = colnames(draws),
    mean     = colMeans(draws),
    median   = apply(draws, 2, median),
    lower    = apply(draws, 2, quantile, 0.025),
    upper    = apply(draws, 2, quantile, 0.975),
    rhat     = apply(draws, 2, posterior::rhat),
    ess_bulk = apply(draws, 2, posterior::ess_bulk)
  )
})

print(n = 25, summary_all)

# fit_st$print(c(
#   "mu", "sigma_b", "phi", "p_national",
#   "beta[1]", "beta[2]", "beta[3]", "beta[4]",
#   "beta[5]", "beta[6]", "beta[7]"
# ))
```

# Posterior Analysis and Visualisation

## Prior vs Posterior Comparison (Ghana Space-Time)

```{r prior-post}
post_draws <- fit_st[["GHA"]]$draws(
  c("mu", "delta_2", "sigma_b", "phi", "sigma_gamma"),
  format = "draws_df"
)

prior_post_plot <- function(post_var, prior_fun, x_range, label) {
  ggplot() +
    geom_function(fun = prior_fun, xlim = x_range,
                  aes(colour = "Prior"), linewidth = 0.9) +
    geom_density(data = post_draws,
                 aes(x = .data[[post_var]], colour = "Posterior"),
                 linewidth = 0.9) +
    scale_colour_manual(values = c(Prior = "grey50", Posterior = "#b2182b")) +
    labs(x = label, y = "Density", colour = NULL) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "none")
}

p_mu     <- prior_post_plot("mu",          ~ dnorm(.x, -1.85, 1), c(-4, 0),  expression(mu))
p_delta  <- prior_post_plot("delta_2",     ~ dnorm(.x, 0, 5),     c(-3, 3),  expression(delta[2]))
p_sigb   <- prior_post_plot("sigma_b",     ~ dexp(.x, 4.6),       c(0, 2),   expression(sigma[b]))
p_phi    <- prior_post_plot("phi",         ~ dbeta(.x, 3, 2),     c(0, 1),   expression(phi))
p_siggam <- prior_post_plot("sigma_gamma", ~ dexp(.x, 2),         c(0, 2),   expression(sigma[gamma]))

(p_mu | p_delta | p_sigb) / (p_phi | p_siggam | plot_spacer()) +
  plot_annotation(title = "Prior vs posterior, space-time BYM2 model (Ghana)")
```

## Covariate Effects Across Countries

```{r covariate-forest}
covar_labels <- c("Elevation", "Rainfall", "Nighttime lights", "Temperature",
                  "Travel time", "Pop. density (log)", "Poverty")

# 2. Process directly from your pre-calculated summary_all dataframe
beta_all <- summary_all %>%
  # Filter only the rows where the variable starts with "beta["
  filter(str_detect(variable, "^beta\\[")) %>%
  mutate(
    # Extract the numeric index from "beta[1]", "beta[2]", etc.
    beta_idx = as.integer(str_extract(variable, "\\d+")),
    
    # Map the index to the correct human-readable label
    covariate = covar_labels[beta_idx],
    
    # Calculate significance strictly using the lower/upper bounds in the table
    # If both bounds are greater than 0, or both are less than 0, it is significant
    is_significant = (lower > 0) | (upper < 0),
    sig_marker = ifelse(is_significant, "*", "")
  )

# 3. Factor the covariates and countries to preserve plotting order
beta_all$covariate <- factor(beta_all$covariate, levels = rev(covar_labels))
beta_all$country <- factor(beta_all$country, levels = c("GHA", "BFA", "CIV"))

# 4. Generate the plot
p_covars <- ggplot(beta_all, aes(x = mean, y = covariate, colour = country, group = country)) +
  
  geom_pointrange(aes(xmin = lower, xmax = upper),
                  position = position_dodge(width = 0.6), size = 0.8) +
                  
  # FIXED ALIGNMENT: Added hjust = 0 and locked the dodge with group = country
  geom_text(aes(x = upper + 0.02, label = sig_marker),
            position = position_dodge(width = 0.6),
            size = 7,          # Slightly larger for visibility on a slide
            vjust = 0.75,      # Pulls the top-heavy asterisk down to center it with the line
            hjust = 0,         # Left-aligns it exactly 0.02 units away from the line
            show.legend = FALSE, 
            colour = "black") +
            
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40", linewidth = 0.8) +
  scale_colour_brewer(palette = "Dark2", name = "Country") +
  
  scale_x_continuous(expand = expansion(mult = c(0.05, 0.15))) + 
  
  labs(x = "Coefficient (logit scale)", y = NULL,
       title = "Covariate effects: Individual Space-Time Models",
       subtitle = "Posterior mean and credible intervals. Asterisks (*) denote significant effects.") +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

ggsave("output/figures/individual_covariate_effects.png", p_covars, width = 10, height = 7, dpi = 300, bg = "white")
p_covars
```

## Prevalence Maps (All Countries, Both Rounds)

```{r prevalence-maps, fig.height=12, fig.width=12}
make_prev_map_df <- function(iso, round) {
  prev_sum <- fit_st[[iso]]$summary("prevalence",
    median = ~ median(.x),
    lower  = ~ quantile(.x, 0.025),
    upper  = ~ quantile(.x, 0.975)
  ) %>%
    mutate(row_id = as.integer(gsub("prevalence\\[|\\]", "", variable))) %>%
    left_join(
      panels[[iso]] %>% mutate(row_id = row_number()) %>%
        select(row_id, district_id, round_id),
      by = "row_id"
    ) %>%
    filter(round_id == round) %>%
    left_join(district_lookups[[iso]] %>% select(district_id, district_code),
              by = "district_id")
  
  districts_raw[[iso]] %>%
    left_join(prev_sum, by = "district_code") %>%
    mutate(country = iso, round_label = paste("Round", round))
}

all_maps <- map_dfr(c("GHA", "BFA", "CIV"),
                    function(iso) bind_rows(
                      make_prev_map_df(iso, 1),
                      make_prev_map_df(iso, 2)
                    ))

ggplot(all_maps) +
  geom_sf(aes(fill = median), colour = "grey40", linewidth = 0.1) +
  scale_fill_viridis_c(name = "Prevalence",
                       labels = scales::percent_format(accuracy = 1),
                       option = "magma", direction = -1) +
  facet_grid(country ~ round_label) +
  labs(title = "Under-5 malaria prevalence across three West African countries",
       subtitle = "Posterior medians from BYM2 space-time model") +
  theme_void(base_size = 11) +
  theme(legend.position = "right")
```

## District Trend Deviations (Ghana)

The interaction term captures district-specific deviations from the national trend, recoverable via $\gamma_i^*$. Negative values indicate districts that declined faster than the national average; positive values indicate stagnant or rising trajectories. Districts with bold outlines have 95% credible intervals that exclude zero.

```{r trend-deviation-map}
dev_summary_gha <- fit_st[["GHA"]]$summary("gamma_star",
  median = ~ median(.x),
  lower  = ~ unname(quantile(.x, 0.025)),
  upper  = ~ unname(quantile(.x, 0.975))
) %>%
  mutate(
    district_id = as.integer(gsub("gamma_star\\[|\\]", "", variable)),
    confident = (lower > 0) | (upper < 0)
  ) %>%
  left_join(district_lookups[["GHA"]] %>%
              select(district_id, district_code, district_name),
            by = "district_id")

map_dev <- districts_raw[["GHA"]] %>%
  left_join(dev_summary_gha, by = "district_code")

dev_limit <- max(abs(range(map_dev$median, na.rm = TRUE)))

ggplot(map_dev) +
  geom_sf(aes(fill = median), colour = "grey40", linewidth = 0.1) +
  scale_fill_gradient2(
    name = expression(gamma[i]^"*"),
    low = "#2166ac", mid = "white", high = "#b2182b",
    midpoint = 0, limits = c(-dev_limit, dev_limit)
  ) +
  geom_sf(data = map_dev %>% filter(confident),
          fill = NA, colour = "black", linewidth = 0.4) +
  labs(
    title = "District-level deviation from national malaria trend, Ghana 2014–2022",
    subtitle = "Blue: faster decline than national. Red: slower / rising. Bold outline: 95% CI excludes 0."
  ) +
  theme_void(base_size = 11)
```

## Top Trend-Deviating Districts

```{r top-deviation-districts}
top_neg <- dev_summary_gha %>% arrange(median) %>% head(10) %>%
  mutate(direction = "Faster decline")
top_pos <- dev_summary_gha %>% arrange(desc(median)) %>% head(10) %>%
  mutate(direction = "Slower / rising")

bind_rows(top_neg, top_pos) %>%
  mutate(district_name = fct_reorder(district_name, median)) %>%
  ggplot(aes(x = median, y = district_name, colour = direction)) +
  geom_pointrange(aes(xmin = lower, xmax = upper)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_colour_manual(values = c("Faster decline" = "#2166ac",
                                 "Slower / rising" = "#b2182b")) +
  labs(x = expression(gamma[i]^"*" * " (logit scale)"), y = NULL,
       title = "Top districts with credible deviation from national trend, Ghana",
       colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

## Uncertainty Maps

Districts with no direct estimates rely entirely on spatial borrowing, their credible intervals are wider, visible as higher values on the uncertainty map.

```{r uncertainty-maps, fig.height=5}
make_uncertainty_df <- function(iso, round) {
  unc <- fit_st[[iso]]$summary("prevalence",
    width = ~ unname(quantile(.x, 0.975) - quantile(.x, 0.025))
  ) %>%
    mutate(row_id = as.integer(gsub("prevalence\\[|\\]", "", variable))) %>%
    left_join(
      panels[[iso]] %>% mutate(row_id = row_number()) %>%
        select(row_id, district_id, round_id, observed),
      by = "row_id"
    ) %>%
    filter(round_id == round) %>%
    left_join(district_lookups[[iso]] %>% select(district_id, district_code),
              by = "district_id")
  
  districts_raw[[iso]] %>%
    left_join(unc, by = "district_code") %>%
    mutate(country = iso)
}

unc_r2 <- map_dfr(c("GHA", "BFA", "CIV"),
                  function(iso) make_uncertainty_df(iso, 2))

ggplot(unc_r2) +
  geom_sf(aes(fill = width), colour = "grey40", linewidth = 0.1) +
  scale_fill_viridis_c(name = "95% CI width",
                       labels = scales::percent_format(accuracy = 1),
                       option = "cividis") +
  facet_wrap(~country) +
  labs(title = "Posterior uncertainty in Round 2 prevalence estimates",
       subtitle = "Wider intervals = greater reliance on spatial smoothing (no direct data)") +
  theme_void(base_size = 11)
```

# Model Diagnostics

## Cross-Sectional Trace Plots

```{r cs-trace}
mcmc_trace(fit_gha_cs$draws(c("mu", "sigma_b", "phi")))
```

## Space-Time Diagnostics (Ghana)

```{r st-diagnostics}

mcmc_rank_overlay(fit_st[["GHA"]]$draws(
  c("mu", "delta_2", "sigma_b", "phi", "sigma_gamma")
))

mcmc_pairs(
  fit_st[["GHA"]]$draws(c("sigma_b", "phi", "sigma_gamma", "delta_2")),
  off_diag_args = list(size = 0.5, alpha = 0.3)
)

```

















# Additional Presentation Plots

## Observed vs Unobserved Districts Map

A map showing which districts have direct DHS survey data (observed) versus those that rely entirely on spatial borrowing from the BYM2 model (unobserved).

```{r observed-unobserved-map, fig.height=10, fig.width=10}
make_obs_map <- function(iso, round) {
  panel_round <- panels[[iso]] %>% filter(round_id == round)
  
  obs_by_district <- panel_round %>%
    select(district_id, observed) %>%
    mutate(status = ifelse(observed == 1, "Observed", "Unobserved")) %>%
    left_join(district_lookups[[iso]] %>% select(district_id, district_code),
              by = "district_id")
  
  districts_raw[[iso]] %>%
    left_join(obs_by_district, by = "district_code") %>%
    mutate(country = iso,
           round_label = paste("Round", round))
}

obs_maps <- map_dfr(c("GHA", "BFA", "CIV"),
                    function(iso) bind_rows(
                      make_obs_map(iso, 1),
                      make_obs_map(iso, 2)
                    ))

p_obs <- ggplot(obs_maps) +
  geom_sf(aes(fill = status), colour = "grey50", linewidth = 0.1) +
  scale_fill_manual(
    values = c("Observed" = "#2563eb", "Unobserved" = "#fbbf24"),
    name = "Survey data"
  ) +
  facet_grid(country ~ round_label) +
  labs(
    title = "Districts with direct DHS data vs unobserved districts",
    subtitle = "Unobserved districts rely entirely on spatial borrowing from neighbours and covariates"
  ) +
  theme_void(base_size = 11) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 11, face = "bold"))

ggsave("output/figures/observed_vs_unobserved_map.png", p_obs, width = 10, height = 10, dpi = 300, bg = "white")
p_obs
```

## Credible Interval Comparison: Observed vs Unobserved Districts

```{r ci-comparison-obs-unobs, fig.height=6, fig.width=10}
make_ci_comparison <- function(iso, round) {
  D <- stan_data[[paste0(iso, "_ST")]]$N
  
  # Use default summary (returns q5, q95, mean, median, sd, etc.)
  prev_summary <- fit_st[[iso]]$summary("prevalence") %>%
    mutate(row_id = as.integer(gsub("prevalence\\[|\\]", "", variable))) %>%
    left_join(
      panels[[iso]] %>% mutate(row_id = row_number()) %>%
        select(row_id, district_id, round_id, observed),
      by = "row_id"
    ) %>%
    filter(round_id == round) %>%
    mutate(
      ci_width = q95 - q5,
      status = ifelse(observed == 1, "Observed", "Unobserved"),
      country = iso
    ) %>%
    left_join(district_lookups[[iso]] %>% select(district_id, district_name),
              by = "district_id")
  
  prev_summary
}

ci_all <- map_dfr(c("GHA", "BFA", "CIV"),
                  function(iso) make_ci_comparison(iso, 2))

# Boxplot comparing CI widths
p_ci_box <- ggplot(ci_all, aes(x = status, y = ci_width, fill = status)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8) +
  facet_wrap(~country, scales = "free_y") +
  scale_fill_manual(values = c("Observed" = "#2563eb", "Unobserved" = "#fbbf24")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Width of 90% credible intervals: observed vs unobserved districts",
    subtitle = "Round 2 estimates. Unobserved districts have wider intervals (more uncertainty)",
    x = NULL, y = "Width of 90% CrI",
    fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

ggsave("output/figures/ci_comparison_obs_unobs.png", p_ci_box, width = 10, height = 6, dpi = 300, bg = "white")
p_ci_box
```

## Caterpillar Plot: Sorted Prevalence with Credible Intervals (Ghana)

```{r caterpillar-plot, fig.height=7, fig.width=10}
ci_gha <- ci_all %>%
  filter(country == "GHA") %>%
  arrange(median) %>%
  mutate(rank = row_number())

p_cat <- ggplot(ci_gha, aes(x = rank, y = median, colour = status)) +
  geom_linerange(aes(ymin = q5, ymax = q95), alpha = 0.5, linewidth = 0.3) +
  geom_point(size = 0.6) +
  scale_colour_manual(values = c("Observed" = "#2563eb", "Unobserved" = "#fbbf24")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Caterpillar plot: district prevalence estimates, Ghana Round 2",
    subtitle = "260 districts sorted by posterior median. Gold = unobserved (wider CrI).",
    x = "District (sorted)", y = "Prevalence (posterior median and 90% CrI)",
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

ggsave("output/figures/caterpillar_plot_ghana_r2.png", p_cat, width = 10, height = 7, dpi = 300, bg = "white")
p_cat
```

## Direct vs Model-Smoothed Estimates (Ghana)

```{r direct-vs-smoothed, fig.height=7, fig.width=9}
gha_panel_r2 <- panels[["GHA"]] %>% filter(round_id == 2)
gha_direct <- gha_panel_r2 %>%
  filter(observed == 1) %>%
  mutate(direct_prev = plogis(y_i))

gha_model <- fit_st[["GHA"]]$summary("prevalence") %>%
  mutate(row_id = as.integer(gsub("prevalence\\[|\\]", "", variable))) %>%
  left_join(
    panels[["GHA"]] %>% mutate(row_id = row_number()) %>%
      select(row_id, district_id, round_id),
    by = "row_id"
  ) %>%
  filter(round_id == 2)

shrinkage_df <- gha_direct %>%
  select(district_id, direct_prev) %>%
  left_join(gha_model %>% select(district_id, model_prev = median,
                                   model_lower = q5, model_upper = q95),
            by = "district_id") %>%
  left_join(district_lookups[["GHA"]] %>% select(district_id, district_name),
            by = "district_id")

p_shrink <- ggplot(shrinkage_df, aes(x = direct_prev, y = model_prev)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_linerange(aes(ymin = model_lower, ymax = model_upper),
                 alpha = 0.15, colour = "#a80050") +
  geom_point(alpha = 0.6, colour = "#a80050", size = 1.5) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "BYM2 shrinkage: direct estimates vs model-smoothed posteriors",
    subtitle = "Ghana Round 2. Points below diagonal = model pulled estimate down. Bars = 90% CrI.",
    x = "Direct estimate (design-based)",
    y = "Model posterior median"
  ) +
  theme_minimal(base_size = 12)

ggsave("output/figures/direct_vs_smoothed_ghana_r2.png", p_shrink, width = 9, height = 7, dpi = 300, bg = "white")
p_shrink
```

<!-- ## Direct Estimates Map (Ghana) -->

<!-- ```{r direct-estimates-map, fig.height=8, fig.width=8} -->
<!-- direct_map_df <- panels[["GHA"]] %>% -->
<!--   filter(round_id == 2) %>% -->
<!--   mutate(direct_prev = ifelse(observed == 1, plogis(y_i), NA_real_)) %>% -->
<!--   left_join(district_lookups[["GHA"]] %>% select(district_id, district_code), -->
<!--             by = "district_id") -->

<!-- p_direct_map <- districts_raw[["GHA"]] %>% -->
<!--   left_join(direct_map_df %>% select(district_code, direct_prev, observed), -->
<!--             by = "district_code") %>% -->
<!--   ggplot() + -->
<!--   geom_sf(aes(fill = direct_prev), colour = "grey40", linewidth = 0.15) + -->
<!--   scale_fill_viridis_c( -->
<!--     name = "Prevalence", -->
<!--     labels = scales::percent_format(accuracy = 1), -->
<!--     option = "magma", direction = -1, -->
<!--     na.value = "grey80" -->
<!--   ) + -->
<!--   labs( -->
<!--     title = "Direct estimates of under-5 malaria prevalence, Ghana 2022", -->
<!--     subtitle = "Design-based Hajek estimates. Grey = no direct data (unobserved districts)." -->
<!--   ) + -->
<!--   theme_void(base_size = 11) + -->
<!--   theme(legend.position = "right") -->

<!-- ggsave("output/figures/direct_estimates_map_ghana_r2.png", p_direct_map, width = 8, height = 8, dpi = 300, bg = "white") -->
<!-- p_direct_map -->
<!-- ``` -->


# Three-Country Pooled Model

```{r}
gha_sf <- districts_raw[["GHA"]] %>%
  mutate(country = "GHA", country_id = 1L,
         local_id = row_number())

civ_sf <- districts_raw[["CIV"]] %>%
  mutate(country = "CIV", country_id = 2L,
         local_id = row_number())

bfa_sf <- districts_raw[["BFA"]] %>%
  mutate(country = "BFA", country_id = 3L,
         local_id = row_number())

D_gha <- nrow(gha_sf)  # 260
D_civ <- nrow(civ_sf)  # 113
D_bfa <- nrow(bfa_sf)  # 351
D_all <- D_gha + D_civ + D_bfa  # 724


#assigning glboal IDs
gha_sf$global_id <- 1:D_gha
civ_sf$global_id <- (D_gha + 1):(D_gha + D_civ)
bfa_sf$global_id <- (D_gha + D_civ + 1):D_all

# binding them all together
all_sf <- bind_rows(gha_sf, civ_sf, bfa_sf)

# constracting the queens contiguity on the combined spatial shapefile
nb_all <- poly2nb(all_sf)

# counting cross border edges 
cross_border <- 0
within_country <- 0
for (i in seq_along(nb_all)) {
  for (j in nb_all[[i]]) {
    if (j > i) {  # avoid double counting
      if (all_sf$country[i] != all_sf$country[j]) {
        cross_border <- cross_border + 1
      } else {
        within_country <- within_country + 1
      }
    }
  }
}

cross_border

# Extracting edge list for Stan
edges <- nb_to_edgelist <- function(nb) {
  n1 <- c(); n2 <- c()
  for (i in seq_along(nb)) {
    for (j in nb[[i]]) {
      if (j > i) {
        n1 <- c(n1, i)
        n2 <- c(n2, j)
      }
    }
  }
  list(node1 = n1, node2 = n2, N_edges = length(n1))
}

el <- nb_to_edgelist(nb_all)
cat("    Edges extracted:", el$N_edges, "\n")

# Computing the BYM2 scaling factor based on the combined spatial dataset
Q <- -nb2mat(nb_all, style = "B", zero.policy = TRUE)
diag(Q) <- -rowSums(Q)
Q_scaled <- INLA::inla.scale.model(Q, constr = list(A = matrix(1, 1, D_all), e = 0))
scaling_factor_all <- exp(mean(log(diag(Q_scaled))))


# Country index for each district (1=GHA, 2=CIV, 3=BFA)
country_vec <- all_sf$country_id  # length 724

# Re-index panels with global district IDs
# Each country's panel has: district_id (local 1:D), round_id (1 or 2),
# y_i, psi_i, observed, and covariates

remap_panel <- function(iso, offset) {
  panels[[iso]] %>%
    mutate(
      global_district_id = district_id + offset,
      country = iso
    )
}

panel_pooled <- bind_rows(
  remap_panel("GHA", 0),
  remap_panel("CIV", D_gha),
  remap_panel("BFA", D_gha + D_civ)
) %>%
  arrange(global_district_id, round_id)

NT <- nrow(panel_pooled)
cat("  Panel rows (NT):", NT, "(should be", D_all * 2, ")\n")

# Observation indices
obs_mask <- panel_pooled$observed == 1
obs_idx  <- which(obs_mask)
N_obs    <- sum(obs_mask)

cat("  Observed:", N_obs, "/ Total:", NT, "\n")

# Build X matrix from model panel (already standardised within country)
model_panel <- read_csv("output/three_country_model_panel.csv",
                         show_col_types = FALSE)

# Join covariates in the same order as panel_pooled
cov_names <- c("elev_std", "rainfall_std", "ntl_std", "temp_std",
               "travel_std", "pop_dens_std", "poverty_std")
K <- length(cov_names)

X_pooled <- panel_pooled %>%
  select(-any_of(cov_names)) %>% 
  left_join(model_panel %>% select(country, district_id, time, all_of(cov_names)),
            by = c("country", 
                   "district_id" = "district_id",
                   "round_id" = "time")) %>%
  select(all_of(cov_names)) %>%
  as.matrix()


cat("  X matrix:", nrow(X_pooled), "x", ncol(X_pooled), "\n")
cat("  NAs in X:", sum(is.na(X_pooled)), "\n")

# Assemble Stan data
stan_data_pooled <- list(
  N       = D_all,
  T       = 2L,
  NT      = NT,
  C       = 3L,
  K       = K,
  N_obs   = N_obs,
  obs_idx = obs_idx,
  y       = panel_pooled$y_i[obs_idx],
  psi     = panel_pooled$psi_i[obs_idx],
  district = panel_pooled$global_district_id,
  time     = panel_pooled$round_id,
  country  = country_vec,
  X        = X_pooled,
  N_edges  = el$N_edges,
  node1    = el$node1,
  node2    = el$node2,
  scaling_factor = scaling_factor_all
)

cat("\n  Stan data assembled.\n")
cat("    N =", stan_data_pooled$N, "\n")
cat("    NT =", stan_data_pooled$NT, "\n")
cat("    C =", stan_data_pooled$C, "\n")
cat("    K =", stan_data_pooled$K, "\n")
cat("    N_obs =", stan_data_pooled$N_obs, "\n")
cat("    N_edges =", stan_data_pooled$N_edges, "\n")


###############################################################################
# STEP 3: Compile and fit
###############################################################################

cat("\n", strrep("=", 70), "\n")
cat("  STEP 3: Fitting pooled model\n")
cat(strrep("=", 70), "\n\n")

mod_pooled <- cmdstan_model("fh_bym2_pooled_3country.stan")

mod_pooled$check_syntax(pedantic = TRUE)

fit_pooled <- mod_pooled$sample(
  data            = stan_data_pooled,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 1500,
  iter_sampling   = 2000,
  adapt_delta     = 0.97,
  max_treedepth   = 13,
  refresh         = 200
)


```
# Covariates Effects Across the Countries

```{r}
covar_labels <- c("Elevation", "Rainfall", "Nighttime lights", "Temperature",
                  "Travel time", "Pop. density (log)", "Poverty")

# Ensure this matches your Stan output exactly: 1=GHA, 2=CIV, 3=BFA
country_names <- c("GHA", "CIV", "BFA") 

# 1. Extract the country-specific slopes (using 0.05 and 0.95 for a 90% CI)
beta_pooled <- fit_pooled$summary("beta",
              mean = mean, 
              median = median,
              lower = ~ unname(quantile(.x, 0.05)),
              upper = ~ unname(quantile(.x, 0.95))) %>%
  tidyr::extract(variable, into = c("country_idx", "covariate_idx"), 
                 regex = "beta\\[(\\d+),(\\d+)\\]", convert = TRUE) %>%
  mutate(
    country = factor(country_names[country_idx], levels = c("GHA", "BFA", "CIV")),
    covariate = covar_labels[covariate_idx],
    # A covariate is significant if the 90% CI does not contain 0
    is_significant = (lower > 0) | (upper < 0),
    sig_marker = ifelse(is_significant, "*", "")
  )

# 2. Extract the global means (using 0.05 and 0.95 for a 90% CI)
beta_bar <- fit_pooled$summary("beta_bar",
              mean = mean,
              lower = ~ unname(quantile(.x, 0.05)),
              upper = ~ unname(quantile(.x, 0.95))) %>%
  mutate(
    covariate_idx = row_number(),
    covariate = covar_labels[covariate_idx],
    country = "Global Mean",
    is_significant = (lower > 0) | (upper < 0),
    sig_marker = ifelse(is_significant, "*", "")
  )

# 3. Factor the covariates to preserve plotting order
beta_pooled$covariate <- factor(beta_pooled$covariate, levels = rev(covar_labels))
beta_bar$covariate <- factor(beta_bar$covariate, levels = rev(covar_labels))

# 4. Generate the plot
p_pooled_covars <- ggplot(beta_pooled, aes(x = mean, y = covariate, colour = country, group = country)) +
  
  # LAYER 1: The deep background (Zero line)
  # Made this solid black and slightly thinner so it stands out against the grey bands
  geom_vline(xintercept = 0, linetype = "dashed", colour = "black", linewidth = 0.6) +
  
  # LAYER 2: THE GLOBAL MEAN (beta_bar) AS A GREY BACKGROUND BAND
  # This replaces the meaningless grid lines with the actual 90% CrI of the global average
  # geom_linerange(data = beta_bar, 
  #                aes(xmin = lower, xmax = upper, y = covariate),
  #                inherit.aes = FALSE, 
  #                colour = "grey85", linewidth = 5) + 
                 
  # Adds a diamond in the exact center of the grey global band
  geom_point(data = beta_bar, 
             aes(x = mean, y = covariate),
             inherit.aes = FALSE, 
             colour = "grey", shape = 18, size = 3) +
  
  # LAYER 3: The foreground (Country-specific estimates)
  geom_pointrange(aes(xmin = lower, xmax = upper),
                  position = position_dodge(width = 0.6), size = 0.8) +
                  
  # LAYER 4: The text/asterisks (Country level)
  geom_text(aes(x = upper + 0.02, label = sig_marker),
            position = position_dodge(width = 0.6),
            size = 7, vjust = 0.75, hjust = 0, show.legend = FALSE, colour = "black") +
            
  # Global Mean Asterisks
  geom_text(data = beta_bar, 
            aes(x = upper + 0.02, y = covariate, label = sig_marker),
            inherit.aes = FALSE,
            colour = "grey50", size = 7, vjust = 0.75, hjust = 0) +
              
  # Formatting
  scale_colour_brewer(palette = "Dark2", name = "Country") +
  scale_x_continuous(expand = expansion(mult = c(0.05, 0.15))) + 
  labs(x = "Coefficient (logit scale)", y = NULL,
       title = "Covariate effects: Pooled Hierarchical Model",
       subtitle = "Thick grey bands represent the Global Mean (90% CrI). Asterisks (*) denote significance.") +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.grid.major.y = element_blank(), # Keeps the background clean
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

# Save and print
ggsave("output/figures/pooled_covariate_effects.png", p_pooled_covars, width = 10, height = 7, dpi = 300, bg = "white")
p_pooled_covars
```


# Discussion

## Cross-Sectional vs Space-Time Comparison















<!-- ```{r} -->
<!-- # --- Shrinkage Plot: Direct vs Pooled Model Estimates --- -->

<!-- # 1. Isolate the observed districts for Round 2 and calculate precision -->
<!-- obs_panel <- panel_pooled %>% -->
<!--   # We only have direct estimates where observed == 1 -->
<!--   filter(observed == 1, round_id == 2) %>% -->
<!--   mutate( -->
<!--     direct_prev = plogis(y_i), -->
<!--     # Lower SE (psi_i) = Higher Precision (Proxy for Sample Size) -->
<!--     precision_proxy = 1 / psi_i -->
<!--   ) -->

<!-- # 2. Extract Posterior Medians for the exact same districts -->
<!-- model_prev <- fit_pooled$summary("prevalence", median = median) %>% -->
<!--   mutate(row_id = as.integer(gsub("prevalence\\[|\\]", "", variable))) %>% -->
<!--   # Map back to the full panel to get district info, then filter -->
<!--   left_join(panel_pooled %>% mutate(row_id = row_number()), by = "row_id") %>% -->
<!--   filter(observed == 1, round_id == 2) %>% -->
<!--   select(row_id, model_prev = median) -->

<!-- # 3. Join the datasets -->
<!-- shrinkage_df <- obs_panel %>% -->
<!--   left_join(model_prev, by = "row_id") -->

<!-- # 4. Generate the faceted Shrinkage Plot -->
<!-- p_shrinkage <- ggplot(shrinkage_df, aes(x = direct_prev, y = model_prev)) + -->
<!--   # The 1:1 Diagonal Line (No Shrinkage / Perfect Agreement) -->
<!--   geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "black", linewidth = 0.8) + -->

<!--   # Draw a faint segment from the point to the diagonal to explicitly highlight the "pull" -->
<!--   geom_segment(aes(xend = direct_prev, yend = direct_prev), alpha = 0.3, colour = "grey60") + -->

<!--   # The Data Points, colored by our sample size proxy -->
<!--   geom_point(aes(fill = precision_proxy), shape = 21, colour = "grey30", size = 3, alpha = 0.85) + -->

<!--   # Facet by country to create the small multiples -->
<!--   facet_wrap(~country, scales = "free") + -->

<!--   # Color Scale: High precision (large sample) = Dark Blue, Low precision = Yellow/Bright -->
<!--   scale_fill_viridis_c( -->
<!--     name = "Data Precision\n(1 / SE)", -->
<!--     option = "cividis", -->
<!--     direction = -1 -->
<!--   ) + -->
<!--   scale_x_continuous(labels = scales::percent_format(accuracy = 1)) + -->
<!--   scale_y_continuous(labels = scales::percent_format(accuracy = 1)) + -->
<!--   labs( -->
<!--     title = "Hierarchical Shrinkage: Joint Pooled Model vs Direct Estimates", -->
<!--     subtitle = "Yellow points (low precision/small sample) are pulled strongly toward the spatial mean. Dark points stay on the diagonal.", -->
<!--     x = "Direct Estimate (Noisy Survey Data)", -->
<!--     y = "Model Posterior Median (Smoothed)" -->
<!--   ) + -->
<!--   theme_minimal(base_size = 13) + -->
<!--   theme( -->
<!--     legend.position = "right", -->
<!--     strip.text = element_text(size = 12, face = "bold"), -->
<!--     panel.grid.minor = element_blank(), -->
<!--     panel.spacing = unit(1.5, "lines") -->
<!--   ) -->

<!-- ggsave("output/figures/pooled_shrinkage_plot.png", p_shrinkage, width = 12, height = 5, dpi = 300, bg = "white") -->
<!-- p_shrinkage -->

<!-- ``` -->


## Three-Country Covariate Maps (Raw Values)

```{r three-country-covariate-maps, fig.height=10, fig.width=14}
full_panel <- read_csv("output/three_country_full_panel.csv", show_col_types = FALSE)

# Fix the coastal extraction artifacts, then filter for Round 2
panel_r2_cov <- full_panel %>%
  # Group by country to calculate a safe replacement value
  group_by(country, time) %>%
  mutate(
    # Calculate the median rainfall for the country (ignoring the negative outliers)
    safe_rainfall = median(rainfall_sep_dec[rainfall_sep_dec >= 0], na.rm = TRUE),
    
    # If rainfall is negative, swap it with the safe regional median
    rainfall_sep_dec = ifelse(rainfall_sep_dec < 0, safe_rainfall, rainfall_sep_dec)
  ) %>%
  ungroup() %>%
  filter(time == 2) %>%
  select(country, district_id, district_name,
         nightlights, elev_mean_pw, rainfall_sep_dec, temp_sep_dec,
         travel_time_pw, census_pop_density, poverty_H)

# Join to the shapefile
covar_map_df <- all_sf %>%
  left_join(panel_r2_cov, by = c("country", "local_id" = "district_id"))

covar_map_df <- all_sf %>%
  left_join(panel_r2_cov, by = c("country", "local_id" = "district_id"))

country_borders <- all_sf %>%
  group_by(country) %>%
  summarise(geometry = st_union(geometry), .groups = "drop")

make_covar_map <- function(data, var, title, palette = "viridis",
                            direction = 1, trans = "identity") {
  ggplot() +
    geom_sf(data = data, aes(fill = .data[[var]]), colour = "grey50", linewidth = 0.05) +
    geom_sf(data = country_borders, fill = NA, colour = "black", linewidth = 0.8) +
    scale_fill_viridis_c(name = title, option = palette, direction = direction,
                          trans = trans, na.value = "grey80") +
    theme_void(base_size = 10) +
    labs(title = title) +
    theme(legend.position = "right",
          plot.title = element_text(size = 11, face = "bold"))
}

p_ntl  <- make_covar_map(covar_map_df, "nightlights", "Nightlights\n(nW/cm2/sr)", "magma")
p_elev <- make_covar_map(covar_map_df, "elev_mean_pw", "Elevation\n(m)", "terrain", direction = -1)
p_rain <- make_covar_map(covar_map_df, "rainfall_sep_dec", "Rainfall\nSep-Dec (mm)", "Blues")
p_temp <- make_covar_map(covar_map_df, "temp_sep_dec", "Temperature\nSep-Dec (C)", "inferno")
p_trav <- make_covar_map(covar_map_df, "travel_time_pw", "Travel time\n(minutes)", "YlOrRd")
p_pov  <- make_covar_map(covar_map_df, "poverty_H", "Poverty\nheadcount", "RdYlGn", direction = -1)

p_all_covars <- (p_ntl | p_elev | p_rain) / (p_temp | p_trav | p_pov) +
  plot_annotation(
    title = "District-level covariates across Ghana, Burkina Faso, and Cote d'Ivoire",
    subtitle = "Raw (non-standardised) values for Round 2. Heavy lines = country borders."
  )

ggsave("output/figures/three_country_covariates_all.png", p_all_covars,
       width = 14, height = 10, dpi = 300, bg = "white")
p_all_covars
```

## Nightlights vs Pooled Prevalence

```{r nightlights-vs-prevalence, fig.height=7, fig.width=14}
p_ntl_big <- ggplot() +
  geom_sf(data = covar_map_df, aes(fill = nightlights), colour = "grey50", linewidth = 0.05) +
  geom_sf(data = country_borders, fill = NA, colour = "black", linewidth = 0.8) +
  scale_fill_viridis_c(name = "Nightlights\n(nW/cm2/sr)", option = "magma",
                        trans = "log1p", na.value = "grey80") +
  labs(title = "Nightlights (VIIRS, log scale)") +
  theme_void(base_size = 12) +
  theme(legend.position = "right", plot.title = element_text(size = 13, face = "bold"))

pooled_prev_r2 <- fit_pooled$summary("prevalence", median = median) %>%
  mutate(row_id = as.integer(gsub("prevalence\\[|\\]", "", variable))) %>%
  left_join(panel_pooled %>% mutate(row_id = row_number()), by = "row_id") %>%
  filter(round_id == 2)

pooled_map_df <- all_sf %>%
  left_join(pooled_prev_r2, by = c("global_id" = "global_district_id"))

p_prev_big <- ggplot() +
  geom_sf(data = pooled_map_df, aes(fill = median), colour = "grey50", linewidth = 0.05) +
  geom_sf(data = country_borders, fill = NA, colour = "black", linewidth = 0.8) +
  scale_fill_viridis_c(name = "Prevalence",
                        labels = scales::percent_format(accuracy = 1),
                        option = "magma", direction = -1, na.value = "grey80") +
  labs(title = "Under-5 malaria prevalence (pooled model)") +
  theme_void(base_size = 12) +
  theme(legend.position = "right", plot.title = element_text(size = 13, face = "bold"))

p_ntl_vs_prev <- p_ntl_big | p_prev_big
ggsave("output/figures/nightlights_vs_prevalence.png", p_ntl_vs_prev,
       width = 14, height = 7, dpi = 300, bg = "white")
p_ntl_vs_prev
```

## Pooled Prevalence Map

```{r pooled-prevalence-map, fig.height=8, fig.width=10}
p_pooled_map <- ggplot() +
  geom_sf(data = pooled_map_df, aes(fill = median), colour = "grey40", linewidth = 0.1) +
  geom_sf(data = country_borders, fill = NA, colour = "black", linewidth = 0.8) +
  scale_fill_viridis_c(name = "Prevalence",
                        labels = scales::percent_format(accuracy = 1),
                        option = "magma", direction = -1) +
  labs(title = "Unified West African Under-5 Malaria Prevalence (2021/2022)",
       subtitle = "Posterior medians from the 724-district pooled hierarchical model") +
  theme_void(base_size = 14) +
  theme(legend.position = "right")

ggsave("output/figures/pooled_prevalence_map.png", p_pooled_map,
       width = 10, height = 8, dpi = 300, bg = "white")
p_pooled_map
```

## Three-Country Adjacency Graph

```{r three-country-adjacency, fig.height=8, fig.width=10}
centroids_all <- st_centroid(all_sf)
coords_all <- st_coordinates(centroids_all)

edges_all_df <- tibble(
  from = el$node1, to = el$node2,
  x_from = coords_all[el$node1, 1], y_from = coords_all[el$node1, 2],
  x_to   = coords_all[el$node2, 1], y_to   = coords_all[el$node2, 2],
  edge_type = ifelse(all_sf$country[el$node1] == all_sf$country[el$node2],
                     "Within-country", "Cross-border")
)

n_within <- sum(edges_all_df$edge_type == "Within-country")
n_cross  <- sum(edges_all_df$edge_type == "Cross-border")

p_adj_3country <- ggplot() +
  geom_sf(data = all_sf, fill = "grey95", colour = "grey80", linewidth = 0.05) +
  geom_segment(data = edges_all_df %>% filter(edge_type == "Within-country"),
               aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
               colour = "#93c5fd", alpha = 0.12, linewidth = 0.15) +
  geom_segment(data = edges_all_df %>% filter(edge_type == "Cross-border"),
               aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
               colour = "#dc2626", alpha = 0.85, linewidth = 0.6) +
  geom_sf(data = country_borders, fill = NA, colour = "black", linewidth = 1) +
  labs(title = paste0("Combined 724-district adjacency graph"),
       subtitle = paste0("Within-country edges (blue): ", n_within,
                          " | Cross-border edges (red): ", n_cross)) +
  theme_void(base_size = 12)

ggsave("output/figures/three_country_adjacency.png", p_adj_3country,
       width = 10, height = 8, dpi = 300, bg = "white")
p_adj_3country
```

## Pooled Estimates: CV Reliability Scatter

Scatter plot of pooled prevalence vs CV, coloured by reliability category, showing how the pooled model brings districts into the reliable region.

```{r cv-reliability-scatter, fig.height=6, fig.width=12}
# Extract prevalence and CV from the pooled model (Round 2)
pooled_full <- fit_pooled$summary("prevalence",
  mean = mean, median = median, sd = sd) %>%
  mutate(row_id = as.integer(gsub("prevalence\\[|\\]", "", variable))) %>%
  left_join(panel_pooled %>% mutate(row_id = row_number()), by = "row_id") %>%
  filter(round_id == 2) %>%
  mutate(
    cv = sd / mean,
    reliability = case_when(
      cv < 0.16 ~ "Reliable (CV < 16%)",
      cv < 0.33 ~ "Caution (CV 16-33%)",
      TRUE      ~ "Unreliable (CV > 33%)"
    ),
    reliability = factor(reliability, levels = c("Reliable (CV < 16%)",
                                                   "Caution (CV 16-33%)",
                                                   "Unreliable (CV > 33%)"))
  )

p_cv_reliability <- ggplot(pooled_full, aes(x = median, y = cv, colour = reliability)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_hline(yintercept = 0.16, linetype = "dashed", colour = "#16a34a", linewidth = 0.5) +
  geom_hline(yintercept = 0.33, linetype = "dashed", colour = "#dc2626", linewidth = 0.5) +
  scale_colour_manual(values = c("Reliable (CV < 16%)" = "#16a34a",
                                  "Caution (CV 16-33%)" = "#f59e0b",
                                  "Unreliable (CV > 33%)" = "#dc2626")) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~country) +
  labs(
    title = "Pooled model: prevalence vs coefficient of variation by reliability",
    subtitle = "Green dashed = 16% CV threshold. Red dashed = 33% threshold. Most districts now in the reliable region.",
    x = "Posterior median prevalence", y = "Coefficient of variation",
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 12, face = "bold"))

ggsave("output/figures/cv_reliability_scatter_pooled.png", p_cv_reliability,
       width = 12, height = 6, dpi = 300, bg = "white")
p_cv_reliability

# Print summary
pooled_full %>%
  count(country, reliability) %>%
  pivot_wider(names_from = reliability, values_from = n, values_fill = 0) %>%
  print()
```

## Pooled CV Map

```{r pooled-cv-map, fig.height=8, fig.width=10}
pooled_cv_df <- pooled_full %>%
  select(global_district_id, cv)

pooled_cv_map <- all_sf %>%
  left_join(pooled_cv_df, by = c("global_id" = "global_district_id"))

p_pooled_cv <- ggplot() +
  geom_sf(data = pooled_cv_map, aes(fill = cv), colour = "grey40", linewidth = 0.1) +
  geom_sf(data = country_borders, fill = NA, colour = "black", linewidth = 0.8) +
  scale_fill_viridis_c(name = "CV",
                        labels = scales::percent_format(accuracy = 1),
                        option = "cividis", direction = -1) +
  labs(title = "Coefficient of Variation: Pooled 3-Country Model (Round 2)",
       subtitle = "Low CV (dark) = high confidence. High CV (bright) = greater uncertainty.") +
  theme_void(base_size = 14) +
  theme(legend.position = "right")

ggsave("output/figures/pooled_cv_map.png", p_pooled_cv,
       width = 10, height = 8, dpi = 300, bg = "white")
p_pooled_cv
```




```{r hh}
library(dplyr)
library(ggplot2)

# 1. Merge the direct and pooled dataframes using inner_join
# This will only keep the districts that actually had survey data
cv_comparison <- cv_df %>%
  select(district_id, country, cv_direct = cv) %>% # Ensure district_id exists in cv_df
  inner_join(
    pooled_full %>% select(district_id, cv_pooled = cv), 
    by = c("district_id") # Adjust join keys if you use row_id instead
  )

# 2. Generate the Variance Reduction Plot
p_cv_comparison <- ggplot(cv_comparison, aes(x = cv_direct, y = cv_pooled)) +
  
  # The 45-degree line: Anything below this line represents a reduction in variance
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey40", linewidth = 1) +
  
  # The 20% Policy Threshold lines
  geom_vline(xintercept = 0.20, colour = "#e3342f", linetype = "dotted", linewidth = 0.8) +
  geom_hline(yintercept = 0.20, colour = "#10b981", linetype = "dotted", linewidth = 0.8) +
  
  # The data points, colored by country
  geom_point(aes(fill = country), shape = 21, colour = "white", size = 3, alpha = 0.7) +
  
  scale_fill_brewer(palette = "Dark2", name = "Country") +
  
  # Force both axes to have the exact same scale and limits so the geometry is honest
  scale_x_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  
  # Annotations
  annotate("text", x = 0.75, y = 0.15, label = "Variance Reduced", 
           colour = "#10b981", fontface = "bold", size = 5) +
  
  labs(
    title = "Variance Reduction: Direct vs. Pooled CV",
    subtitle = "Points below the dashed line indicate precision gained by the hierarchical model.",
    x = "Direct Estimate CV (Survey Data)",
    y = "Pooled Model CV (Hierarchical Smoothing)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")

ggsave("output/figures/variance_reduction_45deg.png", p_cv_comparison, width = 8, height = 8, dpi = 300, bg = "white")
p_cv_comparison
```

```{r jn}
cv_df <- direct_est_all %>%
  filter(round_id == 2) %>%
  mutate(
    p_hat = plogis(y_i),
    se_p = psi_i * p_hat * (1 - p_hat),
    cv = se_p / p_hat,
    
    # UNIFIED THRESHOLDS
    reliability = case_when(
      cv < 0.16 ~ "Reliable (CV < 16%)",
      cv < 0.33 ~ "Caution (CV 16-33%)",
      TRUE      ~ "Unreliable (CV > 33%)"
    ),
    reliability = factor(reliability, levels = c(
      "Reliable (CV < 16%)", 
      "Caution (CV 16-33%)", 
      "Unreliable (CV > 33%)"
    ))
  )

# 2. Generate the faceted scatter plot
p_cv_direct <- ggplot(cv_df, aes(x = p_hat, y = cv, fill = reliability)) +
  geom_hline(yintercept = 0.16, linetype = "dashed", colour = "#16a34a", linewidth = 0.8) +
  geom_hline(yintercept = 0.33, linetype = "dashed", colour = "#dc2626", linewidth = 0.8) +
  
  geom_point(shape = 21, colour = "grey20", size = 2.5, alpha = 0.85) +
  facet_wrap(~country, ncol = 3) +
  
  scale_fill_manual(
    name = "Estimate Reliability",
    values = c(
      "Reliable (CV < 16%)" = "#16a34a",
      "Caution (CV 16-33%)" = "#f59e0b",
      "Unreliable (CV > 33%)" = "#dc2626"
    )
  ) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  # LOCKED Y-AXIS (0 to 1)
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  
  labs(
    title = "Direct Estimates: High Uncertainty (Survey Data Only)",
    subtitle = "Raw estimates before spatial smoothing.",
    x = "Direct Prevalence Estimate",
    y = "Coefficient of Variation"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold")
  )

ggsave("output/figures/direct_cv_scatter_fixed.png", p_cv_direct, width = 12, height = 5, dpi = 300, bg = "white")
```

```{r kn}
# --- Plot 2: Pooled Estimates (Fixed Axes and Thresholds) ---

# 1. Process the posterior pooled data
pooled_full <- fit_pooled$summary("prevalence",
  mean = mean, median = median, sd = sd) %>%
  mutate(row_id = as.integer(gsub("prevalence\\[|\\]", "", variable))) %>%
  left_join(panel_pooled %>% mutate(row_id = row_number()), by = "row_id") %>%
  filter(round_id == 2) %>%
  mutate(
    cv = sd / mean,
    
    # UNIFIED THRESHOLDS (Matches Direct plot exactly)
    reliability = case_when(
      cv < 0.16 ~ "Reliable (CV < 16%)",
      cv < 0.33 ~ "Caution (CV 16-33%)",
      TRUE      ~ "Unreliable (CV > 33%)"
    ),
    reliability = factor(reliability, levels = c(
      "Reliable (CV < 16%)",
      "Caution (CV 16-33%)",
      "Unreliable (CV > 33%)"
    ))
  )

# 2. Generate the faceted scatter plot
p_cv_pooled <- ggplot(pooled_full, aes(x = median, y = cv, fill = reliability)) +
  geom_hline(yintercept = 0.16, linetype = "dashed", colour = "#16a34a", linewidth = 0.8) +
  geom_hline(yintercept = 0.33, linetype = "dashed", colour = "#dc2626", linewidth = 0.8) +
  
  geom_point(shape = 21, colour = "grey20", size = 2.5, alpha = 0.85) +
  facet_wrap(~country, ncol = 3) +
  
  scale_fill_manual(
    name = "Estimate Reliability",
    values = c(
      "Reliable (CV < 16%)" = "#16a34a",
      "Caution (CV 16-33%)" = "#f59e0b",
      "Unreliable (CV > 33%)" = "#dc2626"
    )
  ) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  # LOCKED Y-AXIS (0 to 1)
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  
  labs(
    title = "Pooled Model: Dramatic Reduction in Uncertainty",
    subtitle = "Hierarchical smoothing rescues unobserved and noisy districts.",
    x = "Posterior Median Prevalence",
    y = "Coefficient of Variation",
    fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold")
  )

ggsave("output/figures/cv_reliability_scatter_pooled_fixed.png", p_cv_pooled, width = 12, height = 5, dpi = 300, bg = "white")
```

```{r mhn}
# 1. Categorize all districts based on strict StatCan guidelines
rescue_df <- pooled_full %>% 
  filter(round_id == 2) %>% 
  select(district_id, country, cv_pooled = cv) %>%
  left_join(
    cv_df %>% filter(round_id == 2) %>% select(district_id, cv_direct = cv), 
    by = "district_id"
  ) %>%
  mutate(
    # Categorize the "Before" state (Raw DHS Data)
    status_before = case_when(
      is.na(cv_direct)   ~ "1. No Data (Unobserved)",
      cv_direct <= 0.165 ~ "2. Acceptable (CV ≤ 16.5%)",
      cv_direct <= 0.333 ~ "3. Marginal (CV 16.6-33.3%)",
      TRUE               ~ "4. Unacceptable (CV > 33.3%)"
    ),
    # Categorize the "After" state (Pooled Model)
    status_after = case_when(
      cv_pooled <= 0.165 ~ "Acceptable (CV ≤ 16.5%)",
      cv_pooled <= 0.333 ~ "Marginal (CV 16.6-33.3%)",
      TRUE               ~ "Unacceptable (CV > 33.3%)"
    )
  )

# 2. Generate the Transition Matrix Table
transition_matrix <- rescue_df %>%
  count(status_before, status_after) %>%
  pivot_wider(names_from = status_after, values_from = n, values_fill = 0) %>%
  arrange(status_before)

print(transition_matrix)

# 3. Calculate the "Hero Metric" (Rescued into the Acceptable tier)
rescued_districts <- rescue_df %>%
  filter(
    status_before %in% c("1. No Data (Unobserved)", "3. Marginal (CV 16.6-33.3%)", "4. Unacceptable (CV > 33.3%)") & 
    status_after == "Acceptable (CV ≤ 16.5%)"
  ) %>%
  nrow()

cat("\nTotal districts rescued into the Acceptable zone:", rescued_districts, "\n")
```



```{r hjn}
prev_draws <- fit_pooled$draws("prevalence", format = "df") %>%
  as_tibble() %>%
  select(starts_with("prevalence["), .draw) %>%
  pivot_longer(starts_with("prevalence["), names_to = "variable", values_to = "prev") %>%
  mutate(row_id = as.integer(gsub("prevalence\\[|\\]", "", variable)))

# 2. Join the draws to your panel data to get country, time, and population weights
prev_full <- prev_draws %>%
  left_join(full_panel %>% mutate(row_id = row_number()), by = "row_id")

# 3. Calculate the population-weighted national prevalence for EVERY SINGLE DRAW
national_posteriors <- prev_full %>%
  group_by(.draw, country, time) %>%
  summarise(
    national_prev = weighted.mean(prev, w = census_pop, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Round = factor(ifelse(time == 1, "Round 1 (Baseline)", "Round 2 (Follow-up)"),
                   levels = c("Round 1 (Baseline)", "Round 2 (Follow-up)"))
  )

# 4. Generate the Posterior Boxplot
p_national_post <- ggplot(national_posteriors, aes(x = national_prev, y = country, fill = Round)) +
  
  # Grouped boxplots to show Round 1 and Round 2 side-by-side
  geom_boxplot(
    alpha = 0.85, 
    position = position_dodge(width = 0.75), 
    outlier.alpha = 0.05, 
    outlier.size = 0.5,
    colour = "grey20"
  ) +
  
  # Custom colors (e.g., Gold for Baseline, Blue for Follow-up)
  scale_fill_manual(
    values = c("Round 1 (Baseline)" = "#e6b000", "Round 2 (Follow-up)" = "#3182bd")
  ) +
  
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  
  labs(
    title = "Posterior Uncertainty: National Malaria Prevalence",
    subtitle = "Population-weighted aggregation of 724 district-level posterior draws.",
    x = "Estimated Prevalence",
    y = NULL,
    fill = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

ggsave("output/figures/posterior_boxplot_prevalence.png", p_national_post, width = 10, height = 6, dpi = 300, bg = "white")
```

