---
title: "Cross-Border Bayesian Small Area Estimation of Under-Five Malaria Prevalence in West Africa: A Spatiotemporal BYM2 Fay-Herriot Model Across Ghana, Burkina Faso and Côte d'Ivoire"
author: ''
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
  pdf_document:
    toc: true
  word_document:
    toc: true
subtitle: "A Space-Time BYM2 Fay-Herriot Model Across Ghana, Burkina Faso, and Côte
  d'Ivoire"
bibliography: references.bib
link-citations: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6,
                      cache       = TRUE,
                      cache.path  = "bym2_cache/",
                      cache.lazy  = FALSE,
                      autodep     = TRUE) 
# Core libraries
library(tidyverse)
library(haven)
library(sf)
library(geodata)
library(survey)
library(spdep)
library(Matrix)
library(cmdstanr)
library(bayesplot)
library(tidybayes)
library(patchwork)
library(ggcorrplot)
library(posterior)

# ---- University of Galway brand palette ------------------------------
# Source: Brand Guidelines v13, primary + secondary palettes.
# Use galway_pal()  for categorical scales
# Use galway_grad() for continuous (sequential) scales
# Use galway_div()  for diverging scales (e.g. residuals, anomalies)
galway_maroon       <- "#a80050"   # primary
galway_deep_maroon  <- "#84003d"   # primary
galway_magenta      <- "#e6007e"   # primary accent
galway_aqua         <- "#00b1ba"   # secondary
galway_ocean        <- "#0080a3"   # secondary
galway_green        <- "#00bf6f"   # secondary
galway_lime         <- "#c5e86c"   # secondary
galway_orange       <- "#f58e2f"   # secondary
galway_yellow       <- "#ffc72b"   # secondary
galway_violet       <- "#753abd"   # secondary
galway_ash          <- "#5e514d"   # neutral
galway_stone        <- "#94795d"   # neutral
galway_sandstone    <- "#c4c0b7"   # neutral

# 3-class categorical (e.g. Country = GHA, BFA, CIV)
galway_three <- c(galway_maroon, galway_ocean, galway_green)

# Generic categorical palette (up to 8 distinct hues)
galway_pal <- function(n = 8) {
  base <- c(galway_maroon, galway_ocean, galway_green, galway_orange,
            galway_violet, galway_aqua, galway_yellow, galway_deep_maroon)
  base[seq_len(min(n, length(base)))]
}

# Sequential gradient for prevalence / continuous outcomes
# (light sandstone -> Galway maroon)
galway_grad <- function() {
  ggplot2::scale_fill_gradientn(
    colours = c("#f7eaef", "#e9b9cb", "#d271a0", "#b03073", galway_deep_maroon),
    na.value = "grey90"
  )
}

# Sequential gradient for choropleth (prevalence) maps. We use magma here
# rather than galway_grad() because magma is perceptually uniform and gives
# much stronger luminance contrast across the (0, 1) range than a
# single-hue maroon gradient, which matters specifically for area-fill maps
# where small differences need to be visible at a glance. galway_grad() is
# retained for non-map continuous fills.
prev_grad <- function(name = "Prevalence",
                     labels = scales::percent_format(accuracy = 1),
                     na.value = "grey90", direction = 1) {
  ggplot2::scale_fill_viridis_c(
    name = name, option = "magma", direction = direction,
    labels = labels, na.value = na.value
  )
}

# Diverging scale for residual/deviation maps that need to centre on zero.
prev_div <- function(name = "Deviation", midpoint = 0,
                    na.value = "grey90") {
  ggplot2::scale_fill_gradient2(
    low = "#0080a3", mid = "white", high = "#a80050",
    midpoint = midpoint, name = name, na.value = na.value
  )
}

# Diverging scale: ocean blue <- white -> Galway maroon
galway_div <- function(name = "value", midpoint = 0) {
  ggplot2::scale_fill_gradient2(
    low = galway_ocean, mid = "white", high = galway_maroon,
    midpoint = midpoint, name = name, na.value = "grey90"
  )
}

# A ggplot2 theme to match Galway publication style
theme_galway <- function(base_size = 11) {
  ggplot2::theme_minimal(base_size = base_size) +
    ggplot2::theme(
      plot.title    = ggplot2::element_text(face = "bold", colour = galway_deep_maroon),
      plot.subtitle = ggplot2::element_text(colour = galway_ash),
      strip.text    = ggplot2::element_text(face = "bold", colour = galway_deep_maroon),
      panel.grid.minor = ggplot2::element_blank()
    )
}
```

```{r setup-galway-bayesplot, include=FALSE}
# Apply the Galway maroon to bayesplot defaults
bayesplot::color_scheme_set(scheme = c(
  "#f7eaef", "#e9b9cb", "#d271a0", "#b03073", "#a80050", "#84003d"
))


```{r fit-persistence-helpers, include=FALSE}
dir.create("output/fits",    showWarnings = FALSE, recursive = TRUE)
dir.create("output/figures", showWarnings = FALSE, recursive = TRUE)

persist_fit <- function(fit, name) {
  path <- file.path("output/fits", paste0(name, ".rds"))
  tryCatch(
    fit$save_object(file = path),
    error = function(e) message("Could not persist fit ", name, ": ", conditionMessage(e))
  )
  invisible(fit)
}

is_fit_live <- function(fit) {
  if (is.null(fit)) return(FALSE)
  files <- tryCatch(fit$output_files(), error = function(e) character())
  length(files) > 0L && all(file.exists(files))
}

restore_fit <- function(name) {
  path <- file.path("output/fits", paste0(name, ".rds"))
  if (file.exists(path)) readRDS(path) else NULL
}

ensure_fit <- function(fit, name) {
    if (is_fit_live(fit)) {
    persist_fit(fit, name)
    return(fit)
  }
  restored <- restore_fit(name)
  if (!is.null(restored)) return(restored)
  warning("Fit `", name, "` has no live CSVs and no saved .rds; ",
          "subsequent calls may fail. Re-run the chunk that fits it.")
  fit
}
```

```{r param-summary-helper, include=FALSE}

# Function to pull out model summary statistics when ever a new model is fitted.
# This is done after realising that we will need to continously pull out summary statistics time and again and a helper function becomes handy
DISTRICT_LEVEL <- c("theta", "b", "u", "u_raw", "v", "prevalence",
                    "gamma", "gamma_raw", "gamma_star",
                    "alpha_raw", "beta_raw", "log_lik", "y_rep", "lp__")

.tidy_param_table <- function(draws, digits = 3) {
  tab <- posterior::summarise_draws(
    draws,
    mean, median, sd,
    ~ quantile(.x, probs = c(0.05, 0.95), names = FALSE),
    posterior::default_convergence_measures()
  )
  tab <- as.data.frame(tab)
  if ("5%"  %in% names(tab)) names(tab)[names(tab) == "5%"]  <- "q5"
  if ("95%" %in% names(tab)) names(tab)[names(tab) == "95%"] <- "q95"
  numc <- vapply(tab, is.numeric, logical(1))
  tab[numc] <- lapply(tab[numc], function(x) round(x, digits))
  tab
}

param_summary <- function(fit, label = NULL,
                          keep = c("mu", "sigma_b", "phi", "sigma_gamma",
                                   "alpha", "delta", "delta_bar", "sigma_delta",
                                   "delta_R2_R1", "odds_ratio_R2_R1",
                                   "delta_2", "delta2",  # legacy wave-indicator
                                   "beta", "beta_bar", "tau",
                                   "p_national", "p_round1", "p_round2",
                                   "odds_ratio"),
                          digits = 3) {
  if (!is.null(label)) cat("\n==================== ", label,
                           " ====================\n", sep = "")
  vars <- tryCatch(fit$metadata()$stan_variables, error = function(e) character())
  keep_vars <- setdiff(intersect(keep, vars), DISTRICT_LEVEL)
  if (length(keep_vars) == 0L) {
    cat("  [no model-level parameters matched for this fit]\n")
    return(invisible(NULL))
  }
  tab <- tryCatch(
    .tidy_param_table(fit$draws(variables = keep_vars), digits = digits),
    error = function(e) { cat("  [summary failed:", conditionMessage(e), "]\n"); NULL }
  )
  if (is.null(tab)) return(invisible(NULL))
  print(tab, row.names = FALSE)
  invisible(tab)
}
```


# Data {#sec:data}

## About the Demographic and Health Surveys (DHS)

The Demographic and Health Surveys are nationally representative household
surveys using a stratified two-stage cluster sampling design. The first
stage selects enumeration areas (clusters) within strata defined by
region and urban/rural status; the second stage selects households within
each cluster. Children aged 6--59 months are tested for malaria parasitaemia
using a rapid diagnostic test (RDT) [@florey2014; @croft2018]. Cluster
GPS coordinates are jittered by up to 5 km (urban) or 10 km (rural) for
respondent confidentiality. The household member recode (PR) file
contains the individual-level biomarker; the cluster GPS shapefile
provides the spatial reference.



## Surveys Used in This Study

| Country | Round 1 | Round 2 |
|---|---|---|
| Burkina Faso | DHS 2010 | DHS/MIS 2021 |
| Côte d'Ivoire | DHS 2011 | DHS/MIS 2021 |
| Ghana | DHS 2014 | DHS 2022 |

<!-- Because the DHS data needs formal approve before being used and as we are not allowed to share to anyone who is not involved in this project, for reproducibility of our work, you need to make your own formal request to the dhs program. the file names are defined in the coming sections. the github for this project does not contain any data from the DHS. -->

## District Boundaries

```{r district-boundaries}

# this is a helper function for pulling district boundaries from GADM.
# Alternatively one can use any other shapefiles if they have them locally in their drives. 

# for ghana we pull admin level 2 boundaries while for the other two countries we pull level 3. this is because for level 2 of the other two countries its regional level since they have different definations of districts, communes or regions. Admin level 3 for BK and CIV matches amdin level 2 for Ghana. THis though affects the census covariates as the covairates data for BK and CIV is available mainly at admin 2 level. More is talked about in the final report
load_districts <- function(iso) {
  if (iso == "GHA") {
    gadm(country = iso, level = 2, path = "boundaries/") %>%
      st_as_sf() %>% st_transform(crs = 4326) %>%
      rename(district_name = NAME_2, district_code = GID_2) %>%
      mutate(iso = iso)
  } else {
    gadm(country = iso, level = 3, path = "boundaries/") %>%
      st_as_sf() %>% st_transform(crs = 4326) %>%
      rename(district_name = NAME_3, district_code = GID_3) %>%
      mutate(iso = iso)
  }
}

districts_raw <- list(
  GHA = load_districts("GHA"),
  BFA = load_districts("BFA"),
  CIV = load_districts("CIV")
)

# Create numeric district lookups (Stan requires integer indices)
district_lookups <- map(districts_raw, function(d) {
  d %>% st_drop_geometry() %>%
    select(district_name, district_code, iso) %>%
    mutate(district_id = row_number())
})

# Districts ordered by district_id for adjacency matrix construction
districts_ordered <- map2(districts_raw, district_lookups, function(d, lk) {
  d %>% left_join(lk, by = c("district_name", "district_code", "iso")) %>%
    arrange(district_id)
})

map_dfr(district_lookups, ~tibble(districts = nrow(.x)), .id = "iso")
```

## Survey Index and PR Files

<!-- for reproducibility of these results the user needs to create a a different folder for each country and each DHS survey round as defined below. The files names to download when granted access to  the DHS data are as defined in the blaock of file paths below. -->

```{r survey-index}
survey_index <- tribble(
  ~country,        ~iso, ~year, ~round_id, ~pr_path, ~gps_path,
  "Ghana",         "GHA", 2014, 1L,
  "dhs_gh_14/GHPR72DT/GHPR72FL.DTA", "dhs_gh_14/GHGE71FL/GHGE71FL.shp",
  "Ghana",         "GHA", 2022, 2L,
  "dhs_gh_22/GHPR8CDT/GHPR8CFL.DTA", "dhs_gh_22/GHGE8AFL/GHGE8AFL.shp",
  "Burkina",       "BFA", 2010, 1L,
  "dhs_bf_10/BFPR62DT/BFPR62FL.DTA", "dhs_bf_10/BFGE61FL/BFGE61FL.shp",
  "Burkina",       "BFA", 2021, 2L,
  "dhs_bf_21/BFPR81DT/BFPR81FL.DTA", "dhs_bf_21/BFGE81FL/BFGE81FL.shp",
  "Cote d'Ivoire", "CIV", 2012, 1L,
  "dhs_ci_12/CIPR62DT/CIPR62FL.DTA", "dhs_ci_12/CIGE61FL/CIGE61FL.shp",
  "Cote d'Ivoire", "CIV", 2021, 2L,
  "dhs_ci_21/CIPR81DT/CIPR81FL.DTA", "dhs_ci_21/CIGE81FL/CIGE81FL.shp"
)
```

```{r pr-files}

# helper functon to load the dhs files into the project.

# noted this works if your file paths are defined as above in your project folder otherwise this function will not work and you will need to load them indiviaully which can be draining
load_pr <- function(pr_path, gps_path, iso, year, round_id,
                    districts_ordered, district_lookup) {
  gps <- read_sf(gps_path) %>%
    rename(cluster_id = DHSCLUST) %>%
    st_transform(crs = 4326)
  
  gps_joined <- st_join(gps, districts_ordered, join = st_nearest_feature) %>%
    select(cluster_id, district_name, district_code) %>%
    st_drop_geometry() %>%
    left_join(district_lookup, by = c("district_name", "district_code")) %>%
    select(cluster_id, district_id, district_name)
  
  pr <- read_dta(pr_path) %>%
    rename_with(tolower) %>%
    select(cluster_id = hv001, hh_id = hv002, line_id = hvidx,
           weight_raw = hv005, stratum = hv023,
           age_months = hc1, malaria = hml32) %>%
    filter(age_months < 60, malaria %in% c(0, 1)) %>%
    mutate(weight = weight_raw / 1e6,
           malaria = as.integer(malaria),
           cluster_id = as.integer(cluster_id),
           stratum = as.integer(stratum),
           country = iso, year = year, round_id = round_id) %>%
    left_join(gps_joined, by = "cluster_id")
  
  pr
}

pr_list <- pmap(survey_index, function(country, iso, year, round_id, pr_path, gps_path) {
  load_pr(pr_path, gps_path, iso, year, round_id,
          districts_ordered[[iso]], district_lookups[[iso]])
})

```

# Direct Estimation {#sec:direct}

For each observed district $i$ at time $t$, the design-based prevalence
is computed using the Hájek ratio estimator [@hajek1971],
$$\hat{p}_{it}^{w} = \frac{\sum_k d_k \, z_k}{\sum_k d_k},$$
where $z_k$ is the binary RDT outcome for child $k$ and $d_k$ is the
DHS sampling weight. Design-based standard errors are obtained via
Taylor-series linearisation [@binder1983] using the `survey` package
[@lumley2010], respecting the stratified two-stage cluster design.

$$\beta + \sigma_{i}$$


For the Fay-Herriot likelihood [@wakefield2020], the prevalence is
transformed to the logit scale and the delta method is applied to the
SE:
$$y_{it} = \mathrm{logit}(\hat{p}_{it}^{w}), \qquad
\widehat{\mathrm{SE}}(y_{it}) =
  \frac{\widehat{\mathrm{SE}}(\hat{p}_{it}^{w})}
       {\hat{p}_{it}^{w}(1-\hat{p}_{it}^{w})}.$$
For numerical stability, prevalences are bounded to $[0.02, 0.98]$ and
logit-scale SEs to $[0.1, 2.0]$.

```{r direct-estimates}

# we use the syvdesign packed to calculate the direct estimated defined above
# the function assumes that you have one single dataset with all the 3 countries
get_direct_estimates <- function(pr) {
  design <- svydesign(
    ids = ~cluster_id, strata = ~stratum, weights = ~weight,
    data = pr, nest = TRUE
  )
  svyby(formula = ~malaria, by = ~district_id, design = design,
        FUN = svymean, na.rm = TRUE, keep.var = TRUE) %>%
    rename(p_hat = malaria, se_p = se) %>%
    filter(!is.na(p_hat), !is.na(se_p),
           se_p > 0, p_hat > 0, p_hat < 1) %>%
    mutate(
      country   = pr$country[1],
      iso       = pr$country[1],
      year      = pr$year[1],
      round_id  = pr$round_id[1],
      p_bounded = pmax(0.02, pmin(0.98, p_hat)),
      y_i       = log(p_bounded / (1 - p_bounded)),
      logit_var = se_p^2 / (pmax(0.02, p_hat)^2 * pmax(0.02, 1 - p_hat)^2),
      psi_i     = pmax(0.1, pmin(sqrt(logit_var), 2.0))
    ) %>%
    filter(is.finite(y_i), is.finite(psi_i)) %>%
    select(country, iso, year, round_id, district_id, y_i, psi_i)
}

direct_est_list <- map(pr_list, get_direct_estimates)
direct_est_all  <- bind_rows(direct_est_list)

direct_est_all %>% count(country, year, name = "observed_districts")

```

# Panel Construction {#sec:panel}

```{r build-panel}
build_panel <- function(country_iso, district_lookup, direct_est_all) {
  rounds <- direct_est_all %>%
    filter(country == country_iso) %>%
    distinct(year) %>% arrange(year) %>%
    mutate(round_id = row_number())
  
  est_clean <- direct_est_all %>%
    filter(country == country_iso) %>%
    select(district_id, year, y_i, psi_i)
  
  expand_grid(
    district_id = district_lookup$district_id,
    year        = rounds$year
  ) %>%
    left_join(district_lookup %>% select(district_id, district_name, district_code),
              by = "district_id") %>%
    left_join(est_clean, by = c("district_id", "year")) %>%
    left_join(rounds, by = "year") %>%
    arrange(year, district_id) %>%
    mutate(country = country_iso,
           observed = as.integer(!is.na(y_i)))
}

panels <- map(c("GHA", "BFA", "CIV"),
              ~build_panel(.x, district_lookups[[.x]], direct_est_all))
names(panels) <- c("GHA", "BFA", "CIV")

map_dfr(panels, function(p) {
  p %>% group_by(year) %>%
    summarise(total = n(), observed = sum(observed == 1),
              pct = round(100 * observed / total, 1), .groups = "drop")
}, .id = "country")
```

# Spatial Adjacency and BYM2 Scaling Factor {#sec:adjacency}

District adjacency is defined via queen contiguity using the `spdep`
package [@bivand2013]. Disconnected components are reconnected via a
snap tolerance to ensure a proper ICAR prior [@freni2018]. The BYM2
scaling factor $s$ [@riebler2016] makes the mixing parameter $\phi$
interpretable as the proportion of structured variance regardless of
graph topology.

```{r adjacency-scaling}
compute_scaling_factor <- function(nb, D) {
  adj_mat <- nb2mat(nb, style = "B", zero.policy = TRUE)
  Q_icar  <- Diagonal(x = rowSums(adj_mat)) - as(adj_mat, "sparseMatrix")
  eps <- 0.001
  Q_reg <- Q_icar + eps * Diagonal(D)
  Q_inv_diag <- diag(solve(Q_reg))
  exp(mean(log(Q_inv_diag)))
}

build_spatial <- function(iso, districts_ordered) {
  D <- nrow(districts_ordered)
  nb <- poly2nb(districts_ordered, queen = TRUE)
  if (sum(sapply(nb, length) == 0) > 0) {
    nb <- poly2nb(districts_ordered, queen = TRUE, snap = 0.1)
  }
  
  adj_mat <- nb2mat(nb, style = "B", zero.policy = TRUE)
  edges <- which(adj_mat == 1, arr.ind = TRUE) %>%
    as.data.frame() %>%
    filter(row < col)
  
  scaling_factor <- compute_scaling_factor(nb, D)
  
  list(D = D, nb = nb,
       node1 = edges$row, node2 = edges$col,
       N_edges = nrow(edges),
       scaling_factor = scaling_factor)
}

spatial <- map(c("GHA", "BFA", "CIV"),
               ~build_spatial(.x, districts_ordered[[.x]]))
names(spatial) <- c("GHA", "BFA", "CIV")

map_dfr(spatial, function(s) {
  tibble(N = s$D, N_edges = s$N_edges, scaling_factor = round(s$scaling_factor, 3))
}, .id = "country")
```

# Covariates: Harmonised Multi-Country Panel {#sec:covariates}

Seven covariates are assembled at the district-round level: elevation
(SRTM, population-weighted), rainfall (CHIRPS v2.0, Sep--Dec mean),
nighttime lights (NASA Black Marble VIIRS) [@elvidge2014], temperature
(CRU TS v4.09, Sep--Dec mean), travel time (Weiss/MAP 2020,
population-weighted), log population density (WorldPop 2020), and
multidimensional poverty (OPHI MPI / national census headcount). All
covariates are standardised within country. Pairwise correlations are
kept below 0.7 [@dormann2013].

```{r load-covariates}
covariates_panel <- read_csv("three_country_model_panel.csv") %>%
  select(country, district_id, district_name, time, year, label,
         elev_std, rainfall_std, ntl_std, temp_std,
         travel_std, pop_dens_std, poverty_std)

covariates_panel %>% count(country, time, name = "rows")
```

```{r join-covariates}
join_covariates <- function(iso) {
  panels[[iso]] %>%
    left_join(
      covariates_panel %>%
        filter(country == iso) %>%
        select(district_id, time,
               elev_std, rainfall_std, ntl_std, temp_std,
               travel_std, pop_dens_std, poverty_std),
      by = c("district_id", "round_id" = "time")
    )
}

panels <- list(
  GHA = join_covariates("GHA"),
  BFA = join_covariates("BFA"),
  CIV = join_covariates("CIV")
)

# Verify covariates are present
panels[["GHA"]] %>% select(district_id, year, ends_with("_std")) %>% head()
```

# Exploratory Data Analysis {#sec:eda}

## Correlation Heatmap

```{r correlation-heatmap, fig.height=7, fig.width=8}
covariate_cols <- c("elev_std", "rainfall_std", "ntl_std", "temp_std",
                    "travel_std", "pop_dens_std", "poverty_std")

cor_matrix <- covariates_panel %>%
  select(all_of(covariate_cols)) %>%
  cor(use = "complete.obs")

ggcorrplot(
  cor_matrix,
  type = "lower",
  lab = TRUE, lab_size = 3.5,
  colors = c(galway_ocean, "white", galway_maroon),
  title = "Pairwise correlations among standardised covariates",
  ggtheme = ggplot2::theme_minimal()
) +
  theme(plot.title = element_text(hjust = 0.5))
```

## Distribution of Direct Estimates

```{r direct-est-dist}
direct_est_all %>%
  mutate(prevalence = plogis(y_i)) %>%
  ggplot(aes(x = factor(year), y = prevalence)) +
  geom_violin(fill = "steelblue", alpha = 0.4) +
  geom_jitter(width = 0.1, alpha = 0.4, size = 0.5) +
  facet_wrap(~country, scales = "free_x") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = NULL, y = "Direct prevalence estimate (district)",
       title = "Distribution of district-level direct estimates by country and round",
       subtitle = "Each point is one district; violin shows the distribution") +
  theme_minimal(base_size = 11)
```

## Adjacency Graph (Ghana)

```{r adjacency-graph-plot, fig.height=8}
gha_with_id <- districts_raw[["GHA"]] %>%
  left_join(district_lookups[["GHA"]] %>% select(district_id, district_code),
            by = "district_code") %>%
  arrange(district_id)

centroid_coords <- suppressWarnings(st_coordinates(st_centroid(gha_with_id)))
centroids <- tibble(
  district_id = gha_with_id$district_id,
  x = centroid_coords[, "X"], y = centroid_coords[, "Y"]
)

sp <- spatial[["GHA"]]
edges_df <- tibble(from_id = sp$node1, to_id = sp$node2) %>%
  left_join(centroids %>% rename(x_from = x, y_from = y),
            by = c("from_id" = "district_id")) %>%
  left_join(centroids %>% rename(x_to = x, y_to = y),
            by = c("to_id" = "district_id"))

ggplot() +
  geom_sf(data = districts_raw[["GHA"]],
          fill = "grey95", colour = "grey60", linewidth = 0.2) +
  geom_segment(data = edges_df,
               aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
               colour = galway_maroon, alpha = 0.55, linewidth = 0.3) +
  geom_point(data = centroids, aes(x = x, y = y),
             colour = "black", size = 0.6) +
  labs(title = "District adjacency graph used for the ICAR prior, Ghana",
       subtitle = paste(sp$D, "districts,", sp$N_edges, "edges (queen contiguity)")) +
  theme_void(base_size = 11)
```

## Tyranny of the Average: Why District-Level Estimates Matter

```{r tyranny-of-average, fig.height=5, fig.width=12}
# Two side-by-side maps of Ghana 2022:
#   LEFT  - every district shaded by its REGION's average direct prevalence
#           (i.e. what you would see if you reported only at admin-1)
#   RIGHT - every district shaded by its own direct prevalence
#           (the district-level reality)
# Districts not directly sampled show NA in the right panel.

gha_panel_r2_ty <- panels[["GHA"]] %>%
  filter(round_id == 2) %>%
  left_join(district_lookups[["GHA"]] %>% select(district_id, district_code),
            by = "district_id") %>%
  mutate(prev_direct = ifelse(observed == 1, plogis(y_i), NA_real_))

# Bring NAME_1 (admin-1 / region) from the GADM shapefile onto the panel.
gha_with_region <- districts_raw[["GHA"]] %>%
  st_drop_geometry() %>%
  select(district_code, region = NAME_1) %>%
  right_join(gha_panel_r2_ty, by = "district_code")

# Region-level mean of the direct district estimates (unweighted; illustrative).
region_mean <- gha_with_region %>%
  group_by(region) %>%
  summarise(prev_region = mean(prev_direct, na.rm = TRUE), .groups = "drop")

gha_map_df <- districts_raw[["GHA"]] %>%
  left_join(gha_with_region %>% select(district_code, prev_direct),
            by = "district_code") %>%
  left_join(region_mean, by = c("NAME_1" = "region"))

p_region <- ggplot(gha_map_df) +
  geom_sf(aes(fill = prev_region), colour = "white", linewidth = 0.3) +
  prev_grad(name = "Prevalence") +
  labs(title = "Regional Average",
       subtitle = "Implies uniform, low-level risk") +
  theme_void(base_size = 12) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 11))

p_district <- ggplot(gha_map_df) +
  geom_sf(aes(fill = prev_direct), colour = "white", linewidth = 0.1) +
  prev_grad(name = "Prevalence") +
  labs(title = "District Reality",
       subtitle = "Reveals localised hotspots and cool spots") +
  theme_void(base_size = 12) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 11))

p_tyranny <- p_region | p_district

ggsave("output/figures/tyranny_of_average.png", p_tyranny,
       width = 12, height = 5, dpi = 300, bg = "white")
p_tyranny
```

The contrast between the two panels is the motivation for the rest of
this report. The regional average paints the country as uniformly
moderate-risk; the district reality reveals heterogeneity that policy
needs to see.

# Common Methodology {#sec:methodology}

All four models in this work share a common methodological core: a
Fay-Herriot area-level likelihood, a BYM2 spatial random effect, and
a shared set of priors. Each subsequent model section gives only the
specification elements unique to that model, then proceeds directly to
fitting, results and diagnostics.

## Fay-Herriot Likelihood

$$y_{it} \sim \mathcal{N}(\theta_{it}, \psi_{it}^2),$$
where $\theta_{it}$ is the latent logit prevalence and $\psi_{it}^2$
is treated as known (Section \ref{sec:direct}).

## BYM2 Spatial Random Effect

The district-level spatial random effect is given by [@riebler2016]:
$$b_i = \sigma_b \left(\sqrt{1-\phi}\, v_i + \sqrt{\phi/s}\, u_i\right),
\qquad v_i \sim \mathcal{N}(0,1), \quad u_i \sim \mathrm{ICAR}(W),$$
written in non-centred form to avoid funnel geometry in the posterior.
$W$ is the adjacency matrix of the relevant graph; $s$ is the BYM2
scaling factor; $\sigma_b$ is the marginal spatial SD; and
$\phi \in (0,1)$ is the proportion of structured variance.

## Shared Priors

| Parameter | Prior | Rationale |
|---|---|---|
| $\mu$ | $\mathcal{N}(-1.85, 1)$ | Centred near logit of moderate prevalence |
| $\sigma_b$ | $\mathrm{Exp}(4.6)$ | PC-style; $P(\sigma_b > 1) = 0.01$ [@simpson2017] |
| $\phi$ | $\mathrm{Beta}(3, 2)$ | Mild preference for structured > IID |
| $\beta_k$ | $\mathcal{N}(0, 1)$ | Weakly informative on standardised covariates |
| $\sigma_\gamma$ | $\mathrm{Exp}(2)$ | PC-style on interaction SD (Models 2--3) |
| $\tau_k$ | $\mathrm{Exp}(2)$ | PC-style on between-country SD (Models 3--4) |

## Computation

All models are fitted in Stan [@stan2024] via `cmdstanr`, using the
No-U-Turn Sampler with non-centred parameterisations throughout. Four
chains, 1500 warmup and 2000 sampling iterations each. Cross-sectional
and country-level space-time models use `adapt_delta = 0.99` and
`max_treedepth = 14`; pooled models use `adapt_delta = 0.97` and
`max_treedepth = 13`. Convergence is assessed via $\hat{R}$
[@vehtari2017], bulk/tail ESS, and trace/rank/pairs plots. Models are
compared by approximate leave-one-out cross-validation [@vehtari2017]
using the `loo` package.

## Stan Data Builders

```{r build-stan-data-cs}
build_stan_data_cs <- function(iso, panels, spatial, year, covar_cols = NULL) {
  panel <- panels[[iso]] %>% filter(year == !!year)
  sp <- spatial[[iso]]
  obs_idx <- which(panel$observed == 1)
  
  if (!is.null(covar_cols)) {
    available <- intersect(covar_cols, names(panel))
    X <- if (length(available) > 0) as.matrix(panel[, available]) else matrix(0, nrow(panel), 0)
    K <- ncol(X)
  } else {
    X <- matrix(0, nrow(panel), 0)
    K <- 0L
  }
  
  list(N = sp$D, N_obs = length(obs_idx),
       obs_idx = obs_idx,
       y = panel$y_i[obs_idx], psi = panel$psi_i[obs_idx],
       N_edges = sp$N_edges, node1 = sp$node1, node2 = sp$node2,
       scaling_factor = sp$scaling_factor,
       K = K, X = X)
}
```

```{r build-stan-data-st}
build_stan_data_st <- function(iso, panels, spatial, covar_cols = NULL) {
  panel <- panels[[iso]]
  sp <- spatial[[iso]]
  obs_idx <- which(panel$observed == 1)
  
  if (!is.null(covar_cols)) {
    available <- intersect(covar_cols, names(panel))
    X <- if (length(available) > 0) as.matrix(panel[, available]) else matrix(0, nrow(panel), 0)
    K <- ncol(X)
  } else {
    X <- matrix(0, nrow(panel), 0)
    K <- 0L
  }

  # Centred calendar year of the survey: yr - 2016. The panel carries a
  # `year` column for each row; this is the actual calendar year of the
  # DHS survey in which the (district, round) observation was made.
  year_c <- panel$year - 2016

  list(N = sp$D, T = 2L, NT = nrow(panel),
       N_obs = length(obs_idx), obs_idx = obs_idx,
       district = panel$district_id, time = panel$round_id,
       year_c   = year_c,
       y = panel$y_i[obs_idx], psi = panel$psi_i[obs_idx],
       N_edges = sp$N_edges, node1 = sp$node1, node2 = sp$node2,
       scaling_factor = sp$scaling_factor,
       K = K, X = X)
}
```

```{r assemble-stan-data}
covar_cols <- c("elev_std", "rainfall_std", "ntl_std", "temp_std",
                "travel_std", "pop_dens_std", "poverty_std")

stan_data <- list()

# Cross-sectional reference: Ghana 2022 only
stan_data[["GHA_CS_2022"]] <- build_stan_data_cs(
  "GHA", panels, spatial, year = 2022, covar_cols = covar_cols
)

# Space-time models for all three countries
stan_data[["GHA_ST"]] <- build_stan_data_st("GHA", panels, spatial, covar_cols = covar_cols)
stan_data[["BFA_ST"]] <- build_stan_data_st("BFA", panels, spatial, covar_cols = covar_cols)
stan_data[["CIV_ST"]] <- build_stan_data_st("CIV", panels, spatial, covar_cols = covar_cols)

map_dfr(stan_data, function(d) {
  tibble(
    N = d$N,
    NT = ifelse(is.null(d$NT), d$N, d$NT),
    N_obs = d$N_obs, K = d$K,
    N_edges = d$N_edges,
    scaling_factor = round(d$scaling_factor, 3)
  )
}, .id = "model")
```

# Model 1: Cross-Sectional Ghana Baseline {#sec:model1}

## Specification

For each Ghana district $i$ observed in Round 2,
$$\theta_i = \mu + b_i + \mathbf{x}_i^\top \bm{\beta},$$
with $b_i$ the BYM2 random effect on the Ghana-only adjacency graph.
For comparison, we fit two alternatives that turn off one spatial
component each: a pure IID model ($\phi = 0$) and a pure ICAR model
($\phi = 1$).

## Fitting: IID

```{r fit-iid, cache=TRUE}
iid_code <- '
data {
  int<lower=1> N;
  int<lower=1> N_obs;
  array[N_obs] int<lower=1, upper=N> obs_idx;
  vector[N_obs] y;
  vector<lower=0>[N_obs] psi;
  int<lower=1> K;
  matrix[N, K] X;
}
parameters {
  real mu;
  vector[N] v;
  real<lower=0> sigma_b;
  vector[K] beta;
}
transformed parameters {
  vector[N] theta;
  theta = mu + sigma_b * v + X * beta;
}
model {
  mu ~ normal(-1.85, 1.0);
  sigma_b ~ exponential(4.6);
  beta ~ normal(0, 1);
  v ~ std_normal();
  y ~ normal(theta[obs_idx], psi);
}
generated quantities {
  vector[N_obs] log_lik;
  vector[N] prevalence;
  for (i in 1:N_obs)
    log_lik[i] = normal_lpdf(y[i] | theta[obs_idx[i]], psi[i]);
  for (i in 1:N)
    prevalence[i] = inv_logit(theta[i]);
}
'
writeLines(iid_code, "fh_iid.stan")
mod_iid <- cmdstan_model("fh_iid.stan")

sd_cs <- stan_data[["GHA_CS_2022"]]

fit_iid <- mod_iid$sample(
  data = sd_cs, chains = 4, parallel_chains = 4,
  iter_warmup = 1000, iter_sampling = 1500,
  adapt_delta = 0.95, max_treedepth = 12, refresh = 500
)
fit_iid$diagnostic_summary()

fit_iid <- ensure_fit(fit_iid, "m1_iid")
```

## Model 1 (IID variant): parameters

```{r m1-iid-summary}
param_summary(fit_iid, "MODEL 1 - IID (Ghana 2022 cross-section)")
```

## Fitting: ICAR

```{r fit-icar, cache=TRUE}
icar_code <- '
data {
  int<lower=1> N;
  int<lower=1> N_obs;
  array[N_obs] int<lower=1, upper=N> obs_idx;
  vector[N_obs] y;
  vector<lower=0>[N_obs] psi;
  int<lower=1> N_edges;
  array[N_edges] int<lower=1, upper=N> node1;
  array[N_edges] int<lower=1, upper=N> node2;
  real<lower=0> scaling_factor;
  int<lower=1> K;
  matrix[N, K] X;
}
parameters {
  real mu;
  vector[N] u_raw;
  real<lower=0> sigma_b;
  vector[K] beta;
}
transformed parameters {
  vector[N] theta;
  vector[N] u = u_raw - mean(u_raw);
  theta = mu + sigma_b * u / sqrt(scaling_factor) + X * beta;
}
model {
  mu ~ normal(-1.85, 1.0);
  sigma_b ~ exponential(4.6);
  beta ~ normal(0, 1);
  target += -0.5 * dot_self(u_raw[node1] - u_raw[node2]);
  sum(u_raw) ~ normal(0, 0.001 * N);
  y ~ normal(theta[obs_idx], psi);
}
generated quantities {
  vector[N_obs] log_lik;
  vector[N] prevalence;
  for (i in 1:N_obs)
    log_lik[i] = normal_lpdf(y[i] | theta[obs_idx[i]], psi[i]);
  for (i in 1:N)
    prevalence[i] = inv_logit(theta[i]);
}
'
writeLines(icar_code, "fh_icar.stan")
mod_icar <- cmdstan_model("fh_icar.stan")

fit_icar <- mod_icar$sample(
  data = sd_cs, chains = 4, parallel_chains = 4,
  iter_warmup = 1000, iter_sampling = 1500,
  adapt_delta = 0.95, max_treedepth = 12, refresh = 500
)
fit_icar$diagnostic_summary()


fit_icar <- ensure_fit(fit_icar, "m1_icar")
```

## Model 1 (ICAR variant): parameters

```{r m1-icar-summary}
param_summary(fit_icar, "MODEL 1 - ICAR (Ghana 2022 cross-section)")
```

## Fitting: BYM2 (with log_lik for LOO)

```{r fit-cross-sectional}
mod_cs <- cmdstan_model("fh_bym2_cross_sectional.stan")

fit_gha_cs <- mod_cs$sample(
  data            = stan_data[["GHA_CS_2022"]],
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 1500,
  iter_sampling   = 2000,
  adapt_delta     = 0.99,
  max_treedepth   = 14,
  seed            = 123,
  refresh         = 500
)

fit_gha_cs$diagnostic_summary()


fit_gha_cs <- ensure_fit(fit_gha_cs, "m1_bym2_gha")

fit_gha_cs$print(c(
  "mu", "sigma_b", "phi", "p_national",
  "beta[1]", "beta[2]", "beta[3]", "beta[4]",
  "beta[5]", "beta[6]", "beta[7]"
))
```

## Model 1 (BYM2): parameters

```{r m1-bym2-keyparams}
param_summary(fit_gha_cs, "MODEL 1 - BYM2 (Ghana 2022 cross-section)")
```

## LOO Model Comparison

```{r loo-cv, fig.height=5, fig.width=9}
library(loo)


fit_iid     <- ensure_fit(fit_iid,    "m1_iid")
fit_icar    <- ensure_fit(fit_icar,   "m1_icar")
fit_gha_cs  <- ensure_fit(fit_gha_cs, "m1_bym2_gha")


safe_loo <- function(fit, name) {
  if (!("log_lik" %in% fit$metadata()$stan_variables)) {
    message(name, ": no log_lik variable; skipping LOO.")
    return(NULL)
  }
  tryCatch(
    fit$loo(variables = "log_lik"),
    error = function(e) {
      message(name, " LOO failed: ", conditionMessage(e))
      NULL
    }
  )
}

loo_iid  <- safe_loo(fit_iid,    "IID")
loo_icar <- safe_loo(fit_icar,   "ICAR")
loo_bym2 <- safe_loo(fit_gha_cs, "BYM2")


if (is.null(loo_bym2)) {
  message("Refitting BYM2 with log_lik for LOO comparison...")
  bym2_loo_code <- "
data {
  int<lower=1> N;
  int<lower=1> N_obs;
  array[N_obs] int<lower=1, upper=N> obs_idx;
  vector[N_obs] y;
  vector<lower=0>[N_obs] psi;
  int<lower=1> N_edges;
  array[N_edges] int<lower=1, upper=N> node1;
  array[N_edges] int<lower=1, upper=N> node2;
  real<lower=0> scaling_factor;
  int<lower=1> K;
  matrix[N, K] X;
}
parameters {
  real mu;
  vector[N] v;
  vector[N] u_raw;
  real<lower=0> sigma_b;
  real<lower=0, upper=1> phi;
  vector[K] beta;
}
transformed parameters {
  vector[N] theta;
  vector[N] u = u_raw - mean(u_raw);
  vector[N] b = sigma_b * (sqrt(1 - phi) * v + sqrt(phi / scaling_factor) * u);
  theta = mu + b + X * beta;
}
model {
  mu ~ normal(-1.85, 1.0);
  sigma_b ~ exponential(4.6);
  phi ~ beta(3, 2);
  beta ~ normal(0, 1);
  v ~ std_normal();
  target += -0.5 * dot_self(u_raw[node1] - u_raw[node2]);
  sum(u_raw) ~ normal(0, 0.001 * N);
  y ~ normal(theta[obs_idx], psi);
}
generated quantities {
  vector[N_obs] log_lik;
  vector[N] prevalence;
  for (i in 1:N_obs)
    log_lik[i] = normal_lpdf(y[i] | theta[obs_idx[i]], psi[i]);
  for (i in 1:N)
    prevalence[i] = inv_logit(theta[i]);
}
"
  writeLines(bym2_loo_code, "fh_bym2_loo.stan")
  mod_bym2_loo <- cmdstan_model("fh_bym2_loo.stan")
  fit_bym2_loo <- mod_bym2_loo$sample(
    data = sd_cs, chains = 4, parallel_chains = 4,
    iter_warmup = 1000, iter_sampling = 1500,
    adapt_delta = 0.95, max_treedepth = 12, refresh = 500
  )
  fit_bym2_loo <- ensure_fit(fit_bym2_loo, "m1_bym2_loo")
  loo_bym2 <- safe_loo(fit_bym2_loo, "BYM2 (refit)")
}

loo_list <- list(IID = loo_iid, ICAR = loo_icar, BYM2 = loo_bym2)
loo_list <- Filter(Negate(is.null), loo_list)

if (length(loo_list) >= 2L) {
  loo_comp <- loo_compare(loo_list)
  print(loo_comp, simplify = FALSE)

  loo_df <- as_tibble(loo_comp, rownames = "Model")

  p_loo <- loo_df %>%
    mutate(Model = factor(Model, levels = rev(Model))) %>%
    ggplot(aes(x = Model, y = elpd_diff)) +
    geom_col(aes(fill = Model), width = 0.6, alpha = 0.8) +
    geom_errorbar(aes(ymin = elpd_diff - 2 * se_diff,
                       ymax = elpd_diff + 2 * se_diff), width = 0.2) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
    scale_fill_manual(values = c("BYM2" = galway_maroon, "ICAR" = galway_ocean, "IID" = galway_orange)) +
    coord_flip() +
    labs(title    = "LOO-CV: IID vs ICAR vs BYM2 (Ghana 2022)",
         subtitle = "Delta ELPD relative to best model (higher = better)",
         x = NULL, y = "Delta ELPD") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none")

  ggsave("output/figures/loo_cv_comparison.png", p_loo,
         width = 9, height = 5, dpi = 300, bg = "white")
  p_loo
} else {
  message("Fewer than two LOO objects available; skipping comparison plot.")
}
```

## Maps: Direct vs IID vs ICAR vs BYM2

```{r model-comparison-maps, fig.height=6, fig.width=14}
# Extract prevalence from each model
get_prev <- function(fit) {
  s <- fit$summary("prevalence")
  s$median
}

prev_iid_v  <- get_prev(fit_iid)
prev_icar_v <- get_prev(fit_icar)


if ("prevalence" %in% fit_gha_cs$metadata()$stan_variables) {
  prev_bym2_v <- get_prev(fit_gha_cs)
} else {
  prev_bym2_v <- get_prev(fit_bym2_loo)
}

D_gha_cs <- stan_data[["GHA_CS_2022"]]$N

model_prev <- bind_rows(
  tibble(district_id = 1:D_gha_cs, prev = prev_iid_v[1:D_gha_cs], model = "IID"),
  tibble(district_id = 1:D_gha_cs, prev = prev_icar_v[1:D_gha_cs], model = "ICAR"),
  tibble(district_id = 1:D_gha_cs, prev = prev_bym2_v[1:D_gha_cs], model = "BYM2"),
  panels[["GHA"]] %>% filter(round_id == 2) %>%
    mutate(prev = ifelse(observed == 1, plogis(y_i), NA_real_), model = "Direct") %>%
    select(district_id, prev, model)
) %>%
  mutate(model = factor(model, levels = c("Direct", "IID", "ICAR", "BYM2")))

# Join to shapefile via district_id
gha_shp <- districts_raw[["GHA"]] %>%
  mutate(district_id = row_number())

maps_comp <- map_dfr(levels(model_prev$model), function(m) {
  gha_shp %>%
    left_join(model_prev %>% filter(model == m), by = "district_id") %>%
    mutate(model = m)
}) %>%
  mutate(model = factor(model, levels = c("Direct", "IID", "ICAR", "BYM2")))

p_comp <- ggplot(maps_comp) +
  geom_sf(aes(fill = prev), colour = "white", linewidth = 0.1) +
  prev_grad(name = "Prevalence") +
  facet_wrap(~model, nrow = 1) +
  labs(title = "Model comparison: Direct vs IID vs ICAR vs BYM2 (Ghana 2022)",
       subtitle = "Grey = unobserved in direct estimates. Spatial models fill all 260 districts.") +
  theme_void(base_size = 11) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 12, face = "bold"))

ggsave("output/figures/model_comparison_maps_ghana.png", p_comp, width = 14, height = 6, dpi = 300, bg = "white")
p_comp
```

## Posterior Predictive Checks

```{r ppc-cs}
y_obs_cs <- stan_data[["GHA_CS_2022"]]$y
y_rep_cs <- fit_gha_cs$draws("y_rep", format = "matrix")

ppc_dens_overlay(y = y_obs_cs, yrep = y_rep_cs[1:100, ]) +
  labs(title = "PPC: Density overlay (Ghana 2022 cross-sectional)")

ppc_stat(y = y_obs_cs, yrep = y_rep_cs, stat = "mean") +
  labs(title = "PPC: Mean of y")

ppc_stat(y = y_obs_cs, yrep = y_rep_cs, stat = "sd") +
  labs(title = "PPC: Standard deviation of y")
```

## Convergence Diagnostics

```{r cs-trace}
mcmc_trace(fit_gha_cs$draws(c("mu", "sigma_b", "phi")))
```

# Model 2: Country-Level Space-Time Models {#sec:model2}

## Specification

For each country $c$ fitted separately,
$$\theta_{it} = \mu^{(c)} + b_i^{(c)} + \delta^{(c)} \, \widetilde{\text{yr}}_{it}
              + \gamma_{it}^{(c)} + \mathbf{x}_{it}^\top \bm{\beta}^{(c)},$$
with $b_i^{(c)}$ the BYM2 effect on the country-specific graph,
$\gamma_{it}^{(c)} \sim \mathcal{N}(0, \sigma_\gamma^2)$ the Knorr-Held
Type I [@knorrheld2000] interaction, and $\widetilde{\text{yr}}_{it} =
\text{yr}_{it} - 2016$ the centred calendar year of the survey. The
temporal effect is expressed as a single country-specific linear slope
$\delta^{(c)}$ on calendar year rather than as a wave indicator
$\delta_t^{(c)}$. This matches the parameterisation used by Model 4
(Section \ref{sec:model4}) and is preferred for two reasons. First,
the three DHS Round 1 surveys fall in different calendar years (Burkina
Faso 2010, Côte d'Ivoire 2012, Ghana 2014), so a wave indicator would
conflate each country's elapsed time between rounds with its rate of
change [@wakefield2020]. Second, the calendar-year form generalises
cleanly to more than two rounds. With $T = 2$ the two parameterisations
give identical between-round odds ratios; the round-to-round shift on
the logit scale is recovered as
$\delta^{(c)} \cdot (\widetilde{\text{yr}}_{R2}^{(c)} - \widetilde{\text{yr}}_{R1}^{(c)})
= \delta^{(c)} \cdot \Delta\text{yr}^{(c)}$
and the between-round odds ratio as $\exp(\delta^{(c)} \cdot
\Delta\text{yr}^{(c)})$, where $\Delta\text{yr}^{(c)}$ is the actual
elapsed years between the country's two surveys (8 years for Ghana, 9
for Côte d'Ivoire, 11 for Burkina Faso). With only two rounds the Type
I interaction is the only identifiable choice: structured temporal or
space-time interactions are not separately identified [@goicoa2018].

## Stan model

```{r m2-stan-code}
m2_stan_code <- '
data {
  int<lower=1> N;
  int<lower=1> T;
  int<lower=1> NT;
  int<lower=1> K;

  int<lower=1> N_obs;
  array[N_obs] int<lower=1, upper=NT> obs_idx;
  vector[N_obs] y;
  vector<lower=0>[N_obs] psi;

  array[NT] int<lower=1, upper=N> district;
  array[NT] int<lower=1, upper=T> time;
  vector[NT] year_c;

  matrix[NT, K] X;

  int<lower=1> N_edges;
  array[N_edges] int<lower=1, upper=N> node1;
  array[N_edges] int<lower=1, upper=N> node2;
  real<lower=0> scaling_factor;
}

parameters {
  real mu;
  real delta;

  vector[N] v;
  vector[N] u_raw;
  real<lower=0> sigma_b;
  real<lower=0, upper=1> phi;

  vector[K] beta;

  vector[NT] gamma_raw;
  real<lower=0> sigma_gamma;
}

transformed parameters {
  vector[N] u = u_raw - mean(u_raw);
  vector[N] b = sigma_b * (sqrt(1 - phi) * v
                           + sqrt(phi / scaling_factor) * u);

  vector[NT] gamma_star = sigma_gamma * gamma_raw;

  vector[NT] theta;
  for (n in 1:NT) {
    int d = district[n];
    real xb = 0;
    for (k in 1:K)
      xb += X[n, k] * beta[k];
    theta[n] = mu + b[d] + delta * year_c[n] + gamma_star[n] + xb;
  }
}

model {
  mu ~ normal(-1.85, 1.5);
  delta ~ normal(0, 0.5);

  sigma_b ~ exponential(4.6);
  phi ~ beta(3, 2);
  v ~ std_normal();
  target += -0.5 * dot_self(u_raw[node1] - u_raw[node2]);
  sum(u_raw) ~ normal(0, 0.001 * N);

  beta ~ normal(0, 1);

  sigma_gamma ~ exponential(2);
  gamma_raw ~ std_normal();

  y ~ normal(theta[obs_idx], psi);
}

generated quantities {
  vector[N_obs] log_lik;
  for (i in 1:N_obs)
    log_lik[i] = normal_lpdf(y[i] | theta[obs_idx[i]], psi[i]);

  vector[NT] prevalence;
  for (n in 1:NT)
    prevalence[n] = inv_logit(theta[n]);

  // National prevalence at each round (population-weighted mean would be
  // ideal; here we use the unweighted district mean for consistency with
  // Model 2 of the workbook).
  vector[T] p_national;
  for (t in 1:T) {
    real total = 0;
    int  count = 0;
    for (n in 1:NT) {
      if (time[n] == t) { total += prevalence[n]; count += 1; }
    }
    p_national[t] = total / count;
  }

  // Round 1 and Round 2 calendar years (centred) and the implied
  // round-to-round shift on the logit scale.
  real yr1_c = -999;
  real yr2_c = -999;
  for (n in 1:NT) {
    if (time[n] == 1 && yr1_c == -999) yr1_c = year_c[n];
    if (time[n] == 2 && yr2_c == -999) yr2_c = year_c[n];
  }
  real delta_R2_R1     = delta * (yr2_c - yr1_c);
  real odds_ratio_R2_R1 = exp(delta_R2_R1);
}
'

writeLines(m2_stan_code, "fh_bym2_calyr.stan")
```

## Fitting: All Three Countries

```{r fit-spacetime-all}
mod_st <- cmdstan_model("fh_bym2_calyr.stan")

fit_st <- list()

for (iso in c("GHA", "BFA", "CIV")) {
  cat("\n========== Fitting space-time model for", iso, "==========\n")
  
  fit_st[[iso]] <- mod_st$sample(
    data            = stan_data[[paste0(iso, "_ST")]],
    chains          = 4,
    parallel_chains = 4,
    iter_warmup     = 1500,
    iter_sampling   = 2000,
    adapt_delta     = 0.99,
    max_treedepth   = 14,
    seed            = 1729,
    refresh         = 500
  )
  
  cat("\nDiagnostics for", iso, ":\n")
  print(fit_st[[iso]]$diagnostic_summary())
  # Persist + restore draws so the fit survives across knit sessions.
  # New cache name (`m2_calyr_<iso>`) keeps the wave-indicator fits, if any,
  # at their old paths `m2_st_<iso>`.
  fit_st[[iso]] <- ensure_fit(fit_st[[iso]], paste0("m2_calyr_", iso))
}
```

## Model 2: Country-level space-time parameters

```{r m2-summaries}


safe_print <- function(fit, params, label = NULL) {
  if (!is.null(label)) cat("
--- ", label, " ---
", sep = "")
  available <- fit$metadata()$stan_variables
  base_names <- sub("\\[.*$", "", params)
  keep <- params[base_names %in% available]
  if (length(keep) == 0L) {
    cat("  [none of the requested parameters exist in this fit]
")
    return(invisible(NULL))
  }
  tryCatch(fit$print(keep),
           error = function(e) cat("  [print failed:", conditionMessage(e), "]
"))
}

# Model-level parameters for each country (district-level effects excluded)
for (cc in names(fit_st)) {
  param_summary(fit_st[[cc]], paste0("MODEL 2 - ", cc, " (country-level space-time)"))
}
```

```{r m2-key-params}
# Key parameters across the three country-level space-time models.
# `delta`            annual rate of change on logit scale (per calendar year)
# `delta_R2_R1`      implied Round-2 minus Round-1 shift on logit scale
# `odds_ratio_R2_R1` implied between-round odds ratio = exp(delta_R2_R1)
# `p_national[t]`    modelled national prevalence at round t
key_st_params <- c(
  "mu", "sigma_b", "phi", "sigma_gamma",
  "delta", "delta_R2_R1", "odds_ratio_R2_R1",
  "p_national[1]", "p_national[2]",
  paste0("beta[", 1:7, "]")
)

for (cc in names(fit_st)) {
  safe_print(fit_st[[cc]], key_st_params, label = cc)
}
```

## Summary Across Countries

```{r summary-across-countries}
summary_all <- map_dfr(c("GHA", "BFA", "CIV"), function(iso) {
  draws <- fit_st[[iso]]$draws(
    variables = c("mu", "delta", "delta_R2_R1", "odds_ratio_R2_R1",
                  "sigma_b", "phi", "sigma_gamma",
                  "p_national[1]", "p_national[2]",
                  "beta[1]", "beta[2]", "beta[3]", "beta[4]",
                  "beta[5]", "beta[6]", "beta[7]"),
    format = "draws_matrix"
  )
  
  tibble(
    country  = iso,
    variable = colnames(draws),
    mean     = colMeans(draws),
    median   = apply(draws, 2, median),
    lower    = apply(draws, 2, quantile, 0.025),
    upper    = apply(draws, 2, quantile, 0.975),
    rhat     = apply(draws, 2, posterior::rhat),
    ess_bulk = apply(draws, 2, posterior::ess_bulk)
  )
})

print(n = 25, summary_all)

# fit_st$print(c(
#   "mu", "sigma_b", "phi", "p_national",
#   "beta[1]", "beta[2]", "beta[3]", "beta[4]",
#   "beta[5]", "beta[6]", "beta[7]"
# ))
```

## Prior vs Posterior Comparison

```{r prior-post}
post_draws <- fit_st[["GHA"]]$draws(
  c("mu", "delta", "sigma_b", "phi", "sigma_gamma"),
  format = "draws_df"
)

prior_post_plot <- function(post_var, prior_fun, x_range, label) {
  ggplot() +
    geom_function(fun = prior_fun, xlim = x_range,
                  aes(colour = "Prior"), linewidth = 0.9) +
    geom_density(data = post_draws,
                 aes(x = .data[[post_var]], colour = "Posterior"),
                 linewidth = 0.9) +
    scale_colour_manual(values = c(Prior = galway_ash, Posterior = galway_maroon)) +
    labs(x = label, y = "Density", colour = NULL) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "none")
}

p_mu     <- prior_post_plot("mu",          ~ dnorm(.x, -1.85, 1), c(-4, 0),  expression(mu))
p_delta  <- prior_post_plot("delta",       ~ dnorm(.x, 0, 0.5),   c(-0.5, 0.5), expression(delta))
p_sigb   <- prior_post_plot("sigma_b",     ~ dexp(.x, 4.6),       c(0, 2),   expression(sigma[b]))
p_phi    <- prior_post_plot("phi",         ~ dbeta(.x, 3, 2),     c(0, 1),   expression(phi))
p_siggam <- prior_post_plot("sigma_gamma", ~ dexp(.x, 2),         c(0, 2),   expression(sigma[gamma]))

(p_mu | p_delta | p_sigb) / (p_phi | p_siggam | plot_spacer()) +
  plot_annotation(title = "Prior vs posterior, space-time BYM2 model (Ghana)")
```

## Covariate Effects Across Countries (Country-Specific Models)

```{r covariate-forest}
covar_labels <- c("Elevation", "Rainfall", "Nighttime lights", "Temperature",
                  "Travel time", "Pop. density (log)", "Poverty")


beta_all <- summary_all %>%
  # Filter only the rows where the variable starts with "beta["
  filter(str_detect(variable, "^beta\\[")) %>%
  mutate(
    # Extract the numeric index from "beta[1]", "beta[2]", etc.
    beta_idx = as.integer(str_extract(variable, "\\d+")),
    
    # Map the index to the correct human-readable label
    covariate = covar_labels[beta_idx],
    
    # Calculate significance strictly using the lower/upper bounds in the table
    # If both bounds are greater than 0, or both are less than 0, it is significant
    is_significant = (lower > 0) | (upper < 0),
    sig_marker = ifelse(is_significant, "*", "")
  )

# 3. Factor the covariates and countries to preserve plotting order
beta_all$covariate <- factor(beta_all$covariate, levels = rev(covar_labels))
beta_all$country <- factor(beta_all$country, levels = c("GHA", "BFA", "CIV"))

# 4. Generate the plot
p_covars <- ggplot(beta_all, aes(x = mean, y = covariate, colour = country, group = country)) +
  
  geom_pointrange(aes(xmin = lower, xmax = upper),
                  position = position_dodge(width = 0.6), size = 0.8) +
                  
  # FIXED ALIGNMENT: Added hjust = 0 and locked the dodge with group = country
  geom_text(aes(x = upper + 0.02, label = sig_marker),
            position = position_dodge(width = 0.6),
            size = 7,          # Slightly larger for visibility on a slide
            vjust = 0.75,      # Pulls the top-heavy asterisk down to center it with the line
            hjust = 0,         # Left-aligns it exactly 0.02 units away from the line
            show.legend = FALSE, 
            colour = "black") +
            
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40", linewidth = 0.8) +
  scale_colour_manual(values = galway_three, name = "Country") +
  
  scale_x_continuous(expand = expansion(mult = c(0.05, 0.15))) + 
  
  labs(x = "Coefficient (logit scale)", y = NULL,
       title = "Covariate effects: Individual Space-Time Models",
       subtitle = "Posterior mean and credible intervals. Asterisks (*) denote significant effects.") +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

ggsave("output/figures/individual_covariate_effects.png", p_covars, width = 10, height = 7, dpi = 300, bg = "white")
p_covars
```

## Prevalence Maps: All Countries, Both Rounds

```{r m2-prevalence-maps, fig.height=6, fig.width=12}
# Pull posterior median prevalence per district per round from each country
# fit, then attach to the joined three-country shapefile `all_sf` via the
# country-local district id. The result is one map per round across all 724
# districts, with the three countries appearing as a single contiguous
# West African block.

get_prev <- function(fit) {
  s <- fit$summary("prevalence")
  s$median
}

m2_prev_long <- map_dfr(c("GHA", "BFA", "CIV"), function(iso) {
  prev_v <- get_prev(fit_st[[iso]])
  tibble(
    country  = iso,
    row_id   = seq_along(prev_v),
    prev     = prev_v
  ) %>%
    left_join(
      panels[[iso]] %>% mutate(row_id = row_number()) %>%
        select(row_id, district_id, round_id),
      by = "row_id"
    ) %>%
    select(country, district_id, round_id, prev)
})

m2_maps_df <- all_sf %>%
  left_join(m2_prev_long,
            by = c("country" = "country", "local_id" = "district_id"))

p_m2 <- ggplot(m2_maps_df) +
  geom_sf(aes(fill = prev), colour = "white", linewidth = 0.1) +
  prev_grad(name = "Prevalence") +
  facet_wrap(~ paste("Round", round_id), nrow = 1) +
  labs(title = "Model 2 (country-level space-time): posterior median prevalence",
       subtitle = "Three countries fitted independently and joined on the combined shapefile.") +
  theme_void(base_size = 11) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 12, face = "bold"))

ggsave("output/figures/m2_prevalence_maps_3country.png", p_m2,
       width = 12, height = 6, dpi = 300, bg = "white")
p_m2
```

## District Trend Deviations

```{r trend-deviation-map}
dev_summary_gha <- fit_st[["GHA"]]$summary("gamma_star",
  median = ~ median(.x),
  lower  = ~ unname(quantile(.x, 0.025)),
  upper  = ~ unname(quantile(.x, 0.975))
) %>%
  mutate(
    district_id = as.integer(gsub("gamma_star\\[|\\]", "", variable)),
    confident = (lower > 0) | (upper < 0)
  ) %>%
  left_join(district_lookups[["GHA"]] %>%
              select(district_id, district_code, district_name),
            by = "district_id")

map_dev <- districts_raw[["GHA"]] %>%
  left_join(dev_summary_gha, by = "district_code")

dev_limit <- max(abs(range(map_dev$median, na.rm = TRUE)))

ggplot(map_dev) +
  geom_sf(aes(fill = median), colour = "grey40", linewidth = 0.1) +
  scale_fill_gradient2(
    name = expression(gamma[i]^"*"),
    low = galway_ocean, mid = "white", high = galway_maroon,
    midpoint = 0, limits = c(-dev_limit, dev_limit)
  ) +
  geom_sf(data = map_dev %>% filter(confident),
          fill = NA, colour = "black", linewidth = 0.4) +
  labs(
    title = "District-level deviation from national malaria trend, Ghana 2014–2022",
    subtitle = "Blue: faster decline than national. Red: slower / rising. Bold outline: 95% CI excludes 0."
  ) +
  theme_void(base_size = 11)
```

## Top Trend-Deviating Districts

```{r top-deviation-districts}
top_neg <- dev_summary_gha %>% arrange(median) %>% head(10) %>%
  mutate(direction = "Faster decline")
top_pos <- dev_summary_gha %>% arrange(desc(median)) %>% head(10) %>%
  mutate(direction = "Slower / rising")

bind_rows(top_neg, top_pos) %>%
  mutate(district_name = fct_reorder(district_name, median)) %>%
  ggplot(aes(x = median, y = district_name, colour = direction)) +
  geom_pointrange(aes(xmin = lower, xmax = upper)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_colour_manual(values = c("Faster decline" = galway_ocean,
                                 "Slower / rising" = galway_maroon)) +
  labs(x = expression(gamma[i]^"*" * " (logit scale)"), y = NULL,
       title = "Top districts with credible deviation from national trend, Ghana",
       colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

## Uncertainty Maps

```{r uncertainty-maps, fig.height=5}
make_uncertainty_df <- function(iso, round) {
  unc <- fit_st[[iso]]$summary("prevalence",
    width = ~ unname(quantile(.x, 0.975) - quantile(.x, 0.025))
  ) %>%
    mutate(row_id = as.integer(gsub("prevalence\\[|\\]", "", variable))) %>%
    left_join(
      panels[[iso]] %>% mutate(row_id = row_number()) %>%
        select(row_id, district_id, round_id, observed),
      by = "row_id"
    ) %>%
    filter(round_id == round) %>%
    left_join(district_lookups[[iso]] %>% select(district_id, district_code),
              by = "district_id")
  
  districts_raw[[iso]] %>%
    left_join(unc, by = "district_code") %>%
    mutate(country = iso)
}

unc_r2 <- map_dfr(c("GHA", "BFA", "CIV"),
                  function(iso) make_uncertainty_df(iso, 2))

# Switch to the joined 3-country shapefile.
unc_long <- map_dfr(c("GHA", "BFA", "CIV"), function(iso) {
  s <- fit_st[[iso]]$summary("prevalence",
                              q5  = ~ unname(quantile(.x, 0.025)),
                              q95 = ~ unname(quantile(.x, 0.975)))
  tibble(
    country = iso,
    row_id  = seq_len(nrow(s)),
    width   = s$q95 - s$q5
  ) %>%
    left_join(
      panels[[iso]] %>% mutate(row_id = row_number()) %>%
        select(row_id, district_id, round_id, observed),
      by = "row_id"
    ) %>%
    filter(round_id == 2L) %>%
    select(country, district_id, width, observed)
})

unc_map_df <- all_sf %>%
  left_join(unc_long,
            by = c("country" = "country", "local_id" = "district_id"))

p_unc <- ggplot(unc_map_df) +
  geom_sf(aes(fill = width), colour = "white", linewidth = 0.1) +
  prev_grad(name = "95% CI width",
            labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Posterior uncertainty (Round 2): 95% credible interval widths",
       subtitle = "Wider intervals indicate districts that rely more on spatial smoothing.") +
  theme_void(base_size = 11) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 12, face = "bold"))

ggsave("output/figures/m2_uncertainty_r2_3country.png", p_unc,
       width = 11, height = 6, dpi = 300, bg = "white")
p_unc
```

## Observed vs Unobserved Districts

```{r observed-unobserved-map, fig.height=6, fig.width=12}
# Joined-shapefile version: all three countries on a single map, faceted by
# round. Districts containing at least one DHS cluster in that round are
# shaded; the rest are flagged unobserved.
obs_long <- map_dfr(c("GHA", "BFA", "CIV"), function(iso) {
  panels[[iso]] %>%
    select(district_id, round_id, observed) %>%
    mutate(country = iso,
           status  = ifelse(observed == 1, "Observed", "Unobserved"))
})

obs_map_df <- all_sf %>%
  left_join(obs_long,
            by = c("country" = "country", "local_id" = "district_id"))

p_obs <- ggplot(obs_map_df) +
  geom_sf(aes(fill = status), colour = "white", linewidth = 0.1) +
  scale_fill_manual(
    values = c("Observed" = galway_ocean, "Unobserved" = galway_sandstone),
    name = "Survey data"
  ) +
  facet_wrap(~ paste("Round", round_id), nrow = 1) +
  labs(
    title = "Sampled and unsampled districts by survey round",
    subtitle = "Unobserved districts contain no DHS clusters in that round and rely on the model for inference."
  ) +
  theme_void(base_size = 11) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 12, face = "bold"))

ggsave("output/figures/observed_vs_unobserved_3country.png", p_obs,
       width = 12, height = 6, dpi = 300, bg = "white")
p_obs
```

## Credible Interval Comparison: Observed vs Unobserved

```{r ci-comparison-obs-unobs, fig.height=6, fig.width=10}
make_ci_comparison <- function(iso, round) {
  D <- stan_data[[paste0(iso, "_ST")]]$N
  
  # Use default summary (returns q5, q95, mean, median, sd, etc.)
  prev_summary <- fit_st[[iso]]$summary("prevalence") %>%
    mutate(row_id = as.integer(gsub("prevalence\\[|\\]", "", variable))) %>%
    left_join(
      panels[[iso]] %>% mutate(row_id = row_number()) %>%
        select(row_id, district_id, round_id, observed),
      by = "row_id"
    ) %>%
    filter(round_id == round) %>%
    mutate(
      ci_width = q95 - q5,
      status = ifelse(observed == 1, "Observed", "Unobserved"),
      country = iso
    ) %>%
    left_join(district_lookups[[iso]] %>% select(district_id, district_name),
              by = "district_id")
  
  prev_summary
}

ci_all <- map_dfr(c("GHA", "BFA", "CIV"),
                  function(iso) make_ci_comparison(iso, 2))

# Boxplot comparing CI widths
p_ci_box <- ggplot(ci_all, aes(x = status, y = ci_width, fill = status)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8) +
  facet_wrap(~country, scales = "free_y") +
  scale_fill_manual(values = c("Observed" = galway_ocean, "Unobserved" = galway_orange)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Width of 90% credible intervals: observed vs unobserved districts",
    subtitle = "Round 2 estimates. Unobserved districts have wider intervals (more uncertainty)",
    x = NULL, y = "Width of 90% CrI",
    fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

ggsave("output/figures/ci_comparison_obs_unobs.png", p_ci_box, width = 10, height = 6, dpi = 300, bg = "white")
p_ci_box
```

## Caterpillar Plot: Sorted Prevalence with Credible Intervals

```{r caterpillar-plot, fig.height=7, fig.width=10}
ci_gha <- ci_all %>%
  filter(country == "GHA") %>%
  arrange(median) %>%
  mutate(rank = row_number())

p_cat <- ggplot(ci_gha, aes(x = rank, y = median, colour = status)) +
  geom_linerange(aes(ymin = q5, ymax = q95), alpha = 0.5, linewidth = 0.3) +
  geom_point(size = 0.6) +
  scale_colour_manual(values = c("Observed" = galway_ocean, "Unobserved" = galway_orange)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Caterpillar plot: district prevalence estimates, Ghana Round 2",
    subtitle = "260 districts sorted by posterior median. Gold = unobserved (wider CrI).",
    x = "District (sorted)", y = "Prevalence (posterior median and 90% CrI)",
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

ggsave("output/figures/caterpillar_plot_ghana_r2.png", p_cat, width = 10, height = 7, dpi = 300, bg = "white")
p_cat
```

## Direct vs Model-Smoothed Estimates

```{r direct-vs-smoothed, fig.height=7, fig.width=9}
gha_panel_r2 <- panels[["GHA"]] %>% filter(round_id == 2)
gha_direct <- gha_panel_r2 %>%
  filter(observed == 1) %>%
  mutate(direct_prev = plogis(y_i))

gha_model <- fit_st[["GHA"]]$summary("prevalence") %>%
  mutate(row_id = as.integer(gsub("prevalence\\[|\\]", "", variable))) %>%
  left_join(
    panels[["GHA"]] %>% mutate(row_id = row_number()) %>%
      select(row_id, district_id, round_id),
    by = "row_id"
  ) %>%
  filter(round_id == 2)

shrinkage_df <- gha_direct %>%
  select(district_id, direct_prev) %>%
  left_join(gha_model %>% select(district_id, model_prev = median,
                                   model_lower = q5, model_upper = q95),
            by = "district_id") %>%
  left_join(district_lookups[["GHA"]] %>% select(district_id, district_name),
            by = "district_id")

p_shrink <- ggplot(shrinkage_df, aes(x = direct_prev, y = model_prev)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_linerange(aes(ymin = model_lower, ymax = model_upper),
                 alpha = 0.15, colour = galway_maroon) +
  geom_point(alpha = 0.6, colour = galway_maroon, size = 1.5) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "BYM2 shrinkage: direct estimates vs model-smoothed posteriors",
    subtitle = "Ghana Round 2. Points below diagonal = model pulled estimate down. Bars = 90% CrI.",
    x = "Direct estimate (design-based)",
    y = "Model posterior median"
  ) +
  theme_minimal(base_size = 12)

ggsave("output/figures/direct_vs_smoothed_ghana_r2.png", p_shrink, width = 9, height = 7, dpi = 300, bg = "white")
p_shrink
```

## Space-Time Convergence Diagnostics

```{r st-diagnostics}

mcmc_rank_overlay(fit_st[["GHA"]]$draws(
  c("mu", "delta", "sigma_b", "phi", "sigma_gamma")
))

mcmc_pairs(
  fit_st[["GHA"]]$draws(c("sigma_b", "phi", "sigma_gamma", "delta")),
  off_diag_args = list(size = 0.5, alpha = 0.3)
)

```

# Three-Country Pooled Infrastructure {#sec:pooled-infra}



## Motivation

The country-specific models in Section \ref{sec:model2} treat each
country as a spatial island, despite shared borders, shared *Anopheles*
vectors and shared West African monsoon dynamics. The pooled model
relaxes this by smoothing across the combined 724-district adjacency
graph (including cross-border edges following @saldarriaga2020) and by
partially pooling country-specific covariate slopes through a
hierarchical prior $\beta_{ck} \sim \mathcal{N}(\bar{\beta}_k, \tau_k^2)$.

## Construction

```{r}
gha_sf <- districts_raw[["GHA"]] %>%
  mutate(country = "GHA", country_id = 1L,
         local_id = row_number())

civ_sf <- districts_raw[["CIV"]] %>%
  mutate(country = "CIV", country_id = 2L,
         local_id = row_number())

bfa_sf <- districts_raw[["BFA"]] %>%
  mutate(country = "BFA", country_id = 3L,
         local_id = row_number())

D_gha <- nrow(gha_sf)  # 260
D_civ <- nrow(civ_sf)  # 113
D_bfa <- nrow(bfa_sf)  # 351
D_all <- D_gha + D_civ + D_bfa  # 724


#assigning glboal IDs
gha_sf$global_id <- 1:D_gha
civ_sf$global_id <- (D_gha + 1):(D_gha + D_civ)
bfa_sf$global_id <- (D_gha + D_civ + 1):D_all

# binding them all together
all_sf <- bind_rows(gha_sf, civ_sf, bfa_sf)

# constracting the queens contiguity on the combined spatial shapefile
nb_all <- poly2nb(all_sf)

# counting cross border edges 
cross_border <- 0
within_country <- 0
for (i in seq_along(nb_all)) {
  for (j in nb_all[[i]]) {
    if (j > i) {  # avoid double counting
      if (all_sf$country[i] != all_sf$country[j]) {
        cross_border <- cross_border + 1
      } else {
        within_country <- within_country + 1
      }
    }
  }
}

cross_border

# Extracting edge list for Stan
edges <- nb_to_edgelist <- function(nb) {
  n1 <- c(); n2 <- c()
  for (i in seq_along(nb)) {
    for (j in nb[[i]]) {
      if (j > i) {
        n1 <- c(n1, i)
        n2 <- c(n2, j)
      }
    }
  }
  list(node1 = n1, node2 = n2, N_edges = length(n1))
}

el <- nb_to_edgelist(nb_all)
cat("    Edges extracted:", el$N_edges, "\n")

# Computing the BYM2 scaling factor based on the combined spatial dataset
Q <- -nb2mat(nb_all, style = "B", zero.policy = TRUE)
diag(Q) <- -rowSums(Q)
Q_scaled <- INLA::inla.scale.model(Q, constr = list(A = matrix(1, 1, D_all), e = 0))
scaling_factor_all <- exp(mean(log(diag(Q_scaled))))


# Country index for each district (1=GHA, 2=CIV, 3=BFA)
country_vec <- all_sf$country_id  # length 724

# Re-index panels with global district IDs
# Each country's panel has: district_id (local 1:D), round_id (1 or 2),
# y_i, psi_i, observed, and covariates

remap_panel <- function(iso, offset) {
  panels[[iso]] %>%
    mutate(
      global_district_id = district_id + offset,
      country = iso
    )
}

panel_pooled <- bind_rows(
  remap_panel("GHA", 0),
  remap_panel("CIV", D_gha),
  remap_panel("BFA", D_gha + D_civ)
) %>%
  arrange(global_district_id, round_id)

NT <- nrow(panel_pooled)
cat("  Panel rows (NT):", NT, "(should be", D_all * 2, ")\n")

# Observation indices
obs_mask <- panel_pooled$observed == 1
obs_idx  <- which(obs_mask)
N_obs    <- sum(obs_mask)

cat("  Observed:", N_obs, "/ Total:", NT, "\n")

# Build X matrix from model panel (already standardised within country)
model_panel <- read_csv("output/three_country_model_panel.csv",
                         show_col_types = FALSE)

# Join covariates in the same order as panel_pooled
cov_names <- c("elev_std", "rainfall_std", "ntl_std", "temp_std",
               "travel_std", "pop_dens_std", "poverty_std")
K <- length(cov_names)

X_pooled <- panel_pooled %>%
  select(-any_of(cov_names)) %>% 
  left_join(model_panel %>% select(country, district_id, time, all_of(cov_names)),
            by = c("country", 
                   "district_id" = "district_id",
                   "round_id" = "time")) %>%
  select(all_of(cov_names)) %>%
  as.matrix()


cat("  X matrix:", nrow(X_pooled), "x", ncol(X_pooled), "\n")
cat("  NAs in X:", sum(is.na(X_pooled)), "\n")

# Assemble Stan data
stan_data_pooled <- list(
  N       = D_all,
  T       = 2L,
  NT      = NT,
  C       = 3L,
  K       = K,
  N_obs   = N_obs,
  obs_idx = obs_idx,
  y       = panel_pooled$y_i[obs_idx],
  psi     = panel_pooled$psi_i[obs_idx],
  district = panel_pooled$global_district_id,
  time     = panel_pooled$round_id,
  country  = country_vec,
  X        = X_pooled,
  N_edges  = el$N_edges,
  node1    = el$node1,
  node2    = el$node2,
  scaling_factor = scaling_factor_all
)

cat("\n  Stan data assembled.\n")
cat("    N =", stan_data_pooled$N, "\n")
cat("    NT =", stan_data_pooled$NT, "\n")
cat("    C =", stan_data_pooled$C, "\n")
cat("    K =", stan_data_pooled$K, "\n")
cat("    N_obs =", stan_data_pooled$N_obs, "\n")
cat("    N_edges =", stan_data_pooled$N_edges, "\n")


###############################################################################
# STEP 3: Compile and fit
###############################################################################

cat("\n", strrep("=", 70), "\n")
cat("  STEP 3: Fitting pooled model\n")
cat(strrep("=", 70), "\n\n")

mod_pooled <- cmdstan_model("fh_bym2_pooled_3country.stan")

mod_pooled$check_syntax(pedantic = TRUE)

# The original (un-revised) Model 4 fit (fit_pooled) has been removed.
# Model 4 (revised) replaces it; see the Revised Models section below.
# mod_pooled is still compiled above so the Stan program is documented,
# but not fit here.


```

# Three-Country Cross-Border Models {#sec:cross-border-models}

The country-level models of Section \ref{sec:model2} treat each country as a
spatial island and ignore the documented cross-border ecology of malaria
transmission. This section introduces a sequence of three-country
specifications fitted on the combined 724-district adjacency graph
constructed in Section \ref{sec:adjacency}. Model 3 (three-country
single-time) is fitted to the temporally aligned second survey round and
serves as the methodologically conservative cross-border specification.
Model 4 (three-country two-time) extends Model 3 to both survey rounds with
a continuous calendar-year fixed effect, so that country-specific temporal
terms represent annual rates of change rather than wave-to-wave jumps. This
parameterisation follows the discrete spatio-temporal framework of
[@wakefield2020], which recommends that temporal effects in models that
combine surveys at different times reflect calendar time rather than survey
sequence numbers. Two further specifications relax modelling assumptions
made in Model 4: Model 4a allows covariate slopes to vary between rounds,
and Model 4b replaces the shared BYM2 hyperparameters with
country-specific spatial variance and mixing.

## BYM2 spatial prior

The BYM2 specification of [@riebler2016] decomposes the spatial random
effect into a structured component on an ICAR prior over the
adjacency graph and an unstructured Gaussian component, with a mixing
parameter $\phi$ that lets the data estimate their relative contribution.
For a district $i$,
$$b_i = \sigma_b \left( \sqrt{1 - \phi}\, v_i + \sqrt{\phi / s}\, u_i \right),$$
with $v_i \sim \mathcal{N}(0, 1)$, $u \sim \text{ICAR}(W)$ on the
combined adjacency $W$, and $s$ the scaling factor of [@riebler2016]
computed on $W$. Setting $\phi = 0$ recovers a purely unstructured model
and $\phi = 1$ recovers a pure ICAR model. The cross-sectional baseline
of Section \ref{sec:model1} confirmed empirically that this parameterisation
spans both limiting cases on these data, with the leave-one-out comparison
preferring the BYM2 prior over either limit.

Three features of malaria epidemiology in West Africa motivate the
choice of a spatially structured residual for the cross-border models:

1. *Plasmodium falciparum* dominates transmission in the region and
   shares vector species of the *Anopheles gambiae* and *Anopheles
   funestus* complexes across the three study countries [@who2023].
   Vector dispersal and mobile human populations document cross-border
   transmission flow [@bhatt2015].

2. The West African monsoon drives a regional July to October
   transmission peak and a north to south climatic gradient from the
   Sahelian zone of long dry seasons to the Guinean coast with
   perennial transmission [@bhatt2015; @weiss2020]. The climatic
   drivers of vector breeding and parasite development are
   spatially smooth at the district scale.

3. The structural determinants used as model covariates (elevation,
   rainfall, temperature, urbanisation, accessibility, population
   density, socioeconomic position) are themselves spatially smooth,
   so the residual after conditioning on these covariates is
   expected to retain spatial autocorrelation reflecting unmeasured
   factors such as housing quality, intervention coverage history
   and vector species composition.

## Covariate set

Seven district-level covariates are included in the cross-border models,
selected to capture the dominant ecological and structural determinants of
malaria transmission identified in [@bhatt2015] and [@weiss2020]. Covariate
values for each district-round were extracted using the actual DHS
fieldwork months of that survey, so seasonal variation in rainfall and
temperature is captured at the data level rather than through an explicit
seasonality term.

```{r covariate-justification-table, echo = FALSE}
covar_just <- data.frame(
  Covariate = c("Elevation",
                "Rainfall",
                "Temperature",
                "Nighttime lights",
                "Travel time to nearest city",
                "Log population density",
                "Multidimensional poverty"),
  Mechanism = c("Higher altitude lowers ambient temperature and slows P. falciparum development in Anopheles vectors; transmission is generally lower at higher elevations.",
                "Rainfall sustains vector breeding sites; total wet-season rainfall is a standard environmental driver in regional malaria mapping.",
                "Temperature governs both vector survival and the extrinsic incubation period of P. falciparum within the mosquito.",
                "Urbanisation reduces larval habitat availability and improves housing and healthcare access; nighttime lights from VIIRS proxy urban intensity.",
                "Accessibility from major population centres reflects the practical reach of interventions, diagnostics and treatment.",
                "Population density modifies effective transmission intensity and human-vector contact rates; the log transform handles the heavy right tail.",
                "A composite measure of household-level deprivation captures determinants of ITN ownership, care-seeking behaviour and housing quality."),
  Reference = c("@bhatt2015",
                "@weiss2020",
                "@bhatt2015",
                "@elvidge2014",
                "@weiss2020",
                "@who2023",
                "@florey2014"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
knitr::kable(covar_just, booktabs = TRUE,
             caption = "Covariates used in the cross-border models and their substantive justification.") |>
  kableExtra::kable_styling(latex_options = c("hold_position", "scale_down"))
```

## Common methodology across the cross-border models

All cross-border models share the Fay-Herriot logit-Normal likelihood from
Section \ref{sec:methodology}, the combined adjacency graph $W$ from
Section \ref{sec:adjacency} (Models 4 and 4a use the full graph; Model 4b
uses the within-country subgraph), the per-country covariate panel from
Section \ref{sec:covariates}, country intercepts $\alpha_c$ with
sum-to-zero regional constraint, and hierarchical covariate slopes
$\beta_{c,k} \sim \mathcal{N}(\bar{\beta}_k, \tau_k^2)$. The models differ
in (i) the inclusion or exclusion of Round 1 data, (ii) the
parameterisation of the temporal effect, (iii) whether covariate slopes
are time-invariant, and (iv) whether the BYM2 hyperparameters are shared
or country-specific.

Priors used throughout the cross-border models are:
$\mu \sim \mathcal{N}(-1.85, 1.5^2)$;
$\sigma_b \sim \text{Exp}(4.6)$;
$\phi \sim \text{Beta}(3, 2)$;
$\bar\beta_k \sim \mathcal{N}(0, 1)$;
$\tau_k \sim \text{Exp}(2)$;
$\sigma_\gamma \sim \text{Exp}(2)$.
The temporal hyperparameters introduced in Model 4 use
$\bar\delta \sim \mathcal{N}(0, 0.5^2)$ and
$\sigma_\delta \sim \text{Exp}(5)$, with the prior on $\bar\delta$ tighter
than on the round-indicator shift because $\bar\delta$ is an annual rate.

# Model 3: Three-country single-time {#sec:model3}

Model 3 fits the BYM2 spatial model jointly across all three countries
using only the temporally aligned second survey round (Burkina Faso 2021,
Cote d'Ivoire 2021, Ghana 2022). With every district in the model
surveyed within a 1-year window, cross-border smoothing operates on
contemporaneous surfaces by construction and does not mix space with time.

## Specification

For each district $i$ observed in Round 2,
$$y_i \sim \mathcal{N}(\theta_i, \psi_i^2),$$
$$\theta_i = \mu + \alpha_{c[i]} + b_i + \mathbf{x}_i^{\top}\bm{\beta}_{c[i]},$$
$$b_i = \sigma_b\!\left(\sqrt{1-\phi}\,v_i + \sqrt{\phi/s}\,u_i\right),
\qquad u \sim \mathrm{ICAR}(W),$$
$$\beta_{ck} \sim \mathcal{N}(\bar{\beta}_k, \tau_k^2),$$
with $\sum_c \alpha_c = 0$ and $\sum_i u_i = 0$. The graph $W$ is the
combined 724-district adjacency including 78 cross-border edges; the
scaling factor $s$ is computed on this graph.

## Stan model

```{r m3-stan-code, eval = FALSE}
mod_r2 <- cmdstan_model("fh_bym2_pooled_3country_round2.stan")
```

## Data assembly

```{r m3-build-data}
panel_r2 <- panel_pooled %>% filter(round_id == 2L)
X_r2     <- X_pooled[panel_pooled$round_id == 2L, , drop = FALSE]

obs_idx_r2 <- which(panel_r2$observed == 1L)
y_r2       <- panel_r2$y_logit[obs_idx_r2]
psi_r2     <- panel_r2$psi[obs_idx_r2]

stan_data_r2 <- list(
  N             = D_all,
  C             = 3L,
  K             = ncol(X_r2),
  N_obs         = length(obs_idx_r2),
  obs_idx       = obs_idx_r2,
  y             = y_r2,
  psi           = psi_r2,
  district      = panel_r2$global_district_id,
  country       = country_vec,
  X             = X_r2,
  N_edges       = el$N_edges,
  node1         = el$node1,
  node2         = el$node2,
  scaling_factor = scaling_factor_all
)
```

## Fit

```{r m3-fit, eval = FALSE}
mod_r2 <- cmdstan_model("fh_bym2_pooled_3country_round2.stan")
mod_r2$check_syntax(pedantic = TRUE)

fit_r2 <- mod_r2$sample(
  data            = stan_data_r2,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 1500,
  iter_sampling   = 2000,
  adapt_delta     = 0.97,
  max_treedepth   = 13,
  refresh         = 200,
  seed            = 20260612
)

fit_r2 <- ensure_fit(fit_r2, "m3_single_time")
```

## Posterior summary

```{r m3-summary}
if (!exists("fit_r2") || is.null(fit_r2)) {
  fit_r2 <- restore_fit("m3_single_time")
}
if (is.null(fit_r2)) {
  cat("[m3-summary: no cached fit at output/fits/m3_single_time.rds; run the fit chunk to populate]\n")
} else {
  param_summary(
    fit_r2,
    "Model 3: three-country single-time key parameters",
    keep = c("mu", "sigma_b", "phi", "alpha", "beta_bar", "tau", "p_national")
  )
}
```

## Model 3 Prevalence Map

```{r m3-prevalence-map, fig.height=6, fig.width=8}
# Posterior median prevalence under Model 3 (three-country, Round 2 only)
# joined onto the combined shapefile.
if (!exists("fit_r2") || is.null(fit_r2)) {
  fit_r2 <- restore_fit("m3_single_time")
}

if (!is.null(fit_r2) && "prevalence" %in% fit_r2$metadata()$stan_variables) {
  prev_v <- get_prev(fit_r2)
  m3_prev_df <- tibble(
    global_id = seq_along(prev_v),
    prev      = prev_v
  )
  m3_map_df <- all_sf %>% left_join(m3_prev_df, by = "global_id")

  p_m3 <- ggplot(m3_map_df) +
    geom_sf(aes(fill = prev), colour = "white", linewidth = 0.1) +
    prev_grad(name = "Prevalence") +
    labs(title = "Model 3 (three-country single-time): posterior median prevalence",
         subtitle = "Round 2 only. Shared BYM2 prior on the combined 724-district graph.") +
    theme_void(base_size = 11) +
    theme(legend.position = "bottom",
          strip.text = element_text(size = 12, face = "bold"))

  ggsave("output/figures/m3_prevalence_map.png", p_m3,
         width = 10, height = 6, dpi = 300, bg = "white")
  p_m3
} else {
  cat("[m3-prevalence-map: no cached fit at output/fits/m3_single_time.rds; run the fit chunk to populate]\n")
}
```

# Model 4: Three-country two-time, calendar year {#sec:model4}

Model 4 extends Model 3 to both survey rounds. The country-specific
temporal effect is parameterised as a linear slope on the actual calendar
year of each survey, centred at 2016 for numerical convenience. The
between-round odds ratio for country $c$ is then $\exp(\delta_c \cdot
\Delta\text{yr}_c)$ where $\Delta\text{yr}_c$ is the country's actual
elapsed years between Round 1 and Round 2, so country comparisons are
grounded in calendar time rather than in the round indicator.

## Specification

$$y_{it} \sim \mathcal{N}(\theta_{it}, \psi_{it}^2),$$
$$\theta_{it} = \mu + \alpha_{c[i]} + b_i
              + \delta_{c[i]} \cdot \widetilde{\text{yr}}_{it}
              + \gamma_{it} + \mathbf{x}_{it}^{\top}\bm{\beta}_{c[i]},$$
$$b_i = \sigma_b\!\left(\sqrt{1-\phi}\,v_i + \sqrt{\phi/s}\,u_i\right),
\qquad u \sim \mathrm{ICAR}(W),$$
$$\delta_c \sim \mathcal{N}(\bar\delta, \sigma_\delta^2),$$
$$\beta_{ck} \sim \mathcal{N}(\bar\beta_k, \tau_k^2),$$
with $\sum_c \alpha_c = 0$ and $\sum_i u_i = 0$, and where
$\widetilde{\text{yr}}_{it} = \text{yr}_{it} - 2016$. The Knorr-Held
Type I interaction $\gamma_{it} \sim \mathcal{N}(0, \sigma_\gamma^2)$ is
unstructured in both space and time. The single time-invariant BYM2
spatial field $b_i$ assumes that the residual spatial structure within
each country is stable over the 2010-2022 study window. This assumption
is empirically tested in Model 4a (Section \ref{sec:model4a}), in the
edge-deletion sensitivity (Section \ref{sec:sensitivity-edges}), and in
Model 4b (Section \ref{sec:model4b}).

## Stan model

```{r m4-stan-code}
m4_stan_code <- '
data {
  int<lower=1> N;
  int<lower=1> T;
  int<lower=1> NT;
  int<lower=1> C;
  int<lower=1> K;

  int<lower=1> N_obs;
  array[N_obs] int<lower=1, upper=NT> obs_idx;
  vector[N_obs] y;
  vector<lower=0>[N_obs] psi;

  array[NT] int<lower=1, upper=N> district;
  array[NT] int<lower=1, upper=T> time;
  vector[NT] year_c;
  array[N] int<lower=1, upper=C> country;

  matrix[NT, K] X;

  int<lower=1> N_edges;
  array[N_edges] int<lower=1, upper=N> node1;
  array[N_edges] int<lower=1, upper=N> node2;
  real<lower=0> scaling_factor;
}

parameters {
  real mu;
  vector[C - 1] alpha_raw;

  vector[C] delta;
  real delta_bar;
  real<lower=0> sigma_delta;

  vector[N] v;
  vector[N] u_raw;
  real<lower=0> sigma_b;
  real<lower=0, upper=1> phi;

  matrix[C, K] beta_raw;
  vector[K] beta_bar;
  vector<lower=0>[K] tau;

  vector[NT] gamma_raw;
  real<lower=0> sigma_gamma;
}

transformed parameters {
  vector[C] alpha;
  alpha[1] = 0;
  alpha[2:C] = alpha_raw;

  matrix[C, K] beta;
  for (k in 1:K)
    for (c in 1:C)
      beta[c, k] = beta_bar[k] + tau[k] * beta_raw[c, k];

  vector[N] u = u_raw - mean(u_raw);
  vector[N] b = sigma_b * (sqrt(1 - phi) * v
                           + sqrt(phi / scaling_factor) * u);

  vector[NT] gamma_star = sigma_gamma * gamma_raw;

  vector[NT] theta;
  for (n in 1:NT) {
    int d = district[n];
    int c = country[d];

    real xb = 0;
    for (k in 1:K)
      xb += X[n, k] * beta[c, k];

    theta[n] = mu + alpha[c] + b[d]
               + delta[c] * year_c[n]
               + gamma_star[n]
               + xb;
  }
}

model {
  mu ~ normal(-1.85, 1.5);
  alpha_raw ~ normal(0, 1.5);

  delta_bar ~ normal(0, 0.5);
  sigma_delta ~ exponential(5);
  delta ~ normal(delta_bar, sigma_delta);

  sigma_b ~ exponential(4.6);
  phi ~ beta(3, 2);
  v ~ std_normal();
  target += -0.5 * dot_self(u_raw[node1] - u_raw[node2]);
  sum(u_raw) ~ normal(0, 0.001 * N);

  beta_bar ~ normal(0, 1);
  tau ~ exponential(2);
  to_vector(beta_raw) ~ std_normal();

  sigma_gamma ~ exponential(2);
  gamma_raw ~ std_normal();

  y ~ normal(theta[obs_idx], psi);
}

generated quantities {
  vector[N_obs] log_lik;
  for (i in 1:N_obs)
    log_lik[i] = normal_lpdf(y[i] | theta[obs_idx[i]], psi[i]);

  vector[NT] prevalence;
  for (n in 1:NT)
    prevalence[n] = inv_logit(theta[n]);

  matrix[C, T] p_national;
  for (c in 1:C) {
    for (t in 1:T) {
      real total = 0;
      int count = 0;
      for (n in 1:NT) {
        if (country[district[n]] == c && time[n] == t) {
          total += prevalence[n];
          count += 1;
        }
      }
      p_national[c, t] = total / count;
    }
  }

  vector[C] odds_ratio_per_country;
  for (c in 1:C) {
    real y1 = 0;
    real y2 = 0;
    int found1 = 0;
    int found2 = 0;
    for (n in 1:NT) {
      if (country[district[n]] == c) {
        if (time[n] == 1 && found1 == 0) { y1 = year_c[n]; found1 = 1; }
        if (time[n] == 2 && found2 == 0) { y2 = year_c[n]; found2 = 1; }
      }
    }
    odds_ratio_per_country[c] = exp(delta[c] * (y2 - y1));
  }
}
'

writeLines(m4_stan_code, "fh_bym2_3country_two_time_calyr.stan")
```

## Build the calendar-year panel

```{r m4-build-cy}
year_lookup <- survey_index %>%
  transmute(country = iso, round_id, survey_year = year)

panel_pooled_cy <- panel_pooled %>%
  left_join(year_lookup, by = c("country", "round_id"))

stopifnot(all(!is.na(panel_pooled_cy$survey_year)))

panel_pooled_cy$year_c <- panel_pooled_cy$survey_year - 2016
```

## Data assembly

```{r m4-stan-data}
stan_data_m4_revised <- stan_data_pooled
stan_data_m4_revised$year_c <- panel_pooled_cy$year_c
stopifnot(length(stan_data_m4_revised$year_c) == stan_data_pooled$NT)
```

## Fit

```{r m4-fit, eval = FALSE}
mod_m4_revised <- cmdstan_model("fh_bym2_3country_two_time_calyr.stan")
mod_m4_revised$check_syntax(pedantic = TRUE)

fit_m4_revised <- mod_m4_revised$sample(
  data            = stan_data_m4_revised,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 1500,
  iter_sampling   = 2000,
  adapt_delta     = 0.97,
  max_treedepth   = 13,
  refresh         = 200,
  seed            = 20260612
)

fit_m4_revised <- ensure_fit(fit_m4_revised, "m4_revised_calyr")
```

## Posterior summary

```{r m4-summary}
if (!exists("fit_m4_revised") || is.null(fit_m4_revised)) {
  fit_m4_revised <- restore_fit("m4_revised_calyr")
}
if (is.null(fit_m4_revised)) {
  cat("[m4-summary: no cached fit at output/fits/m4_revised_calyr.rds; run the fit chunk to populate]\n")
} else {
  param_summary(
    fit_m4_revised,
    "Model 4: three-country two-time key parameters",
    keep = c("mu", "sigma_b", "phi", "sigma_gamma",
             "alpha", "delta", "delta_bar", "sigma_delta",
             "beta_bar", "tau",
             "p_national", "odds_ratio_per_country")
  )
}
```

## Model 4 Prevalence Maps

```{r m4-prevalence-maps, fig.height=6, fig.width=12}
# Posterior median prevalence under Model 4 (three-country, two-time) for
# both rounds, joined onto the combined shapefile. The prevalence vector has
# length NT = 1448 (724 districts x 2 rounds); we use panel_pooled_cy (which
# already carries the calendar year alongside round_id and country) to map
# each row of prevalence onto the corresponding (global district, round).
if (!exists("fit_m4_revised") || is.null(fit_m4_revised)) {
  fit_m4_revised <- restore_fit("m4_revised_calyr")
}

if (!is.null(fit_m4_revised) && "prevalence" %in% fit_m4_revised$metadata()$stan_variables) {
  prev_v <- get_prev(fit_m4_revised)
  stopifnot(length(prev_v) == nrow(panel_pooled_cy))

  m4_prev_long <- panel_pooled_cy %>%
    mutate(prev = prev_v) %>%
    select(global_id = global_district_id, round_id, prev)

  m4_map_df <- all_sf %>%
    left_join(m4_prev_long, by = "global_id")

  p_m4 <- ggplot(m4_map_df) +
    geom_sf(aes(fill = prev), colour = "white", linewidth = 0.1) +
    prev_grad(name = "Prevalence") +
    facet_wrap(~ paste("Round", round_id), nrow = 1) +
    labs(title = "Model 4 (three-country two-time, calendar year): posterior median prevalence",
         subtitle = "Shared BYM2 prior on the combined 724-district graph; country-specific calendar-year slopes.") +
    theme_void(base_size = 11) +
    theme(legend.position = "bottom",
          strip.text = element_text(size = 12, face = "bold"))

  ggsave("output/figures/m4_prevalence_maps_3country.png", p_m4,
         width = 12, height = 6, dpi = 300, bg = "white")
  p_m4
} else {
  cat("[m4-prevalence-maps: no cached fit at output/fits/m4_revised_calyr.rds; run the fit chunk to populate]\n")
}
```

# Model 4a: Time-varying covariate slopes {#sec:model4a}

Model 4 assumes the covariate slopes $\bm{\beta}_{c}$ are time-invariant.
Model 4a relaxes this by allowing country-specific covariate slopes to
differ between rounds, giving 42 slope parameters partially pooled around
seven regional means. The rest of the specification is identical to Model 4.

## Specification

$$y_{it} \sim \mathcal{N}(\theta_{it}, \psi_{it}^2),$$
$$\theta_{it} = \mu + \alpha_{c[i]} + b_i
              + \delta_{c[i]} \cdot \widetilde{\text{yr}}_{it}
              + \gamma_{it} + \mathbf{x}_{it}^{\top}\bm{\beta}_{c[i], t[i]},$$
$$\beta_{c,t,k} \sim \mathcal{N}(\bar\beta_k, \tau_k^2).$$
The spatial field $b_i$, country intercepts $\alpha_c$ with regional
sum-to-zero, country-year slopes $\delta_c \sim \mathcal{N}(\bar\delta,
\sigma_\delta^2)$, and the Knorr-Held Type I interaction $\gamma_{it}$
are unchanged from Model 4.

## Stan model

```{r m4a-stan-code}
m4a_stan_code <- '
data {
  int<lower=1> N;
  int<lower=1> T;
  int<lower=1> NT;
  int<lower=1> C;
  int<lower=1> K;

  int<lower=1> N_obs;
  array[N_obs] int<lower=1, upper=NT> obs_idx;
  vector[N_obs] y;
  vector<lower=0>[N_obs] psi;

  array[NT] int<lower=1, upper=N> district;
  array[NT] int<lower=1, upper=T> time;
  vector[NT] year_c;
  array[N] int<lower=1, upper=C> country;

  matrix[NT, K] X;

  int<lower=1> N_edges;
  array[N_edges] int<lower=1, upper=N> node1;
  array[N_edges] int<lower=1, upper=N> node2;
  real<lower=0> scaling_factor;
}

parameters {
  real mu;
  vector[C - 1] alpha_raw;

  vector[C] delta;
  real delta_bar;
  real<lower=0> sigma_delta;

  vector[N] v;
  vector[N] u_raw;
  real<lower=0> sigma_b;
  real<lower=0, upper=1> phi;

  array[C, T] vector[K] beta_raw;
  vector[K] beta_bar;
  vector<lower=0>[K] tau;

  vector[NT] gamma_raw;
  real<lower=0> sigma_gamma;
}

transformed parameters {
  vector[C] alpha;
  alpha[1] = 0;
  alpha[2:C] = alpha_raw;

  array[C, T] vector[K] beta;
  for (c in 1:C)
    for (t in 1:T)
      for (k in 1:K)
        beta[c, t, k] = beta_bar[k] + tau[k] * beta_raw[c, t, k];

  vector[N] u = u_raw - mean(u_raw);
  vector[N] b = sigma_b * (sqrt(1 - phi) * v
                           + sqrt(phi / scaling_factor) * u);

  vector[NT] gamma_star = sigma_gamma * gamma_raw;

  vector[NT] theta;
  for (n in 1:NT) {
    int d = district[n];
    int t = time[n];
    int c = country[d];

    real xb = 0;
    for (k in 1:K)
      xb += X[n, k] * beta[c, t, k];

    theta[n] = mu + alpha[c] + b[d]
               + delta[c] * year_c[n]
               + gamma_star[n]
               + xb;
  }
}

model {
  mu ~ normal(-1.85, 1.5);
  alpha_raw ~ normal(0, 1.5);

  delta_bar ~ normal(0, 0.5);
  sigma_delta ~ exponential(5);
  delta ~ normal(delta_bar, sigma_delta);

  sigma_b ~ exponential(4.6);
  phi ~ beta(3, 2);
  v ~ std_normal();
  target += -0.5 * dot_self(u_raw[node1] - u_raw[node2]);
  sum(u_raw) ~ normal(0, 0.001 * N);

  beta_bar ~ normal(0, 1);
  tau ~ exponential(2);
  for (c in 1:C)
    for (t in 1:T)
      beta_raw[c, t] ~ std_normal();

  sigma_gamma ~ exponential(2);
  gamma_raw ~ std_normal();

  y ~ normal(theta[obs_idx], psi);
}

generated quantities {
  vector[N_obs] log_lik;
  for (i in 1:N_obs)
    log_lik[i] = normal_lpdf(y[i] | theta[obs_idx[i]], psi[i]);
}
'

writeLines(m4a_stan_code, "fh_bym2_3country_two_time_calyr_4a.stan")
```

## Fit

```{r m4a-fit, eval = FALSE}
mod_m4a <- cmdstan_model("fh_bym2_3country_two_time_calyr_4a.stan")
mod_m4a$check_syntax(pedantic = TRUE)

fit_m4a <- mod_m4a$sample(
  data            = stan_data_m4_revised,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 1500,
  iter_sampling   = 2000,
  adapt_delta     = 0.97,
  max_treedepth   = 13,
  refresh         = 200,
  seed            = 20260612
)

fit_m4a <- ensure_fit(fit_m4a, "m4a_timevarying_betas")
```

## Posterior summary

```{r m4a-summary}
if (!exists("fit_m4a") || is.null(fit_m4a)) {
  fit_m4a <- restore_fit("m4a_timevarying_betas")
}
if (is.null(fit_m4a)) {
  cat("[m4a-summary: no cached fit at output/fits/m4a_timevarying_betas.rds; run the fit chunk to populate]\n")
} else {
  param_summary(
    fit_m4a,
    "Model 4a: time-varying covariate slopes key parameters",
    keep = c("mu", "sigma_b", "phi", "sigma_gamma",
             "alpha", "delta", "delta_bar", "sigma_delta",
             "beta_bar", "tau")
  )
}
```

## Time-varying covariate slopes

```{r m4a-beta-table}
if (!exists("fit_m4a") || is.null(fit_m4a)) {
  fit_m4a <- restore_fit("m4a_timevarying_betas")
}
if (is.null(fit_m4a)) {
  cat("[m4a-beta-table: no cached fit at output/fits/m4a_timevarying_betas.rds; run the fit chunk to populate]\n")
} else {
  beta_4a <- posterior::summarise_draws(
    fit_m4a$draws(variables = "beta"),
    median = median,
    q5  = ~ quantile(.x, probs = 0.05, names = FALSE),
    q95 = ~ quantile(.x, probs = 0.95, names = FALSE)
  )
  beta_4a <- as.data.frame(beta_4a)
  
  country_names <- c("GHA", "CIV", "BFA")
  cov_names <- c("Elevation", "Rainfall", "Nighttime lights", "Temperature",
                 "Travel time", "Pop. density", "Poverty")
  
  parsed <- regmatches(beta_4a$variable,
                       regexec("beta\\[(\\d+),(\\d+),(\\d+)\\]", beta_4a$variable))
  beta_4a$country   <- country_names[as.integer(sapply(parsed, `[`, 2))]
  beta_4a$round     <- paste0("R", sapply(parsed, `[`, 3))
  beta_4a$covariate <- cov_names[as.integer(sapply(parsed, `[`, 4))]
  
  library(dplyr); library(tidyr)
  beta_4a_wide <- beta_4a |>
    mutate(value = sprintf("%.2f [%.2f, %.2f]", median, q5, q95)) |>
    select(country, covariate, round, value) |>
    pivot_wider(names_from = round, values_from = value) |>
    arrange(country, covariate)
  
  knitr::kable(beta_4a_wide, booktabs = TRUE,
               caption = "Model 4a country-specific covariate slopes by round. Posterior median with 90 per cent credible interval.")
}
```

# Model 4b: Country-specific BYM2 Components {#sec:model4b}

The edge-deletion sensitivity model (Section \ref{sec:sensitivity-edges}, below)
removes cross-border edges but retains a single set of BYM2 hyperparameters
$(\sigma_b, \phi)$ shared across the three countries. A further relaxation
following the disconnected-graph framework of [@freni2018] is to allow each
country's spatial component to have its own variance $\sigma_{b,c}$ and its
own mixing parameter $\phi_c$. This treats the three countries as
substantively distinct in their malaria epidemiology, intervention history
and ecological setting, so that the magnitude and the
structured-versus-unstructured split of residual spatial variation may
differ between countries. Country-specific BYM2 hyperparameters are the
standard treatment for disconnected graphs in [@freni2018] and have been
implemented in the Donegan ICAR/BYM2 Stan code [@donegan2021].

## Specification

For each district-round observation,
$$y_{it} \sim \mathcal{N}(\theta_{it}, \psi_{it}^2),$$
$$\theta_{it} = \mu + \alpha_{c[i]} + b_i
              + \delta_{c[i]} \cdot \widetilde{\text{yr}}_{it}
              + \gamma_{it} + \mathbf{x}_{it}^{\top}\bm{\beta}_{c[i]}.$$
The country-indexed BYM2 spatial term is
$$b_i = \sigma_{b,\,c[i]}\!\left(\sqrt{1 - \phi_{c[i]}}\,v_i
        + \sqrt{\phi_{c[i]} / s_{c[i]}}\,u_i\right),$$
with $v_i \sim \mathcal{N}(0, 1)$ and $u$ following an ICAR prior on the
within-country adjacency $W'$:
$$p(u \mid W') \propto
   \exp\!\left(-\tfrac{1}{2}\sum_{(i,j) \in E_{\text{within}}}(u_i - u_j)^2\right),$$
with per-country sum-to-zero constraints $\sum_{i \in c} u_i = 0$ for each
$c$, and per-country scaling factor $s_c$ computed from country $c$'s
within-country subgraph.

The country-specific hyperparameters have weakly informative priors,
$$\sigma_{b,c} \sim \text{Exponential}(4.6), \qquad
  \phi_c \sim \text{Beta}(3, 2).$$
All other components are exactly as in Model 4: country intercepts
$\alpha_c$ with regional sum-to-zero, country-year linear slopes
$\delta_c \sim \mathcal{N}(\bar\delta, \sigma_\delta^2)$, the Knorr-Held
Type I interaction $\gamma_{it} \sim \mathcal{N}(0, \sigma_\gamma^2)$, and
hierarchical covariate slopes $\beta_{c,k} \sim \mathcal{N}(\bar\beta_k,
\tau_k^2)$.

## Stan model

```{r m4b-stan-code}
m4b_stan_code <- '
data {
  int<lower=1> N;
  int<lower=1> T;
  int<lower=1> NT;
  int<lower=1> C;
  int<lower=1> K;

  int<lower=1> N_obs;
  array[N_obs] int<lower=1, upper=NT> obs_idx;
  vector[N_obs] y;
  vector<lower=0>[N_obs] psi;

  array[NT] int<lower=1, upper=N> district;
  array[NT] int<lower=1, upper=T> time;
  vector[NT] year_c;
  array[N] int<lower=1, upper=C> country;

  matrix[NT, K] X;

  int<lower=1> N_edges;
  array[N_edges] int<lower=1, upper=N> node1;
  array[N_edges] int<lower=1, upper=N> node2;
  vector<lower=0>[C] scaling_factor_country;
}

parameters {
  real mu;
  vector[C - 1] alpha_raw;

  vector[C] delta;
  real delta_bar;
  real<lower=0> sigma_delta;

  vector[N] v;
  vector[N] u_raw;
  vector<lower=0>[C] sigma_b;
  vector<lower=0, upper=1>[C] phi;

  matrix[C, K] beta_raw;
  vector[K] beta_bar;
  vector<lower=0>[K] tau;

  vector[NT] gamma_raw;
  real<lower=0> sigma_gamma;
}

transformed parameters {
  vector[C] alpha;
  alpha[1] = 0;
  alpha[2:C] = alpha_raw;

  matrix[C, K] beta;
  for (k in 1:K)
    for (c in 1:C)
      beta[c, k] = beta_bar[k] + tau[k] * beta_raw[c, k];

  vector[N] u = u_raw - mean(u_raw);

  vector[N] b;
  for (i in 1:N) {
    int c = country[i];
    b[i] = sigma_b[c] * ( sqrt(1 - phi[c]) * v[i]
                          + sqrt(phi[c] / scaling_factor_country[c]) * u[i] );
  }

  vector[NT] gamma_star = sigma_gamma * gamma_raw;

  vector[NT] theta;
  for (n in 1:NT) {
    int d = district[n];
    int c = country[d];

    real xb = 0;
    for (k in 1:K)
      xb += X[n, k] * beta[c, k];

    theta[n] = mu + alpha[c] + b[d]
               + delta[c] * year_c[n]
               + gamma_star[n]
               + xb;
  }
}

model {
  mu ~ normal(-1.85, 1.5);
  alpha_raw ~ normal(0, 1.5);

  delta_bar ~ normal(0, 0.5);
  sigma_delta ~ exponential(5);
  delta ~ normal(delta_bar, sigma_delta);

  sigma_b ~ exponential(4.6);
  phi ~ beta(3, 2);
  v ~ std_normal();
  target += -0.5 * dot_self(u_raw[node1] - u_raw[node2]);
  sum(u_raw) ~ normal(0, 0.001 * N);

  beta_bar ~ normal(0, 1);
  tau ~ exponential(2);
  to_vector(beta_raw) ~ std_normal();

  sigma_gamma ~ exponential(2);
  gamma_raw ~ std_normal();

  y ~ normal(theta[obs_idx], psi);
}

generated quantities {
  vector[N_obs] log_lik;
  for (i in 1:N_obs)
    log_lik[i] = normal_lpdf(y[i] | theta[obs_idx[i]], psi[i]);

  vector[NT] prevalence;
  for (n in 1:NT)
    prevalence[n] = inv_logit(theta[n]);

  matrix[C, T] p_national;
  for (c in 1:C) {
    for (t in 1:T) {
      real total = 0;
      int count = 0;
      for (n in 1:NT) {
        if (country[district[n]] == c && time[n] == t) {
          total += prevalence[n];
          count += 1;
        }
      }
      p_national[c, t] = total / count;
    }
  }
}
'

writeLines(m4b_stan_code, "fh_bym2_3country_two_time_calyr_country_bym2.stan")
```

## Data assembly

```{r m4b-stan-data, eval = FALSE}
stopifnot(exists("node1_within"), exists("node2_within"),
          exists("sfac_by_country"), exists("stan_data_m4_revised"))

scaling_factor_country <- unname(sfac_by_country[as.character(1:3)])

stan_data_m4b <- stan_data_m4_revised
stan_data_m4b$node1                  <- node1_within
stan_data_m4b$node2                  <- node2_within
stan_data_m4b$N_edges                <- length(node1_within)
stan_data_m4b$scaling_factor         <- NULL
stan_data_m4b$scaling_factor_country <- scaling_factor_country
```

## Fit

```{r m4b-fit, eval = FALSE}
mod_m4b <- cmdstan_model("fh_bym2_3country_two_time_calyr_country_bym2.stan")
mod_m4b$check_syntax(pedantic = TRUE)

fit_m4b <- mod_m4b$sample(
  data            = stan_data_m4b,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 1500,
  iter_sampling   = 2000,
  adapt_delta     = 0.97,
  max_treedepth   = 13,
  refresh         = 200,
  seed            = 20260612
)

fit_m4b <- ensure_fit(fit_m4b, "m4b_country_bym2")
```

## Posterior summary

```{r m4b-summary}
if (!exists("fit_m4b") || is.null(fit_m4b)) {
  fit_m4b <- restore_fit("m4b_country_bym2")
}
if (is.null(fit_m4b)) {
  cat("[m4b-summary: no cached fit at output/fits/m4b_country_bym2.rds; run the fit chunk to populate]\n")
} else {
  param_summary(
    fit_m4b,
    "Model 4b: country-specific BYM2 key parameters",
    keep = c("mu", "sigma_b", "phi", "sigma_gamma",
             "alpha", "delta", "delta_bar", "sigma_delta",
             "beta_bar", "tau", "p_national")
  )
}
```




# Sensitivity to Cross-Border Adjacency {#sec:sensitivity-edges}

The three-country cross-border models include 78 cross-border edges
joining adjacent districts across national borders in the combined
724-district adjacency graph. As a sensitivity check on the contribution
of those edges, Model 4 is refitted with the cross-border edges removed,
leaving only the 1,978 within-country edges. This adjacency-structure
sensitivity follows the recommended practice of [@earnest2007] for
CAR-based disease mapping. Removing the cross-border edges disconnects
the combined graph into three components, one per country; the BYM2
scaling factor [@riebler2016] is recomputed per component and applied per
district following the disconnected-graph treatment of [@freni2018].

## Specification

The edge-deleted model has the same likelihood and linear predictor as
Model 4. The only modification is that the ICAR sum runs over within-country
edges only,
$$p(u \mid W') \propto
  \exp\!\left(-\tfrac{1}{2}\sum_{(i,j) \in E_{\text{within}}}(u_i - u_j)^2\right),$$
with per-country sum-to-zero constraints $\sum_{i \in c} u_i = 0$ and a
per-district scaling factor $s'_{c[i]}$ computed from the country
subgraph.

## Within-country edge list and per-country scaling

```{r ed-data, eval = FALSE}
edge_country_a <- all_sf$country[el$node1]
edge_country_b <- all_sf$country[el$node2]
within_mask    <- edge_country_a == edge_country_b

node1_within    <- el$node1[within_mask]
node2_within    <- el$node2[within_mask]
N_edges_within  <- length(node1_within)

compute_scaling_factor_subgraph <- function(node1, node2, n_nodes) {
  A <- Matrix::sparseMatrix(i = c(node1, node2), j = c(node2, node1),
                             x = 1, dims = c(n_nodes, n_nodes))
  D <- Matrix::Diagonal(x = Matrix::rowSums(A))
  Q <- D - A
  Q_scaled <- INLA::inla.scale.model(
    as(Q, "dgTMatrix"),
    constr = list(A = matrix(1, 1, n_nodes), e = 0)
  )
  exp(mean(log(Matrix::diag(Q_scaled))))
}

country_of_node <- all_sf$country_id
sf_by_country   <- split(seq_len(D_all), country_of_node)
sfac_by_country <- setNames(numeric(3), c("1", "2", "3"))

for (cc in 1:3) {
  nodes_cc <- sf_by_country[[as.character(cc)]]
  mask_cc  <- (el$node1 %in% nodes_cc) & (el$node2 %in% nodes_cc)
  local_id <- setNames(seq_along(nodes_cc), as.character(nodes_cc))
  n1_local <- local_id[as.character(el$node1[mask_cc])]
  n2_local <- local_id[as.character(el$node2[mask_cc])]
  sfac_by_country[as.character(cc)] <-
    compute_scaling_factor_subgraph(n1_local, n2_local, length(nodes_cc))
}

scaling_factor_per_district <- sfac_by_country[as.character(country_of_node)]
```

## Stan model

```{r ed-stan-code}
ed_stan_code <- '
data {
  int<lower=1> N;
  int<lower=1> T;
  int<lower=1> NT;
  int<lower=1> C;
  int<lower=1> K;

  int<lower=1> N_obs;
  array[N_obs] int<lower=1, upper=NT> obs_idx;
  vector[N_obs] y;
  vector<lower=0>[N_obs] psi;

  array[NT] int<lower=1, upper=N> district;
  array[NT] int<lower=1, upper=T> time;
  vector[NT] year_c;
  array[N] int<lower=1, upper=C> country;

  matrix[NT, K] X;

  int<lower=1> N_edges;
  array[N_edges] int<lower=1, upper=N> node1;
  array[N_edges] int<lower=1, upper=N> node2;
  vector<lower=0>[N] scaling_factor;
}

parameters {
  real mu;
  vector[C - 1] alpha_raw;
  vector[C] delta;
  real delta_bar;
  real<lower=0> sigma_delta;
  vector[N] v;
  vector[N] u_raw;
  real<lower=0> sigma_b;
  real<lower=0, upper=1> phi;
  matrix[C, K] beta_raw;
  vector[K] beta_bar;
  vector<lower=0>[K] tau;
  vector[NT] gamma_raw;
  real<lower=0> sigma_gamma;
}

transformed parameters {
  vector[C] alpha;
  alpha[1] = 0;
  alpha[2:C] = alpha_raw;

  matrix[C, K] beta;
  for (k in 1:K)
    for (c in 1:C)
      beta[c, k] = beta_bar[k] + tau[k] * beta_raw[c, k];

  vector[N] u = u_raw - mean(u_raw);
  vector[N] b = sigma_b * (sqrt(1 - phi) * v
                           + sqrt(phi ./ scaling_factor) .* u);

  vector[NT] gamma_star = sigma_gamma * gamma_raw;

  vector[NT] theta;
  for (n in 1:NT) {
    int d = district[n];
    int c = country[d];
    real xb = 0;
    for (k in 1:K)
      xb += X[n, k] * beta[c, k];
    theta[n] = mu + alpha[c] + b[d]
               + delta[c] * year_c[n]
               + gamma_star[n]
               + xb;
  }
}

model {
  mu ~ normal(-1.85, 1.5);
  alpha_raw ~ normal(0, 1.5);
  delta_bar ~ normal(0, 0.5);
  sigma_delta ~ exponential(5);
  delta ~ normal(delta_bar, sigma_delta);
  sigma_b ~ exponential(4.6);
  phi ~ beta(3, 2);
  v ~ std_normal();
  target += -0.5 * dot_self(u_raw[node1] - u_raw[node2]);
  sum(u_raw) ~ normal(0, 0.001 * N);
  beta_bar ~ normal(0, 1);
  tau ~ exponential(2);
  to_vector(beta_raw) ~ std_normal();
  sigma_gamma ~ exponential(2);
  gamma_raw ~ std_normal();
  y ~ normal(theta[obs_idx], psi);
}

generated quantities {
  vector[N_obs] log_lik;
  for (i in 1:N_obs)
    log_lik[i] = normal_lpdf(y[i] | theta[obs_idx[i]], psi[i]);
}
'

writeLines(ed_stan_code, "fh_bym2_3country_two_time_calyr_edge_del.stan")
```

## Fit

```{r ed-fit, eval = FALSE}
stan_data_edge_del <- stan_data_m4_revised
stan_data_edge_del$node1          <- node1_within
stan_data_edge_del$node2          <- node2_within
stan_data_edge_del$N_edges        <- N_edges_within
stan_data_edge_del$scaling_factor <- scaling_factor_per_district

mod_edge_del <- cmdstan_model("fh_bym2_3country_two_time_calyr_edge_del.stan")
mod_edge_del$check_syntax(pedantic = TRUE)

fit_edge_del <- mod_edge_del$sample(
  data            = stan_data_edge_del,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 1500,
  iter_sampling   = 2000,
  adapt_delta     = 0.97,
  max_treedepth   = 13,
  refresh         = 200,
  seed            = 20260612
)

fit_edge_del <- ensure_fit(fit_edge_del, "m4_edge_del")
```

## Posterior summary

```{r ed-summary}
if (!exists("fit_edge_del") || is.null(fit_edge_del)) {
  fit_edge_del <- restore_fit("m4_edge_del")
}
if (is.null(fit_edge_del)) {
  cat("[ed-summary: no cached fit at output/fits/m4_edge_del.rds; run the fit chunk to populate]\n")
} else {
  param_summary(
    fit_edge_del,
    "Model 4 edge-deletion sensitivity key parameters",
    keep = c("mu", "sigma_b", "phi", "sigma_gamma",
             "alpha", "delta", "delta_bar", "sigma_delta",
             "beta_bar", "tau")
  )
}
```

# Sensitivity Comparison Across Cross-Border Models {#sec:sensitivity-comparison}

The cross-border specifications fitted above explore four modelling
choices in turn: the time-invariance of the BYM2 spatial field (Model 4
versus Model 4a, which lets covariate slopes vary by round), the
contribution of cross-border adjacency (Model 4 versus the edge-deletion
sensitivity), and the assumption of shared BYM2 hyperparameters across
countries (Model 4 versus Model 4b). Each pair is compared by leave-one-out
cross-validation [@vehtari2017] on the expected log predictive density
(ELPD), using the `loo` package implementation. The comparisons within each
block use models fitted to the same observation set, which is the
requirement for LOO-CV comparability.

```{r sensitivity-loo-compare}
# Restore the fits from disk where they are not already in memory.
restore_to_global <- function(fit_var, disk_name) {
  if (!exists(fit_var, envir = .GlobalEnv) ||
      is.null(get(fit_var, envir = .GlobalEnv))) {
    restored <- restore_fit(disk_name)
    if (!is.null(restored)) assign(fit_var, restored, envir = .GlobalEnv)
  }
}
restore_to_global("fit_m4_revised", "m4_revised_calyr")
restore_to_global("fit_m4a",        "m4a_timevarying_betas")
restore_to_global("fit_m4b",        "m4b_country_bym2")
restore_to_global("fit_edge_del",   "m4_edge_del")

m4_family <- list()
for (nm_label in list(
       c("fit_m4_revised", "Model 4: shared BYM2, full graph"),
       c("fit_m4a",        "Model 4a: time-varying betas"),
       c("fit_m4b",        "Model 4b: country-specific BYM2"),
       c("fit_edge_del",   "Model 4 edge-deletion (within-country only)")
     )) {
  fit_name  <- nm_label[1]
  loo_label <- nm_label[2]
  if (!exists(fit_name) || is.null(get(fit_name))) next
  fit <- get(fit_name)
  if (!"log_lik" %in% fit$metadata()$stan_variables) next
  loo_obj <- tryCatch(fit$loo(), error = function(e) NULL)
  if (!is.null(loo_obj)) m4_family[[loo_label]] <- loo_obj
}

if (length(m4_family) >= 2L) {
  cmp <- loo::loo_compare(m4_family)
  cmp_df <- data.frame(
    Model     = rownames(cmp),
    elpd_diff = round(cmp[, "elpd_diff"], 1),
    se_diff   = round(cmp[, "se_diff"],   1),
    elpd_loo  = round(sapply(m4_family[rownames(cmp)],
                             function(l) l$estimates["elpd_loo", "Estimate"]), 1),
    se_elpd   = round(sapply(m4_family[rownames(cmp)],
                             function(l) l$estimates["elpd_loo", "SE"]), 1),
    p_loo     = round(sapply(m4_family[rownames(cmp)],
                             function(l) l$estimates["p_loo", "Estimate"]), 1),
    stringsAsFactors = FALSE,
    row.names = NULL
  )

  knitr::kable(
    cmp_df,
    booktabs = TRUE,
    col.names = c("Model", "ELPD diff", "SE diff", "ELPD", "SE(ELPD)", "p_loo"),
    caption = "Leave-one-out cross-validation across the cross-border two-time models."
  )
} else {
  cat("[sensitivity-loo-compare: fewer than 2 LOO objects available; cannot compare]\n")
}
```

The reading rule for the table is the standard one of [@vehtari2017]: a
difference in ELPD smaller than two standard errors of that difference is
not detectable, and the simpler model in the comparison is preferred on
parsimony grounds.



# External Validation against the Malaria Atlas Project {#sec:map-validation}

The Malaria Atlas Project [MAP; @bhatt2015; @weiss2020] publishes
model-based geostatistical surfaces of *Plasmodium falciparum* parasite
rate for sub-Saharan Africa using a different evidence base and a
different modelling tradition (unit-level point-referenced geostatistics
with continuous-time embedding) than the area-level Fay-Herriot framework
used here. The MAP surfaces provide an external reference against which
the district-level posterior estimates can be compared.

The comparison is framed as **concordance with an independent model**
rather than validation against ground truth, since neither source observes
true district prevalence directly. Two differences in target quantity are
acknowledged explicitly:

- MAP describes an age-standardised parasite rate in children aged two
  to ten years (PfPR$_{2-10}$), whereas the present estimates describe
  RDT positivity in children aged six to fifty-nine months. A level
  offset is therefore expected; the emphasis falls on spatial and rank
  agreement.
- MAP is a continuous-time geostatistical surface fitted to point-level
  data and aggregated; the present estimates are direct area-level
  posteriors. Aggregation noise contributes to any disagreement.

