---
title: "Bayesian Small Area Estimation of Under-5 Malaria Prevalence in West Africa"
author: ''
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
subtitle: "A Space-Time BYM2 Fay-Herriot Model Across Ghana, Burkina Faso, and Côte
  d'Ivoire"
bibliography: references.bib
link-citations: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6,
                      cache       = TRUE,
                      cache.path  = "bym2_cache/",
                      cache.lazy  = FALSE,
                      autodep     = TRUE)
# Core libraries
library(tidyverse)
library(haven)
library(sf)
library(geodata)
library(survey)
library(spdep)
library(Matrix)
library(cmdstanr)
library(bayesplot)
library(tidybayes)
library(patchwork)
library(ggcorrplot)
library(posterior)
```

```{r fit-persistence-helpers, include=FALSE}
dir.create("output/fits",    showWarnings = FALSE, recursive = TRUE)
dir.create("output/figures", showWarnings = FALSE, recursive = TRUE)

persist_fit <- function(fit, name) {
  path <- file.path("output/fits", paste0(name, ".rds"))
  tryCatch(
    fit$save_object(file = path),
    error = function(e) message("Could not persist fit ", name, ": ", conditionMessage(e))
  )
  invisible(fit)
}

is_fit_live <- function(fit) {
  if (is.null(fit)) return(FALSE)
  files <- tryCatch(fit$output_files(), error = function(e) character())
  length(files) > 0L && all(file.exists(files))
}

restore_fit <- function(name) {
  path <- file.path("output/fits", paste0(name, ".rds"))
  if (file.exists(path)) readRDS(path) else NULL
}

ensure_fit <- function(fit, name) {
    if (is_fit_live(fit)) {
    persist_fit(fit, name)
    return(fit)
  }
  restored <- restore_fit(name)
  if (!is.null(restored)) return(restored)
  warning("Fit `", name, "` has no live CSVs and no saved .rds; ",
          "subsequent calls may fail. Re-run the chunk that fits it.")
  fit
}
```


# Data {#sec:data}

## About the Demographic and Health Surveys (DHS)

The Demographic and Health Surveys are nationally representative household
surveys using a stratified two-stage cluster sampling design. The first
stage selects enumeration areas (clusters) within strata defined by
region and urban/rural status; the second stage selects households within
each cluster. Children aged 6--59 months are tested for malaria parasitaemia
using a rapid diagnostic test (RDT) [@florey2014; @croft2018]. Cluster
GPS coordinates are jittered by up to 5 km (urban) or 10 km (rural) for
respondent confidentiality. The household member recode (PR) file
contains the individual-level biomarker; the cluster GPS shapefile
provides the spatial reference.

## Surveys Used in This Study

| Country | Round 1 | Round 2 |
|---|---|---|
| Burkina Faso | DHS 2010 | DHS/MIS 2021 |
| Côte d'Ivoire | DHS 2011 | DHS/MIS 2021 |
| Ghana | DHS 2014 | DHS 2022 |

## District Boundaries

```{r district-boundaries}
load_districts <- function(iso) {
  if (iso == "GHA") {
    gadm(country = iso, level = 2, path = "boundaries/") %>%
      st_as_sf() %>% st_transform(crs = 4326) %>%
      rename(district_name = NAME_2, district_code = GID_2) %>%
      mutate(iso = iso)
  } else {
    gadm(country = iso, level = 3, path = "boundaries/") %>%
      st_as_sf() %>% st_transform(crs = 4326) %>%
      rename(district_name = NAME_3, district_code = GID_3) %>%
      mutate(iso = iso)
  }
}

districts_raw <- list(
  GHA = load_districts("GHA"),
  BFA = load_districts("BFA"),
  CIV = load_districts("CIV")
)

# Create numeric district lookups (Stan requires integer indices)
district_lookups <- map(districts_raw, function(d) {
  d %>% st_drop_geometry() %>%
    select(district_name, district_code, iso) %>%
    mutate(district_id = row_number())
})

# Districts ordered by district_id for adjacency matrix construction
districts_ordered <- map2(districts_raw, district_lookups, function(d, lk) {
  d %>% left_join(lk, by = c("district_name", "district_code", "iso")) %>%
    arrange(district_id)
})

map_dfr(district_lookups, ~tibble(districts = nrow(.x)), .id = "iso")
```

## Survey Index and PR Files

```{r survey-index}
survey_index <- tribble(
  ~country,        ~iso, ~year, ~round_id, ~pr_path, ~gps_path,
  "Ghana",         "GHA", 2014, 1L,
  "dhs_gh_14/GHPR72DT/GHPR72FL.DTA", "dhs_gh_14/GHGE71FL/GHGE71FL.shp",
  "Ghana",         "GHA", 2022, 2L,
  "dhs_gh_22/GHPR8CDT/GHPR8CFL.DTA", "dhs_gh_22/GHGE8AFL/GHGE8AFL.shp",
  "Burkina",       "BFA", 2010, 1L,
  "dhs_bf_10/BFPR62DT/BFPR62FL.DTA", "dhs_bf_10/BFGE61FL/BFGE61FL.shp",
  "Burkina",       "BFA", 2021, 2L,
  "dhs_bf_21/BFPR81DT/BFPR81FL.DTA", "dhs_bf_21/BFGE81FL/BFGE81FL.shp",
  "Cote d'Ivoire", "CIV", 2012, 1L,
  "dhs_ci_12/CIPR62DT/CIPR62FL.DTA", "dhs_ci_12/CIGE61FL/CIGE61FL.shp",
  "Cote d'Ivoire", "CIV", 2021, 2L,
  "dhs_ci_21/CIPR81DT/CIPR81FL.DTA", "dhs_ci_21/CIGE81FL/CIGE81FL.shp"
)
```

```{r pr-files}
load_pr <- function(pr_path, gps_path, iso, year, round_id,
                    districts_ordered, district_lookup) {
  gps <- read_sf(gps_path) %>%
    rename(cluster_id = DHSCLUST) %>%
    st_transform(crs = 4326)
  
  gps_joined <- st_join(gps, districts_ordered, join = st_nearest_feature) %>%
    select(cluster_id, district_name, district_code) %>%
    st_drop_geometry() %>%
    left_join(district_lookup, by = c("district_name", "district_code")) %>%
    select(cluster_id, district_id, district_name)
  
  pr <- read_dta(pr_path) %>%
    rename_with(tolower) %>%
    select(cluster_id = hv001, hh_id = hv002, line_id = hvidx,
           weight_raw = hv005, stratum = hv023,
           age_months = hc1, malaria = hml32) %>%
    filter(age_months < 60, malaria %in% c(0, 1)) %>%
    mutate(weight = weight_raw / 1e6,
           malaria = as.integer(malaria),
           cluster_id = as.integer(cluster_id),
           stratum = as.integer(stratum),
           country = iso, year = year, round_id = round_id) %>%
    left_join(gps_joined, by = "cluster_id")
  
  pr
}

pr_list <- pmap(survey_index, function(country, iso, year, round_id, pr_path, gps_path) {
  load_pr(pr_path, gps_path, iso, year, round_id,
          districts_ordered[[iso]], district_lookups[[iso]])
})

```

# Direct Estimation {#sec:direct}

For each observed district $i$ at time $t$, the design-based prevalence
is computed using the Hájek ratio estimator [@hajek1971],
$$\hat{p}_{it}^{w} = \frac{\sum_k d_k \, z_k}{\sum_k d_k},$$
where $z_k$ is the binary RDT outcome for child $k$ and $d_k$ is the
DHS sampling weight. Design-based standard errors are obtained via
Taylor-series linearisation [@binder1983] using the `survey` package
[@lumley2010], respecting the stratified two-stage cluster design.

For the Fay-Herriot likelihood [@wakefield2020], the prevalence is
transformed to the logit scale and the delta method is applied to the
SE:
$$y_{it} = \mathrm{logit}(\hat{p}_{it}^{w}), \qquad
\widehat{\mathrm{SE}}(y_{it}) =
  \frac{\widehat{\mathrm{SE}}(\hat{p}_{it}^{w})}
       {\hat{p}_{it}^{w}(1-\hat{p}_{it}^{w})}.$$
For numerical stability, prevalences are bounded to $[0.02, 0.98]$ and
logit-scale SEs to $[0.1, 2.0]$.

```{r direct-estimates}
get_direct_estimates <- function(pr) {
  design <- svydesign(
    ids = ~cluster_id, strata = ~stratum, weights = ~weight,
    data = pr, nest = TRUE
  )
  svyby(formula = ~malaria, by = ~district_id, design = design,
        FUN = svymean, na.rm = TRUE, keep.var = TRUE) %>%
    rename(p_hat = malaria, se_p = se) %>%
    filter(!is.na(p_hat), !is.na(se_p),
           se_p > 0, p_hat > 0, p_hat < 1) %>%
    mutate(
      country   = pr$country[1],
      iso       = pr$country[1],
      year      = pr$year[1],
      round_id  = pr$round_id[1],
      p_bounded = pmax(0.02, pmin(0.98, p_hat)),
      y_i       = log(p_bounded / (1 - p_bounded)),
      logit_var = se_p^2 / (pmax(0.02, p_hat)^2 * pmax(0.02, 1 - p_hat)^2),
      psi_i     = pmax(0.1, pmin(sqrt(logit_var), 2.0))
    ) %>%
    filter(is.finite(y_i), is.finite(psi_i)) %>%
    select(country, iso, year, round_id, district_id, y_i, psi_i)
}

direct_est_list <- map(pr_list, get_direct_estimates)
direct_est_all  <- bind_rows(direct_est_list)

direct_est_all %>% count(country, year, name = "observed_districts")

```

# Panel Construction {#sec:panel}

```{r build-panel}
build_panel <- function(country_iso, district_lookup, direct_est_all) {
  rounds <- direct_est_all %>%
    filter(country == country_iso) %>%
    distinct(year) %>% arrange(year) %>%
    mutate(round_id = row_number())
  
  est_clean <- direct_est_all %>%
    filter(country == country_iso) %>%
    select(district_id, year, y_i, psi_i)
  
  expand_grid(
    district_id = district_lookup$district_id,
    year        = rounds$year
  ) %>%
    left_join(district_lookup %>% select(district_id, district_name, district_code),
              by = "district_id") %>%
    left_join(est_clean, by = c("district_id", "year")) %>%
    left_join(rounds, by = "year") %>%
    arrange(year, district_id) %>%
    mutate(country = country_iso,
           observed = as.integer(!is.na(y_i)))
}

panels <- map(c("GHA", "BFA", "CIV"),
              ~build_panel(.x, district_lookups[[.x]], direct_est_all))
names(panels) <- c("GHA", "BFA", "CIV")

map_dfr(panels, function(p) {
  p %>% group_by(year) %>%
    summarise(total = n(), observed = sum(observed == 1),
              pct = round(100 * observed / total, 1), .groups = "drop")
}, .id = "country")
```

# Spatial Adjacency and BYM2 Scaling Factor {#sec:adjacency}

District adjacency is defined via queen contiguity using the `spdep`
package [@bivand2013]. Disconnected components are reconnected via a
snap tolerance to ensure a proper ICAR prior [@freni2018]. The BYM2
scaling factor $s$ [@riebler2016] makes the mixing parameter $\phi$
interpretable as the proportion of structured variance regardless of
graph topology.

```{r adjacency-scaling}
compute_scaling_factor <- function(nb, D) {
  adj_mat <- nb2mat(nb, style = "B", zero.policy = TRUE)
  Q_icar  <- Diagonal(x = rowSums(adj_mat)) - as(adj_mat, "sparseMatrix")
  eps <- 0.001
  Q_reg <- Q_icar + eps * Diagonal(D)
  Q_inv_diag <- diag(solve(Q_reg))
  exp(mean(log(Q_inv_diag)))
}

build_spatial <- function(iso, districts_ordered) {
  D <- nrow(districts_ordered)
  nb <- poly2nb(districts_ordered, queen = TRUE)
  if (sum(sapply(nb, length) == 0) > 0) {
    nb <- poly2nb(districts_ordered, queen = TRUE, snap = 0.1)
  }
  
  adj_mat <- nb2mat(nb, style = "B", zero.policy = TRUE)
  edges <- which(adj_mat == 1, arr.ind = TRUE) %>%
    as.data.frame() %>%
    filter(row < col)
  
  scaling_factor <- compute_scaling_factor(nb, D)
  
  list(D = D, nb = nb,
       node1 = edges$row, node2 = edges$col,
       N_edges = nrow(edges),
       scaling_factor = scaling_factor)
}

spatial <- map(c("GHA", "BFA", "CIV"),
               ~build_spatial(.x, districts_ordered[[.x]]))
names(spatial) <- c("GHA", "BFA", "CIV")

map_dfr(spatial, function(s) {
  tibble(N = s$D, N_edges = s$N_edges, scaling_factor = round(s$scaling_factor, 3))
}, .id = "country")
```

# Covariates: Harmonised Multi-Country Panel {#sec:covariates}

Seven covariates are assembled at the district-round level: elevation
(SRTM, population-weighted), rainfall (CHIRPS v2.0, Sep--Dec mean),
nighttime lights (NASA Black Marble VIIRS) [@elvidge2014], temperature
(CRU TS v4.09, Sep--Dec mean), travel time (Weiss/MAP 2020,
population-weighted), log population density (WorldPop 2020), and
multidimensional poverty (OPHI MPI / national census headcount). All
covariates are standardised within country. Pairwise correlations are
kept below 0.7 [@dormann2013].

```{r load-covariates}
covariates_panel <- read_csv("three_country_model_panel.csv") %>%
  select(country, district_id, district_name, time, year, label,
         elev_std, rainfall_std, ntl_std, temp_std,
         travel_std, pop_dens_std, poverty_std)

covariates_panel %>% count(country, time, name = "rows")
```

```{r join-covariates}
join_covariates <- function(iso) {
  panels[[iso]] %>%
    left_join(
      covariates_panel %>%
        filter(country == iso) %>%
        select(district_id, time,
               elev_std, rainfall_std, ntl_std, temp_std,
               travel_std, pop_dens_std, poverty_std),
      by = c("district_id", "round_id" = "time")
    )
}

panels <- list(
  GHA = join_covariates("GHA"),
  BFA = join_covariates("BFA"),
  CIV = join_covariates("CIV")
)

# Verify covariates are present
panels[["GHA"]] %>% select(district_id, year, ends_with("_std")) %>% head()
```

# Exploratory Data Analysis {#sec:eda}

## Correlation Heatmap

```{r correlation-heatmap, fig.height=7, fig.width=8}
covariate_cols <- c("elev_std", "rainfall_std", "ntl_std", "temp_std",
                    "travel_std", "pop_dens_std", "poverty_std")

cor_matrix <- covariates_panel %>%
  select(all_of(covariate_cols)) %>%
  cor(use = "complete.obs")

ggcorrplot(
  cor_matrix,
  type = "lower",
  lab = TRUE, lab_size = 3.5,
  colors = c("#2166ac", "white", "#b2182b"),
  title = "Pairwise correlations among standardised covariates",
  ggtheme = ggplot2::theme_minimal()
) +
  theme(plot.title = element_text(hjust = 0.5))
```

## Distribution of Direct Estimates

```{r direct-est-dist}
direct_est_all %>%
  mutate(prevalence = plogis(y_i)) %>%
  ggplot(aes(x = factor(year), y = prevalence)) +
  geom_violin(fill = "steelblue", alpha = 0.4) +
  geom_jitter(width = 0.1, alpha = 0.4, size = 0.5) +
  facet_wrap(~country, scales = "free_x") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = NULL, y = "Direct prevalence estimate (district)",
       title = "Distribution of district-level direct estimates by country and round",
       subtitle = "Each point is one district; violin shows the distribution") +
  theme_minimal(base_size = 11)
```

## Adjacency Graph (Ghana)

```{r adjacency-graph-plot, fig.height=8}
gha_with_id <- districts_raw[["GHA"]] %>%
  left_join(district_lookups[["GHA"]] %>% select(district_id, district_code),
            by = "district_code") %>%
  arrange(district_id)

centroid_coords <- suppressWarnings(st_coordinates(st_centroid(gha_with_id)))
centroids <- tibble(
  district_id = gha_with_id$district_id,
  x = centroid_coords[, "X"], y = centroid_coords[, "Y"]
)

sp <- spatial[["GHA"]]
edges_df <- tibble(from_id = sp$node1, to_id = sp$node2) %>%
  left_join(centroids %>% rename(x_from = x, y_from = y),
            by = c("from_id" = "district_id")) %>%
  left_join(centroids %>% rename(x_to = x, y_to = y),
            by = c("to_id" = "district_id"))

ggplot() +
  geom_sf(data = districts_raw[["GHA"]],
          fill = "grey95", colour = "grey60", linewidth = 0.2) +
  geom_segment(data = edges_df,
               aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
               colour = "#b2182b", alpha = 0.55, linewidth = 0.3) +
  geom_point(data = centroids, aes(x = x, y = y),
             colour = "black", size = 0.6) +
  labs(title = "District adjacency graph used for the ICAR prior, Ghana",
       subtitle = paste(sp$D, "districts,", sp$N_edges, "edges (queen contiguity)")) +
  theme_void(base_size = 11)
```

# Common Methodology {#sec:methodology}

All four models in this work share a common methodological core: a
Fay-Herriot area-level likelihood, a BYM2 spatial random effect, and
a shared set of priors. Each subsequent model section gives only the
specification elements unique to that model, then proceeds directly to
fitting, results and diagnostics.

## Fay-Herriot Likelihood

$$y_{it} \sim \mathcal{N}(\theta_{it}, \psi_{it}^2),$$
where $\theta_{it}$ is the latent logit prevalence and $\psi_{it}^2$
is treated as known (Section \ref{sec:direct}).

## BYM2 Spatial Random Effect

The district-level spatial random effect is given by [@riebler2016]:
$$b_i = \sigma_b \left(\sqrt{1-\phi}\, v_i + \sqrt{\phi/s}\, u_i\right),
\qquad v_i \sim \mathcal{N}(0,1), \quad u_i \sim \mathrm{ICAR}(W),$$
written in non-centred form to avoid funnel geometry in the posterior.
$W$ is the adjacency matrix of the relevant graph; $s$ is the BYM2
scaling factor; $\sigma_b$ is the marginal spatial SD; and
$\phi \in (0,1)$ is the proportion of structured variance.

## Shared Priors

| Parameter | Prior | Rationale |
|---|---|---|
| $\mu$ | $\mathcal{N}(-1.85, 1)$ | Centred near logit of moderate prevalence |
| $\sigma_b$ | $\mathrm{Exp}(4.6)$ | PC-style; $P(\sigma_b > 1) = 0.01$ [@simpson2017] |
| $\phi$ | $\mathrm{Beta}(3, 2)$ | Mild preference for structured > IID |
| $\beta_k$ | $\mathcal{N}(0, 1)$ | Weakly informative on standardised covariates |
| $\sigma_\gamma$ | $\mathrm{Exp}(2)$ | PC-style on interaction SD (Models 2--3) |
| $\tau_k$ | $\mathrm{Exp}(2)$ | PC-style on between-country SD (Models 3--4) |

## Computation

All models are fitted in Stan [@stan2024] via `cmdstanr`, using the
No-U-Turn Sampler with non-centred parameterisations throughout. Four
chains, 1500 warmup and 2000 sampling iterations each. Cross-sectional
and country-level space-time models use `adapt_delta = 0.99` and
`max_treedepth = 14`; pooled models use `adapt_delta = 0.97` and
`max_treedepth = 13`. Convergence is assessed via $\hat{R}$
[@vehtari2017], bulk/tail ESS, and trace/rank/pairs plots. Models are
compared by approximate leave-one-out cross-validation [@vehtari2017]
using the `loo` package.

## Stan Data Builders

```{r build-stan-data-cs}
build_stan_data_cs <- function(iso, panels, spatial, year, covar_cols = NULL) {
  panel <- panels[[iso]] %>% filter(year == !!year)
  sp <- spatial[[iso]]
  obs_idx <- which(panel$observed == 1)
  
  if (!is.null(covar_cols)) {
    available <- intersect(covar_cols, names(panel))
    X <- if (length(available) > 0) as.matrix(panel[, available]) else matrix(0, nrow(panel), 0)
    K <- ncol(X)
  } else {
    X <- matrix(0, nrow(panel), 0)
    K <- 0L
  }
  
  list(N = sp$D, N_obs = length(obs_idx),
       obs_idx = obs_idx,
       y = panel$y_i[obs_idx], psi = panel$psi_i[obs_idx],
       N_edges = sp$N_edges, node1 = sp$node1, node2 = sp$node2,
       scaling_factor = sp$scaling_factor,
       K = K, X = X)
}
```

```{r build-stan-data-st}
build_stan_data_st <- function(iso, panels, spatial, covar_cols = NULL) {
  panel <- panels[[iso]]
  sp <- spatial[[iso]]
  obs_idx <- which(panel$observed == 1)
  
  if (!is.null(covar_cols)) {
    available <- intersect(covar_cols, names(panel))
    X <- if (length(available) > 0) as.matrix(panel[, available]) else matrix(0, nrow(panel), 0)
    K <- ncol(X)
  } else {
    X <- matrix(0, nrow(panel), 0)
    K <- 0L
  }
  
  list(N = sp$D, T = 2L, NT = nrow(panel),
       N_obs = length(obs_idx), obs_idx = obs_idx,
       district = panel$district_id, time = panel$round_id,
       y = panel$y_i[obs_idx], psi = panel$psi_i[obs_idx],
       N_edges = sp$N_edges, node1 = sp$node1, node2 = sp$node2,
       scaling_factor = sp$scaling_factor,
       K = K, X = X)
}
```

```{r assemble-stan-data}
covar_cols <- c("elev_std", "rainfall_std", "ntl_std", "temp_std",
                "travel_std", "pop_dens_std", "poverty_std")

stan_data <- list()

# Cross-sectional reference: Ghana 2022 only
stan_data[["GHA_CS_2022"]] <- build_stan_data_cs(
  "GHA", panels, spatial, year = 2022, covar_cols = covar_cols
)

# Space-time models for all three countries
stan_data[["GHA_ST"]] <- build_stan_data_st("GHA", panels, spatial, covar_cols = covar_cols)
stan_data[["BFA_ST"]] <- build_stan_data_st("BFA", panels, spatial, covar_cols = covar_cols)
stan_data[["CIV_ST"]] <- build_stan_data_st("CIV", panels, spatial, covar_cols = covar_cols)

map_dfr(stan_data, function(d) {
  tibble(
    N = d$N,
    NT = ifelse(is.null(d$NT), d$N, d$NT),
    N_obs = d$N_obs, K = d$K,
    N_edges = d$N_edges,
    scaling_factor = round(d$scaling_factor, 3)
  )
}, .id = "model")
```

# Model 1: Cross-Sectional Ghana Baseline {#sec:model1}

## Specification

For each Ghana district $i$ observed in Round 2,
$$\theta_i = \mu + b_i + \mathbf{x}_i^\top \bm{\beta},$$
with $b_i$ the BYM2 random effect on the Ghana-only adjacency graph.
For comparison, we fit two alternatives that turn off one spatial
component each: a pure IID model ($\phi = 0$) and a pure ICAR model
($\phi = 1$).

## Fitting: IID

```{r fit-iid, cache=TRUE}
iid_code <- '
data {
  int<lower=1> N;
  int<lower=1> N_obs;
  array[N_obs] int<lower=1, upper=N> obs_idx;
  vector[N_obs] y;
  vector<lower=0>[N_obs] psi;
  int<lower=1> K;
  matrix[N, K] X;
}
parameters {
  real mu;
  vector[N] v;
  real<lower=0> sigma_b;
  vector[K] beta;
}
transformed parameters {
  vector[N] theta;
  theta = mu + sigma_b * v + X * beta;
}
model {
  mu ~ normal(-1.85, 1.0);
  sigma_b ~ exponential(4.6);
  beta ~ normal(0, 1);
  v ~ std_normal();
  y ~ normal(theta[obs_idx], psi);
}
generated quantities {
  vector[N_obs] log_lik;
  vector[N] prevalence;
  for (i in 1:N_obs)
    log_lik[i] = normal_lpdf(y[i] | theta[obs_idx[i]], psi[i]);
  for (i in 1:N)
    prevalence[i] = inv_logit(theta[i]);
}
'
writeLines(iid_code, "fh_iid.stan")
mod_iid <- cmdstan_model("fh_iid.stan")

sd_cs <- stan_data[["GHA_CS_2022"]]

fit_iid <- mod_iid$sample(
  data = sd_cs, chains = 4, parallel_chains = 4,
  iter_warmup = 1000, iter_sampling = 1500,
  adapt_delta = 0.95, max_treedepth = 12, refresh = 500
)
fit_iid$diagnostic_summary()

fit_iid <- ensure_fit(fit_iid, "m1_iid")
```

### IID Model: Full Posterior Summary

```{r m1-iid-summary}
# Full posterior summary of the IID-only cross-sectional model
tryCatch(print(fit_iid$summary(), n = Inf),
         error = function(e) cat("[IID summary failed:", conditionMessage(e), "]
"))
```

## Fitting: ICAR

```{r fit-icar, cache=TRUE}
icar_code <- '
data {
  int<lower=1> N;
  int<lower=1> N_obs;
  array[N_obs] int<lower=1, upper=N> obs_idx;
  vector[N_obs] y;
  vector<lower=0>[N_obs] psi;
  int<lower=1> N_edges;
  array[N_edges] int<lower=1, upper=N> node1;
  array[N_edges] int<lower=1, upper=N> node2;
  real<lower=0> scaling_factor;
  int<lower=1> K;
  matrix[N, K] X;
}
parameters {
  real mu;
  vector[N] u_raw;
  real<lower=0> sigma_b;
  vector[K] beta;
}
transformed parameters {
  vector[N] theta;
  vector[N] u = u_raw - mean(u_raw);
  theta = mu + sigma_b * u / sqrt(scaling_factor) + X * beta;
}
model {
  mu ~ normal(-1.85, 1.0);
  sigma_b ~ exponential(4.6);
  beta ~ normal(0, 1);
  target += -0.5 * dot_self(u_raw[node1] - u_raw[node2]);
  sum(u_raw) ~ normal(0, 0.001 * N);
  y ~ normal(theta[obs_idx], psi);
}
generated quantities {
  vector[N_obs] log_lik;
  vector[N] prevalence;
  for (i in 1:N_obs)
    log_lik[i] = normal_lpdf(y[i] | theta[obs_idx[i]], psi[i]);
  for (i in 1:N)
    prevalence[i] = inv_logit(theta[i]);
}
'
writeLines(icar_code, "fh_icar.stan")
mod_icar <- cmdstan_model("fh_icar.stan")

fit_icar <- mod_icar$sample(
  data = sd_cs, chains = 4, parallel_chains = 4,
  iter_warmup = 1000, iter_sampling = 1500,
  adapt_delta = 0.95, max_treedepth = 12, refresh = 500
)
fit_icar$diagnostic_summary()


fit_icar <- ensure_fit(fit_icar, "m1_icar")
```

### ICAR Model: Full Posterior Summary

<!-- ```{r m1-icar-summary} -->
<!-- # Full posterior summary of the ICAR-only cross-sectional model -->
<!-- tryCatch(print(fit_icar$summary(), n = Inf), -->
<!--          error = function(e) cat("[ICAR summary failed:", conditionMessage(e), "] -->
<!-- ")) -->
<!-- ``` -->

## Fitting: BYM2 (with log_lik for LOO)

```{r fit-cross-sectional}
mod_cs <- cmdstan_model("fh_bym2_cross_sectional.stan")

fit_gha_cs <- mod_cs$sample(
  data            = stan_data[["GHA_CS_2022"]],
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 1500,
  iter_sampling   = 2000,
  adapt_delta     = 0.99,
  max_treedepth   = 14,
  seed            = 123,
  refresh         = 500
)

fit_gha_cs$diagnostic_summary()


fit_gha_cs <- ensure_fit(fit_gha_cs, "m1_bym2_gha")

fit_gha_cs$print(c(
  "mu", "sigma_b", "phi", "p_national",
  "beta[1]", "beta[2]", "beta[3]", "beta[4]",
  "beta[5]", "beta[6]", "beta[7]"
))
```

### BYM2 Cross-Sectional Model: Full Posterior Summary

<!-- # ```{r m1-bym2-summary} -->
<!-- # # Full posterior summary of the BYM2 cross-sectional model  -->
<!-- # tryCatch(print(fit_gha_cs$summary(), n = Inf), -->
<!-- #          error = function(e) cat("[BYM2 summary failed:", conditionMessage(e), "] -->
<!-- # ")) -->
<!-- # ``` -->

```{r m1-bym2-keyparams}
# Compact view of the key BYM2 parameters
key_bym2_params <- c(
  "mu", "sigma_b", "phi", "p_national",
  paste0("beta[", 1:7, "]")
)
available <- fit_gha_cs$metadata()$stan_variables
base_names <- sub("\\[.*$", "", key_bym2_params)
keep <- key_bym2_params[base_names %in% available]
if (length(keep) > 0L) {
  tryCatch(fit_gha_cs$print(keep),
           error = function(e) cat("[key params failed:", conditionMessage(e), "]
"))
} else {
  cat("[no key parameters matched]
")
}
```

## LOO Model Comparison

```{r loo-cv, fig.height=5, fig.width=9}
library(loo)


fit_iid     <- ensure_fit(fit_iid,    "m1_iid")
fit_icar    <- ensure_fit(fit_icar,   "m1_icar")
fit_gha_cs  <- ensure_fit(fit_gha_cs, "m1_bym2_gha")


safe_loo <- function(fit, name) {
  if (!("log_lik" %in% fit$metadata()$stan_variables)) {
    message(name, ": no log_lik variable; skipping LOO.")
    return(NULL)
  }
  tryCatch(
    fit$loo(variables = "log_lik"),
    error = function(e) {
      message(name, " LOO failed: ", conditionMessage(e))
      NULL
    }
  )
}

loo_iid  <- safe_loo(fit_iid,    "IID")
loo_icar <- safe_loo(fit_icar,   "ICAR")
loo_bym2 <- safe_loo(fit_gha_cs, "BYM2")


if (is.null(loo_bym2)) {
  message("Refitting BYM2 with log_lik for LOO comparison...")
  bym2_loo_code <- "
data {
  int<lower=1> N;
  int<lower=1> N_obs;
  array[N_obs] int<lower=1, upper=N> obs_idx;
  vector[N_obs] y;
  vector<lower=0>[N_obs] psi;
  int<lower=1> N_edges;
  array[N_edges] int<lower=1, upper=N> node1;
  array[N_edges] int<lower=1, upper=N> node2;
  real<lower=0> scaling_factor;
  int<lower=1> K;
  matrix[N, K] X;
}
parameters {
  real mu;
  vector[N] v;
  vector[N] u_raw;
  real<lower=0> sigma_b;
  real<lower=0, upper=1> phi;
  vector[K] beta;
}
transformed parameters {
  vector[N] theta;
  vector[N] u = u_raw - mean(u_raw);
  vector[N] b = sigma_b * (sqrt(1 - phi) * v + sqrt(phi / scaling_factor) * u);
  theta = mu + b + X * beta;
}
model {
  mu ~ normal(-1.85, 1.0);
  sigma_b ~ exponential(4.6);
  phi ~ beta(3, 2);
  beta ~ normal(0, 1);
  v ~ std_normal();
  target += -0.5 * dot_self(u_raw[node1] - u_raw[node2]);
  sum(u_raw) ~ normal(0, 0.001 * N);
  y ~ normal(theta[obs_idx], psi);
}
generated quantities {
  vector[N_obs] log_lik;
  vector[N] prevalence;
  for (i in 1:N_obs)
    log_lik[i] = normal_lpdf(y[i] | theta[obs_idx[i]], psi[i]);
  for (i in 1:N)
    prevalence[i] = inv_logit(theta[i]);
}
"
  writeLines(bym2_loo_code, "fh_bym2_loo.stan")
  mod_bym2_loo <- cmdstan_model("fh_bym2_loo.stan")
  fit_bym2_loo <- mod_bym2_loo$sample(
    data = sd_cs, chains = 4, parallel_chains = 4,
    iter_warmup = 1000, iter_sampling = 1500,
    adapt_delta = 0.95, max_treedepth = 12, refresh = 500
  )
  fit_bym2_loo <- ensure_fit(fit_bym2_loo, "m1_bym2_loo")
  loo_bym2 <- safe_loo(fit_bym2_loo, "BYM2 (refit)")
}

loo_list <- list(IID = loo_iid, ICAR = loo_icar, BYM2 = loo_bym2)
loo_list <- Filter(Negate(is.null), loo_list)

if (length(loo_list) >= 2L) {
  loo_comp <- loo_compare(loo_list)
  print(loo_comp, simplify = FALSE)

  loo_df <- as_tibble(loo_comp, rownames = "Model")

  p_loo <- loo_df %>%
    mutate(Model = factor(Model, levels = rev(Model))) %>%
    ggplot(aes(x = Model, y = elpd_diff)) +
    geom_col(aes(fill = Model), width = 0.6, alpha = 0.8) +
    geom_errorbar(aes(ymin = elpd_diff - 2 * se_diff,
                       ymax = elpd_diff + 2 * se_diff), width = 0.2) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
    scale_fill_manual(values = c("BYM2" = "#a80050", "ICAR" = "#2563eb", "IID" = "#f5a623")) +
    coord_flip() +
    labs(title    = "LOO-CV: IID vs ICAR vs BYM2 (Ghana 2022)",
         subtitle = "Delta ELPD relative to best model (higher = better)",
         x = NULL, y = "Delta ELPD") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none")

  ggsave("output/figures/loo_cv_comparison.png", p_loo,
         width = 9, height = 5, dpi = 300, bg = "white")
  p_loo
} else {
  message("Fewer than two LOO objects available; skipping comparison plot.")
}
```

## Maps: Direct vs IID vs ICAR vs BYM2

```{r model-comparison-maps, fig.height=6, fig.width=14}
# Extract prevalence from each model
get_prev <- function(fit) {
  s <- fit$summary("prevalence")
  s$median
}

prev_iid_v  <- get_prev(fit_iid)
prev_icar_v <- get_prev(fit_icar)


if ("prevalence" %in% fit_gha_cs$metadata()$stan_variables) {
  prev_bym2_v <- get_prev(fit_gha_cs)
} else {
  prev_bym2_v <- get_prev(fit_bym2_loo)
}

D_gha_cs <- stan_data[["GHA_CS_2022"]]$N

model_prev <- bind_rows(
  tibble(district_id = 1:D_gha_cs, prev = prev_iid_v[1:D_gha_cs], model = "IID"),
  tibble(district_id = 1:D_gha_cs, prev = prev_icar_v[1:D_gha_cs], model = "ICAR"),
  tibble(district_id = 1:D_gha_cs, prev = prev_bym2_v[1:D_gha_cs], model = "BYM2"),
  panels[["GHA"]] %>% filter(round_id == 2) %>%
    mutate(prev = ifelse(observed == 1, plogis(y_i), NA_real_), model = "Direct") %>%
    select(district_id, prev, model)
) %>%
  mutate(model = factor(model, levels = c("Direct", "IID", "ICAR", "BYM2")))

# Join to shapefile via district_id
gha_shp <- districts_raw[["GHA"]] %>%
  mutate(district_id = row_number())

maps_comp <- map_dfr(levels(model_prev$model), function(m) {
  gha_shp %>%
    left_join(model_prev %>% filter(model == m), by = "district_id") %>%
    mutate(model = m)
}) %>%
  mutate(model = factor(model, levels = c("Direct", "IID", "ICAR", "BYM2")))

p_comp <- ggplot(maps_comp) +
  geom_sf(aes(fill = prev), colour = "grey50", linewidth = 0.1) +
  scale_fill_viridis_c(name = "Prevalence",
                       labels = scales::percent_format(accuracy = 1),
                       option = "magma", direction = -1,
                       na.value = "grey80") +
  facet_wrap(~model, nrow = 1) +
  labs(title = "Model comparison: Direct vs IID vs ICAR vs BYM2 (Ghana 2022)",
       subtitle = "Grey = unobserved in direct estimates. Spatial models fill all 260 districts.") +
  theme_void(base_size = 11) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 12, face = "bold"))

ggsave("output/figures/model_comparison_maps_ghana.png", p_comp, width = 14, height = 6, dpi = 300, bg = "white")
p_comp
```

## Posterior Predictive Checks

```{r ppc-cs}
y_obs_cs <- stan_data[["GHA_CS_2022"]]$y
y_rep_cs <- fit_gha_cs$draws("y_rep", format = "matrix")

ppc_dens_overlay(y = y_obs_cs, yrep = y_rep_cs[1:100, ]) +
  labs(title = "PPC: Density overlay (Ghana 2022 cross-sectional)")

ppc_stat(y = y_obs_cs, yrep = y_rep_cs, stat = "mean") +
  labs(title = "PPC: Mean of y")

ppc_stat(y = y_obs_cs, yrep = y_rep_cs, stat = "sd") +
  labs(title = "PPC: Standard deviation of y")
```

## Convergence Diagnostics

```{r cs-trace}
mcmc_trace(fit_gha_cs$draws(c("mu", "sigma_b", "phi")))
```

# Model 2: Country-Level Space-Time Models {#sec:model2}

## Specification

For each country $c$ fitted separately,
$$\theta_{it} = \mu^{(c)} + b_i^{(c)} + \delta_t^{(c)} + \gamma_{it}^{(c)}
              + \mathbf{x}_{it}^\top \bm{\beta}^{(c)},$$
with $b_i^{(c)}$ the BYM2 effect on the country-specific graph and
$\gamma_{it}^{(c)} \sim \mathcal{N}(0, \sigma_\gamma^2)$ the Knorr-Held
Type I [@knorrheld2000] interaction. With $T=2$, Type I is the honest
choice: structured temporal or space-time interactions are not
identifiable [@goicoa2018].

## Fitting: All Three Countries

```{r fit-spacetime-all}
mod_st <- cmdstan_model("fh_bym2_typeI.stan")

fit_st <- list()

for (iso in c("GHA", "BFA", "CIV")) {
  cat("\n========== Fitting space-time model for", iso, "==========\n")
  
  fit_st[[iso]] <- mod_st$sample(
    data            = stan_data[[paste0(iso, "_ST")]],
    chains          = 4,
    parallel_chains = 4,
    iter_warmup     = 1500,
    iter_sampling   = 2000,
    adapt_delta     = 0.99,
    max_treedepth   = 14,
    seed            = 1729,
    refresh         = 500
  )
  
  cat("\nDiagnostics for", iso, ":\n")
  print(fit_st[[iso]]$diagnostic_summary())
  # Persist + restore draws so the fit survives across knit sessions
  fit_st[[iso]] <- ensure_fit(fit_st[[iso]], paste0("m2_st_", iso))
}
```

## Country-Level Space-Time: Full Posterior Summaries

```{r m2-summaries}

safe_full_summary <- function(fit, label) {
  cat("
========== ", label, " ==========
", sep = "")
  tryCatch(
    {
      print(fit$summary(), n = Inf)
    },
    error = function(e) {
      cat("  [could not print full summary:", conditionMessage(e), "]
")
    }
  )
}


safe_print <- function(fit, params, label = NULL) {
  if (!is.null(label)) cat("
--- ", label, " ---
", sep = "")
  available <- fit$metadata()$stan_variables
  base_names <- sub("\\[.*$", "", params)
  keep <- params[base_names %in% available]
  if (length(keep) == 0L) {
    cat("  [none of the requested parameters exist in this fit]
")
    return(invisible(NULL))
  }
  tryCatch(fit$print(keep),
           error = function(e) cat("  [print failed:", conditionMessage(e), "]
"))
}

for (cc in names(fit_st)) {
  safe_full_summary(fit_st[[cc]], paste(cc, "SPACE-TIME"))
}
```

```{r m2-key-params}
# Key parameters across the three country-level space-time models
key_st_params <- c(
  "mu", "sigma_b", "phi", "sigma_gamma",
  "delta2", "delta[2]",                       
  "p_national[1]", "p_national[2]",
  paste0("beta[", 1:7, "]")
)

for (cc in names(fit_st)) {
  safe_print(fit_st[[cc]], key_st_params, label = cc)
}
```

## Summary Across Countries

```{r summary-across-countries}
summary_all <- map_dfr(c("GHA", "BFA", "CIV"), function(iso) {
  draws <- fit_st[[iso]]$draws(
    variables = c("mu", "delta_2", "sigma_b", "phi", "sigma_gamma",
                  "p_round1", "p_round2"),
    format = "draws_matrix"
  )
  
  tibble(
    country  = iso,
    variable = colnames(draws),
    mean     = colMeans(draws),
    median   = apply(draws, 2, median),
    lower    = apply(draws, 2, quantile, 0.025),
    upper    = apply(draws, 2, quantile, 0.975),
    rhat     = apply(draws, 2, posterior::rhat),
    ess_bulk = apply(draws, 2, posterior::ess_bulk)
  )
})

print(n = 25, summary_all)

# fit_st$print(c(
#   "mu", "sigma_b", "phi", "p_national",
#   "beta[1]", "beta[2]", "beta[3]", "beta[4]",
#   "beta[5]", "beta[6]", "beta[7]"
# ))
```

## Prior vs Posterior Comparison

```{r prior-post}
post_draws <- fit_st[["GHA"]]$draws(
  c("mu", "delta_2", "sigma_b", "phi", "sigma_gamma"),
  format = "draws_df"
)

prior_post_plot <- function(post_var, prior_fun, x_range, label) {
  ggplot() +
    geom_function(fun = prior_fun, xlim = x_range,
                  aes(colour = "Prior"), linewidth = 0.9) +
    geom_density(data = post_draws,
                 aes(x = .data[[post_var]], colour = "Posterior"),
                 linewidth = 0.9) +
    scale_colour_manual(values = c(Prior = "grey50", Posterior = "#b2182b")) +
    labs(x = label, y = "Density", colour = NULL) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "none")
}

p_mu     <- prior_post_plot("mu",          ~ dnorm(.x, -1.85, 1), c(-4, 0),  expression(mu))
p_delta  <- prior_post_plot("delta_2",     ~ dnorm(.x, 0, 5),     c(-3, 3),  expression(delta[2]))
p_sigb   <- prior_post_plot("sigma_b",     ~ dexp(.x, 4.6),       c(0, 2),   expression(sigma[b]))
p_phi    <- prior_post_plot("phi",         ~ dbeta(.x, 3, 2),     c(0, 1),   expression(phi))
p_siggam <- prior_post_plot("sigma_gamma", ~ dexp(.x, 2),         c(0, 2),   expression(sigma[gamma]))

(p_mu | p_delta | p_sigb) / (p_phi | p_siggam | plot_spacer()) +
  plot_annotation(title = "Prior vs posterior, space-time BYM2 model (Ghana)")
```

## Covariate Effects Across Countries (Country-Specific Models)

```{r covariate-forest}
covar_labels <- c("Elevation", "Rainfall", "Nighttime lights", "Temperature",
                  "Travel time", "Pop. density (log)", "Poverty")


beta_all <- summary_all %>%
  # Filter only the rows where the variable starts with "beta["
  filter(str_detect(variable, "^beta\\[")) %>%
  mutate(
    # Extract the numeric index from "beta[1]", "beta[2]", etc.
    beta_idx = as.integer(str_extract(variable, "\\d+")),
    
    # Map the index to the correct human-readable label
    covariate = covar_labels[beta_idx],
    
    # Calculate significance strictly using the lower/upper bounds in the table
    # If both bounds are greater than 0, or both are less than 0, it is significant
    is_significant = (lower > 0) | (upper < 0),
    sig_marker = ifelse(is_significant, "*", "")
  )

# 3. Factor the covariates and countries to preserve plotting order
beta_all$covariate <- factor(beta_all$covariate, levels = rev(covar_labels))
beta_all$country <- factor(beta_all$country, levels = c("GHA", "BFA", "CIV"))

# 4. Generate the plot
p_covars <- ggplot(beta_all, aes(x = mean, y = covariate, colour = country, group = country)) +
  
  geom_pointrange(aes(xmin = lower, xmax = upper),
                  position = position_dodge(width = 0.6), size = 0.8) +
                  
  # FIXED ALIGNMENT: Added hjust = 0 and locked the dodge with group = country
  geom_text(aes(x = upper + 0.02, label = sig_marker),
            position = position_dodge(width = 0.6),
            size = 7,          # Slightly larger for visibility on a slide
            vjust = 0.75,      # Pulls the top-heavy asterisk down to center it with the line
            hjust = 0,         # Left-aligns it exactly 0.02 units away from the line
            show.legend = FALSE, 
            colour = "black") +
            
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40", linewidth = 0.8) +
  scale_colour_brewer(palette = "Dark2", name = "Country") +
  
  scale_x_continuous(expand = expansion(mult = c(0.05, 0.15))) + 
  
  labs(x = "Coefficient (logit scale)", y = NULL,
       title = "Covariate effects: Individual Space-Time Models",
       subtitle = "Posterior mean and credible intervals. Asterisks (*) denote significant effects.") +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

ggsave("output/figures/individual_covariate_effects.png", p_covars, width = 10, height = 7, dpi = 300, bg = "white")
p_covars
```

## Prevalence Maps: All Countries, Both Rounds

```{r prevalence-maps, fig.height=12, fig.width=12}
make_prev_map_df <- function(iso, round) {
  prev_sum <- fit_st[[iso]]$summary("prevalence",
    median = ~ median(.x),
    lower  = ~ quantile(.x, 0.025),
    upper  = ~ quantile(.x, 0.975)
  ) %>%
    mutate(row_id = as.integer(gsub("prevalence\\[|\\]", "", variable))) %>%
    left_join(
      panels[[iso]] %>% mutate(row_id = row_number()) %>%
        select(row_id, district_id, round_id),
      by = "row_id"
    ) %>%
    filter(round_id == round) %>%
    left_join(district_lookups[[iso]] %>% select(district_id, district_code),
              by = "district_id")
  
  districts_raw[[iso]] %>%
    left_join(prev_sum, by = "district_code") %>%
    mutate(country = iso, round_label = paste("Round", round))
}

all_maps <- map_dfr(c("GHA", "BFA", "CIV"),
                    function(iso) bind_rows(
                      make_prev_map_df(iso, 1),
                      make_prev_map_df(iso, 2)
                    ))

ggplot(all_maps) +
  geom_sf(aes(fill = median), colour = "grey40", linewidth = 0.1) +
  scale_fill_viridis_c(name = "Prevalence",
                       labels = scales::percent_format(accuracy = 1),
                       option = "magma", direction = -1) +
  facet_grid(country ~ round_label) +
  labs(title = "Under-5 malaria prevalence across three West African countries",
       subtitle = "Posterior medians from BYM2 space-time model") +
  theme_void(base_size = 11) +
  theme(legend.position = "right")
```

## District Trend Deviations

```{r trend-deviation-map}
dev_summary_gha <- fit_st[["GHA"]]$summary("gamma_star",
  median = ~ median(.x),
  lower  = ~ unname(quantile(.x, 0.025)),
  upper  = ~ unname(quantile(.x, 0.975))
) %>%
  mutate(
    district_id = as.integer(gsub("gamma_star\\[|\\]", "", variable)),
    confident = (lower > 0) | (upper < 0)
  ) %>%
  left_join(district_lookups[["GHA"]] %>%
              select(district_id, district_code, district_name),
            by = "district_id")

map_dev <- districts_raw[["GHA"]] %>%
  left_join(dev_summary_gha, by = "district_code")

dev_limit <- max(abs(range(map_dev$median, na.rm = TRUE)))

ggplot(map_dev) +
  geom_sf(aes(fill = median), colour = "grey40", linewidth = 0.1) +
  scale_fill_gradient2(
    name = expression(gamma[i]^"*"),
    low = "#2166ac", mid = "white", high = "#b2182b",
    midpoint = 0, limits = c(-dev_limit, dev_limit)
  ) +
  geom_sf(data = map_dev %>% filter(confident),
          fill = NA, colour = "black", linewidth = 0.4) +
  labs(
    title = "District-level deviation from national malaria trend, Ghana 2014–2022",
    subtitle = "Blue: faster decline than national. Red: slower / rising. Bold outline: 95% CI excludes 0."
  ) +
  theme_void(base_size = 11)
```

## Top Trend-Deviating Districts

```{r top-deviation-districts}
top_neg <- dev_summary_gha %>% arrange(median) %>% head(10) %>%
  mutate(direction = "Faster decline")
top_pos <- dev_summary_gha %>% arrange(desc(median)) %>% head(10) %>%
  mutate(direction = "Slower / rising")

bind_rows(top_neg, top_pos) %>%
  mutate(district_name = fct_reorder(district_name, median)) %>%
  ggplot(aes(x = median, y = district_name, colour = direction)) +
  geom_pointrange(aes(xmin = lower, xmax = upper)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_colour_manual(values = c("Faster decline" = "#2166ac",
                                 "Slower / rising" = "#b2182b")) +
  labs(x = expression(gamma[i]^"*" * " (logit scale)"), y = NULL,
       title = "Top districts with credible deviation from national trend, Ghana",
       colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

## Uncertainty Maps

```{r uncertainty-maps, fig.height=5}
make_uncertainty_df <- function(iso, round) {
  unc <- fit_st[[iso]]$summary("prevalence",
    width = ~ unname(quantile(.x, 0.975) - quantile(.x, 0.025))
  ) %>%
    mutate(row_id = as.integer(gsub("prevalence\\[|\\]", "", variable))) %>%
    left_join(
      panels[[iso]] %>% mutate(row_id = row_number()) %>%
        select(row_id, district_id, round_id, observed),
      by = "row_id"
    ) %>%
    filter(round_id == round) %>%
    left_join(district_lookups[[iso]] %>% select(district_id, district_code),
              by = "district_id")
  
  districts_raw[[iso]] %>%
    left_join(unc, by = "district_code") %>%
    mutate(country = iso)
}

unc_r2 <- map_dfr(c("GHA", "BFA", "CIV"),
                  function(iso) make_uncertainty_df(iso, 2))

ggplot(unc_r2) +
  geom_sf(aes(fill = width), colour = "grey40", linewidth = 0.1) +
  scale_fill_viridis_c(name = "95% CI width",
                       labels = scales::percent_format(accuracy = 1),
                       option = "cividis") +
  facet_wrap(~country) +
  labs(title = "Posterior uncertainty in Round 2 prevalence estimates",
       subtitle = "Wider intervals = greater reliance on spatial smoothing (no direct data)") +
  theme_void(base_size = 11)
```

## Observed vs Unobserved Districts

```{r observed-unobserved-map, fig.height=10, fig.width=10}
make_obs_map <- function(iso, round) {
  panel_round <- panels[[iso]] %>% filter(round_id == round)
  
  obs_by_district <- panel_round %>%
    select(district_id, observed) %>%
    mutate(status = ifelse(observed == 1, "Observed", "Unobserved")) %>%
    left_join(district_lookups[[iso]] %>% select(district_id, district_code),
              by = "district_id")
  
  districts_raw[[iso]] %>%
    left_join(obs_by_district, by = "district_code") %>%
    mutate(country = iso,
           round_label = paste("Round", round))
}

obs_maps <- map_dfr(c("GHA", "BFA", "CIV"),
                    function(iso) bind_rows(
                      make_obs_map(iso, 1),
                      make_obs_map(iso, 2)
                    ))

p_obs <- ggplot(obs_maps) +
  geom_sf(aes(fill = status), colour = "grey50", linewidth = 0.1) +
  scale_fill_manual(
    values = c("Observed" = "#2563eb", "Unobserved" = "#fbbf24"),
    name = "Survey data"
  ) +
  facet_grid(country ~ round_label) +
  labs(
    title = "Districts with direct DHS data vs unobserved districts",
    subtitle = "Unobserved districts rely entirely on spatial borrowing from neighbours and covariates"
  ) +
  theme_void(base_size = 11) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 11, face = "bold"))

ggsave("output/figures/observed_vs_unobserved_map.png", p_obs, width = 10, height = 10, dpi = 300, bg = "white")
p_obs
```

## Credible Interval Comparison: Observed vs Unobserved

```{r ci-comparison-obs-unobs, fig.height=6, fig.width=10}
make_ci_comparison <- function(iso, round) {
  D <- stan_data[[paste0(iso, "_ST")]]$N
  
  # Use default summary (returns q5, q95, mean, median, sd, etc.)
  prev_summary <- fit_st[[iso]]$summary("prevalence") %>%
    mutate(row_id = as.integer(gsub("prevalence\\[|\\]", "", variable))) %>%
    left_join(
      panels[[iso]] %>% mutate(row_id = row_number()) %>%
        select(row_id, district_id, round_id, observed),
      by = "row_id"
    ) %>%
    filter(round_id == round) %>%
    mutate(
      ci_width = q95 - q5,
      status = ifelse(observed == 1, "Observed", "Unobserved"),
      country = iso
    ) %>%
    left_join(district_lookups[[iso]] %>% select(district_id, district_name),
              by = "district_id")
  
  prev_summary
}

ci_all <- map_dfr(c("GHA", "BFA", "CIV"),
                  function(iso) make_ci_comparison(iso, 2))

# Boxplot comparing CI widths
p_ci_box <- ggplot(ci_all, aes(x = status, y = ci_width, fill = status)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8) +
  facet_wrap(~country, scales = "free_y") +
  scale_fill_manual(values = c("Observed" = "#2563eb", "Unobserved" = "#fbbf24")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Width of 90% credible intervals: observed vs unobserved districts",
    subtitle = "Round 2 estimates. Unobserved districts have wider intervals (more uncertainty)",
    x = NULL, y = "Width of 90% CrI",
    fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

ggsave("output/figures/ci_comparison_obs_unobs.png", p_ci_box, width = 10, height = 6, dpi = 300, bg = "white")
p_ci_box
```

## Caterpillar Plot: Sorted Prevalence with Credible Intervals

```{r caterpillar-plot, fig.height=7, fig.width=10}
ci_gha <- ci_all %>%
  filter(country == "GHA") %>%
  arrange(median) %>%
  mutate(rank = row_number())

p_cat <- ggplot(ci_gha, aes(x = rank, y = median, colour = status)) +
  geom_linerange(aes(ymin = q5, ymax = q95), alpha = 0.5, linewidth = 0.3) +
  geom_point(size = 0.6) +
  scale_colour_manual(values = c("Observed" = "#2563eb", "Unobserved" = "#fbbf24")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Caterpillar plot: district prevalence estimates, Ghana Round 2",
    subtitle = "260 districts sorted by posterior median. Gold = unobserved (wider CrI).",
    x = "District (sorted)", y = "Prevalence (posterior median and 90% CrI)",
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

ggsave("output/figures/caterpillar_plot_ghana_r2.png", p_cat, width = 10, height = 7, dpi = 300, bg = "white")
p_cat
```

## Direct vs Model-Smoothed Estimates

```{r direct-vs-smoothed, fig.height=7, fig.width=9}
gha_panel_r2 <- panels[["GHA"]] %>% filter(round_id == 2)
gha_direct <- gha_panel_r2 %>%
  filter(observed == 1) %>%
  mutate(direct_prev = plogis(y_i))

gha_model <- fit_st[["GHA"]]$summary("prevalence") %>%
  mutate(row_id = as.integer(gsub("prevalence\\[|\\]", "", variable))) %>%
  left_join(
    panels[["GHA"]] %>% mutate(row_id = row_number()) %>%
      select(row_id, district_id, round_id),
    by = "row_id"
  ) %>%
  filter(round_id == 2)

shrinkage_df <- gha_direct %>%
  select(district_id, direct_prev) %>%
  left_join(gha_model %>% select(district_id, model_prev = median,
                                   model_lower = q5, model_upper = q95),
            by = "district_id") %>%
  left_join(district_lookups[["GHA"]] %>% select(district_id, district_name),
            by = "district_id")

p_shrink <- ggplot(shrinkage_df, aes(x = direct_prev, y = model_prev)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_linerange(aes(ymin = model_lower, ymax = model_upper),
                 alpha = 0.15, colour = "#a80050") +
  geom_point(alpha = 0.6, colour = "#a80050", size = 1.5) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "BYM2 shrinkage: direct estimates vs model-smoothed posteriors",
    subtitle = "Ghana Round 2. Points below diagonal = model pulled estimate down. Bars = 90% CrI.",
    x = "Direct estimate (design-based)",
    y = "Model posterior median"
  ) +
  theme_minimal(base_size = 12)

ggsave("output/figures/direct_vs_smoothed_ghana_r2.png", p_shrink, width = 9, height = 7, dpi = 300, bg = "white")
p_shrink
```

## Space-Time Convergence Diagnostics

```{r st-diagnostics}

mcmc_rank_overlay(fit_st[["GHA"]]$draws(
  c("mu", "delta_2", "sigma_b", "phi", "sigma_gamma")
))

mcmc_pairs(
  fit_st[["GHA"]]$draws(c("sigma_b", "phi", "sigma_gamma", "delta_2")),
  off_diag_args = list(size = 0.5, alpha = 0.3)
)

```

# Model 3: Pooled Three-Country Space-Time Model {#sec:model3}

## Motivation

The country-specific models in Section \ref{sec:model2} treat each
country as a spatial island, despite shared borders, shared *Anopheles*
vectors and shared West African monsoon dynamics. The pooled model
relaxes this by smoothing across the combined 724-district adjacency
graph (including cross-border edges following @saldarriaga2020) and by
partially pooling country-specific covariate slopes through a
hierarchical prior $\beta_{ck} \sim \mathcal{N}(\bar{\beta}_k, \tau_k^2)$.

## Construction

```{r}
gha_sf <- districts_raw[["GHA"]] %>%
  mutate(country = "GHA", country_id = 1L,
         local_id = row_number())

civ_sf <- districts_raw[["CIV"]] %>%
  mutate(country = "CIV", country_id = 2L,
         local_id = row_number())

bfa_sf <- districts_raw[["BFA"]] %>%
  mutate(country = "BFA", country_id = 3L,
         local_id = row_number())

D_gha <- nrow(gha_sf)  # 260
D_civ <- nrow(civ_sf)  # 113
D_bfa <- nrow(bfa_sf)  # 351
D_all <- D_gha + D_civ + D_bfa  # 724


#assigning glboal IDs
gha_sf$global_id <- 1:D_gha
civ_sf$global_id <- (D_gha + 1):(D_gha + D_civ)
bfa_sf$global_id <- (D_gha + D_civ + 1):D_all

# binding them all together
all_sf <- bind_rows(gha_sf, civ_sf, bfa_sf)

# constracting the queens contiguity on the combined spatial shapefile
nb_all <- poly2nb(all_sf)

# counting cross border edges 
cross_border <- 0
within_country <- 0
for (i in seq_along(nb_all)) {
  for (j in nb_all[[i]]) {
    if (j > i) {  # avoid double counting
      if (all_sf$country[i] != all_sf$country[j]) {
        cross_border <- cross_border + 1
      } else {
        within_country <- within_country + 1
      }
    }
  }
}

cross_border

# Extracting edge list for Stan
edges <- nb_to_edgelist <- function(nb) {
  n1 <- c(); n2 <- c()
  for (i in seq_along(nb)) {
    for (j in nb[[i]]) {
      if (j > i) {
        n1 <- c(n1, i)
        n2 <- c(n2, j)
      }
    }
  }
  list(node1 = n1, node2 = n2, N_edges = length(n1))
}

el <- nb_to_edgelist(nb_all)
cat("    Edges extracted:", el$N_edges, "\n")

# Computing the BYM2 scaling factor based on the combined spatial dataset
Q <- -nb2mat(nb_all, style = "B", zero.policy = TRUE)
diag(Q) <- -rowSums(Q)
Q_scaled <- INLA::inla.scale.model(Q, constr = list(A = matrix(1, 1, D_all), e = 0))
scaling_factor_all <- exp(mean(log(diag(Q_scaled))))


# Country index for each district (1=GHA, 2=CIV, 3=BFA)
country_vec <- all_sf$country_id  # length 724

# Re-index panels with global district IDs
# Each country's panel has: district_id (local 1:D), round_id (1 or 2),
# y_i, psi_i, observed, and covariates

remap_panel <- function(iso, offset) {
  panels[[iso]] %>%
    mutate(
      global_district_id = district_id + offset,
      country = iso
    )
}

panel_pooled <- bind_rows(
  remap_panel("GHA", 0),
  remap_panel("CIV", D_gha),
  remap_panel("BFA", D_gha + D_civ)
) %>%
  arrange(global_district_id, round_id)

NT <- nrow(panel_pooled)
cat("  Panel rows (NT):", NT, "(should be", D_all * 2, ")\n")

# Observation indices
obs_mask <- panel_pooled$observed == 1
obs_idx  <- which(obs_mask)
N_obs    <- sum(obs_mask)

cat("  Observed:", N_obs, "/ Total:", NT, "\n")

# Build X matrix from model panel (already standardised within country)
model_panel <- read_csv("output/three_country_model_panel.csv",
                         show_col_types = FALSE)

# Join covariates in the same order as panel_pooled
cov_names <- c("elev_std", "rainfall_std", "ntl_std", "temp_std",
               "travel_std", "pop_dens_std", "poverty_std")
K <- length(cov_names)

X_pooled <- panel_pooled %>%
  select(-any_of(cov_names)) %>% 
  left_join(model_panel %>% select(country, district_id, time, all_of(cov_names)),
            by = c("country", 
                   "district_id" = "district_id",
                   "round_id" = "time")) %>%
  select(all_of(cov_names)) %>%
  as.matrix()


cat("  X matrix:", nrow(X_pooled), "x", ncol(X_pooled), "\n")
cat("  NAs in X:", sum(is.na(X_pooled)), "\n")

# Assemble Stan data
stan_data_pooled <- list(
  N       = D_all,
  T       = 2L,
  NT      = NT,
  C       = 3L,
  K       = K,
  N_obs   = N_obs,
  obs_idx = obs_idx,
  y       = panel_pooled$y_i[obs_idx],
  psi     = panel_pooled$psi_i[obs_idx],
  district = panel_pooled$global_district_id,
  time     = panel_pooled$round_id,
  country  = country_vec,
  X        = X_pooled,
  N_edges  = el$N_edges,
  node1    = el$node1,
  node2    = el$node2,
  scaling_factor = scaling_factor_all
)

cat("\n  Stan data assembled.\n")
cat("    N =", stan_data_pooled$N, "\n")
cat("    NT =", stan_data_pooled$NT, "\n")
cat("    C =", stan_data_pooled$C, "\n")
cat("    K =", stan_data_pooled$K, "\n")
cat("    N_obs =", stan_data_pooled$N_obs, "\n")
cat("    N_edges =", stan_data_pooled$N_edges, "\n")


###############################################################################
# STEP 3: Compile and fit
###############################################################################

cat("\n", strrep("=", 70), "\n")
cat("  STEP 3: Fitting pooled model\n")
cat(strrep("=", 70), "\n\n")

mod_pooled <- cmdstan_model("fh_bym2_pooled_3country.stan")

mod_pooled$check_syntax(pedantic = TRUE)

fit_pooled <- mod_pooled$sample(
  data            = stan_data_pooled,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 1500,
  iter_sampling   = 2000,
  adapt_delta     = 0.97,
  max_treedepth   = 13,
  refresh         = 200
)


fit_pooled <- ensure_fit(fit_pooled, "m3_pooled")


```


## Covariate Effects Across Countries (Pooled Model)

```{r}
covar_labels <- c("Elevation", "Rainfall", "Nighttime lights", "Temperature",
                  "Travel time", "Pop. density (log)", "Poverty")


country_names <- c("GHA", "CIV", "BFA") 

# 1. Extract the country-specific slopes (using 0.05 and 0.95 for a 90% CI)
beta_pooled <- fit_pooled$summary("beta",
              mean = mean, 
              median = median,
              lower = ~ unname(quantile(.x, 0.05)),
              upper = ~ unname(quantile(.x, 0.95))) %>%
  tidyr::extract(variable, into = c("country_idx", "covariate_idx"), 
                 regex = "beta\\[(\\d+),(\\d+)\\]", convert = TRUE) %>%
  mutate(
    country = factor(country_names[country_idx], levels = c("GHA", "BFA", "CIV")),
    covariate = covar_labels[covariate_idx],
    # A covariate is significant if the 90% CI does not contain 0
    is_significant = (lower > 0) | (upper < 0),
    sig_marker = ifelse(is_significant, "*", "")
  )

# 2. Extract the global means (using 0.05 and 0.95 for a 90% CI)
beta_bar <- fit_pooled$summary("beta_bar",
              mean = mean,
              lower = ~ unname(quantile(.x, 0.05)),
              upper = ~ unname(quantile(.x, 0.95))) %>%
  mutate(
    covariate_idx = row_number(),
    covariate = covar_labels[covariate_idx],
    country = "Global Mean",
    is_significant = (lower > 0) | (upper < 0),
    sig_marker = ifelse(is_significant, "*", "")
  )

# 3. Factor the covariates to preserve plotting order
beta_pooled$covariate <- factor(beta_pooled$covariate, levels = rev(covar_labels))
beta_bar$covariate <- factor(beta_bar$covariate, levels = rev(covar_labels))

# 4. Generate the plot
p_pooled_covars <- ggplot(beta_pooled, aes(x = mean, y = covariate, colour = country, group = country)) +
  
  # LAYER 1: The deep background (Zero line)
  # Made this solid black and slightly thinner so it stands out against the grey bands
  geom_vline(xintercept = 0, linetype = "dashed", colour = "black", linewidth = 0.6) +
  
  # LAYER 2: THE GLOBAL MEAN (beta_bar) AS A GREY BACKGROUND BAND
  # This replaces the meaningless grid lines with the actual 90% CrI of the global average
  # geom_linerange(data = beta_bar, 
  #                aes(xmin = lower, xmax = upper, y = covariate),
  #                inherit.aes = FALSE, 
  #                colour = "grey85", linewidth = 5) + 
                 
  # Adds a diamond in the exact center of the grey global band
  geom_point(data = beta_bar, 
             aes(x = mean, y = covariate),
             inherit.aes = FALSE, 
             colour = "grey", shape = 18, size = 3) +
  
  # LAYER 3: The foreground (Country-specific estimates)
  geom_pointrange(aes(xmin = lower, xmax = upper),
                  position = position_dodge(width = 0.6), size = 0.8) +
                  
  # LAYER 4: The text/asterisks (Country level)
  geom_text(aes(x = upper + 0.02, label = sig_marker),
            position = position_dodge(width = 0.6),
            size = 7, vjust = 0.75, hjust = 0, show.legend = FALSE, colour = "black") +
            
  # Global Mean Asterisks
  geom_text(data = beta_bar, 
            aes(x = upper + 0.02, y = covariate, label = sig_marker),
            inherit.aes = FALSE,
            colour = "grey50", size = 7, vjust = 0.75, hjust = 0) +
              
  # Formatting
  scale_colour_brewer(palette = "Dark2", name = "Country") +
  scale_x_continuous(expand = expansion(mult = c(0.05, 0.15))) + 
  labs(x = "Coefficient (logit scale)", y = NULL,
       title = "Covariate effects: Pooled Hierarchical Model",
       subtitle = "Thick grey bands represent the Global Mean (90% CrI). Asterisks (*) denote significance.") +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.grid.major.y = element_blank(), 
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )


ggsave("output/figures/pooled_covariate_effects.png", p_pooled_covars, width = 10, height = 7, dpi = 300, bg = "white")
p_pooled_covars
```




## Pooled Model: Full Posterior Summary

<!-- ```{r m3-pooled-summary} -->
<!-- # Full posterior summary of the pooled three-country space-time model -->
<!-- tryCatch(print(fit_pooled$summary(), n = Inf), -->
<!--          error = function(e) cat("[pooled summary failed:", conditionMessage(e), "] -->
<!-- ")) -->
<!-- ``` -->

```{r m3-pooled-key-params}

if (!exists("safe_print")) {
  safe_print <- function(fit, params, label = NULL) {
    if (!is.null(label)) cat("
--- ", label, " ---
", sep = "")
    available <- fit$metadata()$stan_variables
    base_names <- sub("\\[.*$", "", params)
    keep <- params[base_names %in% available]
    if (length(keep) == 0L) {
      cat("  [none of the requested parameters exist]
"); return(invisible(NULL))
    }
    tryCatch(fit$print(keep),
             error = function(e) cat("  [print failed:", conditionMessage(e), "]
"))
  }
}

key_pooled_params <- c(
  "mu", "sigma_b", "phi", "sigma_gamma",
  paste0("alpha[", 1:3, "]"),
  # country-specific time shifts (Stan may name these delta_country or delta_ct)
  "delta_country[1,2]", "delta_country[2,2]", "delta_country[3,2]",
  "delta_ct[1,2]", "delta_ct[2,2]", "delta_ct[3,2]",
  paste0("tau[", 1:7, "]"),
  paste0("beta_bar[", 1:7, "]")
)
safe_print(fit_pooled, key_pooled_params, label = "Pooled key parameters")
```

```{r m3-country-betas}
# Country-specific beta coefficients: 3 countries x 7 covariates = 21 betas.

if (!exists("safe_print")) {
  safe_print <- function(fit, params, label = NULL) {
    if (!is.null(label)) cat("
--- ", label, " ---
", sep = "")
    available <- fit$metadata()$stan_variables
    base_names <- sub("\\[.*$", "", params)
    keep <- params[base_names %in% available]
    if (length(keep) == 0L) {
      cat("  [none of the requested parameters exist]
"); return(invisible(NULL))
    }
    tryCatch(fit$print(keep),
             error = function(e) cat("  [print failed:", conditionMessage(e), "]
"))
  }
}

cov_names_short <- c("elev", "rain", "nightlights", "temp",
                     "travel", "popdens", "poverty")
country_labels  <- c("GHA", "CIV", "BFA")    # Stan index 1=GHA, 2=CIV, 3=BFA

for (cc in 1:3) {
  params <- sprintf("beta[%d,%d]", cc, 1:7)
  safe_print(fit_pooled, params,
             label = paste0("Country ", cc, " (", country_labels[cc],
                            ") covariate effects"))
}
```

## Pooled Discussion


## Cross-Sectional vs Space-Time Comparison


<!-- ```{r} -->
<!-- # --- Shrinkage Plot: Direct vs Pooled Model Estimates --- -->

<!-- # 1. Isolate the observed districts for Round 2 and calculate precision -->
<!-- obs_panel <- panel_pooled %>% -->
<!--   # We only have direct estimates where observed == 1 -->
<!--   filter(observed == 1, round_id == 2) %>% -->
<!--   mutate( -->
<!--     direct_prev = plogis(y_i), -->
<!--     # Lower SE (psi_i) = Higher Precision (Proxy for Sample Size) -->
<!--     precision_proxy = 1 / psi_i -->
<!--   ) -->

<!-- # 2. Extract Posterior Medians for the exact same districts -->
<!-- model_prev <- fit_pooled$summary("prevalence", median = median) %>% -->
<!--   mutate(row_id = as.integer(gsub("prevalence\\[|\\]", "", variable))) %>% -->
<!--   # Map back to the full panel to get district info, then filter -->
<!--   left_join(panel_pooled %>% mutate(row_id = row_number()), by = "row_id") %>% -->
<!--   filter(observed == 1, round_id == 2) %>% -->
<!--   select(row_id, model_prev = median) -->

<!-- # 3. Join the datasets -->
<!-- shrinkage_df <- obs_panel %>% -->
<!--   left_join(model_prev, by = "row_id") -->

<!-- # 4. Generate the faceted Shrinkage Plot -->
<!-- p_shrinkage <- ggplot(shrinkage_df, aes(x = direct_prev, y = model_prev)) + -->
<!--   # The 1:1 Diagonal Line (No Shrinkage / Perfect Agreement) -->
<!--   geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "black", linewidth = 0.8) + -->

<!--   # Draw a faint segment from the point to the diagonal to explicitly highlight the "pull" -->
<!--   geom_segment(aes(xend = direct_prev, yend = direct_prev), alpha = 0.3, colour = "grey60") + -->

<!--   # The Data Points, colored by our sample size proxy -->
<!--   geom_point(aes(fill = precision_proxy), shape = 21, colour = "grey30", size = 3, alpha = 0.85) + -->

<!--   # Facet by country to create the small multiples -->
<!--   facet_wrap(~country, scales = "free") + -->

<!--   # Color Scale: High precision (large sample) = Dark Blue, Low precision = Yellow/Bright -->
<!--   scale_fill_viridis_c( -->
<!--     name = "Data Precision\n(1 / SE)", -->
<!--     option = "cividis", -->
<!--     direction = -1 -->
<!--   ) + -->
<!--   scale_x_continuous(labels = scales::percent_format(accuracy = 1)) + -->
<!--   scale_y_continuous(labels = scales::percent_format(accuracy = 1)) + -->
<!--   labs( -->
<!--     title = "Hierarchical Shrinkage: Joint Pooled Model vs Direct Estimates", -->
<!--     subtitle = "Yellow points (low precision/small sample) are pulled strongly toward the spatial mean. Dark points stay on the diagonal.", -->
<!--     x = "Direct Estimate (Noisy Survey Data)", -->
<!--     y = "Model Posterior Median (Smoothed)" -->
<!--   ) + -->
<!--   theme_minimal(base_size = 13) + -->
<!--   theme( -->
<!--     legend.position = "right", -->
<!--     strip.text = element_text(size = 12, face = "bold"), -->
<!--     panel.grid.minor = element_blank(), -->
<!--     panel.spacing = unit(1.5, "lines") -->
<!--   ) -->

<!-- ggsave("output/figures/pooled_shrinkage_plot.png", p_shrinkage, width = 12, height = 5, dpi = 300, bg = "white") -->
<!-- p_shrinkage -->

<!-- ``` -->




## Pooled Model Maps

```{r three-country-covariate-maps, fig.height=10, fig.width=14}
full_panel <- read_csv("output/three_country_full_panel.csv", show_col_types = FALSE)

# Fix the coastal extraction artifacts, then filter for Round 2
panel_r2_cov <- full_panel %>%
  # Group by country to calculate a safe replacement value
  group_by(country, time) %>%
  mutate(
    # Calculate the median rainfall for the country (ignoring the negative outliers)
    safe_rainfall = median(rainfall_sep_dec[rainfall_sep_dec >= 0], na.rm = TRUE),
    
    
    rainfall_sep_dec = ifelse(rainfall_sep_dec < 0, safe_rainfall, rainfall_sep_dec)
  ) %>%
  ungroup() %>%
  filter(time == 2) %>%
  select(country, district_id, district_name,
         nightlights, elev_mean_pw, rainfall_sep_dec, temp_sep_dec,
         travel_time_pw, census_pop_density, poverty_H)

# Join to the shapefile
covar_map_df <- all_sf %>%
  left_join(panel_r2_cov, by = c("country", "local_id" = "district_id"))

covar_map_df <- all_sf %>%
  left_join(panel_r2_cov, by = c("country", "local_id" = "district_id"))

country_borders <- all_sf %>%
  group_by(country) %>%
  summarise(geometry = st_union(geometry), .groups = "drop")

make_covar_map <- function(data, var, title, palette = "viridis",
                            direction = 1, trans = "identity") {
  ggplot() +
    geom_sf(data = data, aes(fill = .data[[var]]), colour = "grey50", linewidth = 0.05) +
    geom_sf(data = country_borders, fill = NA, colour = "black", linewidth = 0.8) +
    scale_fill_viridis_c(name = title, option = palette, direction = direction,
                          trans = trans, na.value = "grey80") +
    theme_void(base_size = 10) +
    labs(title = title) +
    theme(legend.position = "right",
          plot.title = element_text(size = 11, face = "bold"))
}

p_ntl  <- make_covar_map(covar_map_df, "nightlights", "Nightlights\n(nW/cm2/sr)", "magma")
p_elev <- make_covar_map(covar_map_df, "elev_mean_pw", "Elevation\n(m)", "terrain", direction = -1)
p_rain <- make_covar_map(covar_map_df, "rainfall_sep_dec", "Rainfall\nSep-Dec (mm)", "Blues")
p_temp <- make_covar_map(covar_map_df, "temp_sep_dec", "Temperature\nSep-Dec (C)", "inferno")
p_trav <- make_covar_map(covar_map_df, "travel_time_pw", "Travel time\n(minutes)", "YlOrRd")
p_pov  <- make_covar_map(covar_map_df, "poverty_H", "Poverty\nheadcount", "RdYlGn", direction = -1)

p_all_covars <- (p_ntl | p_elev | p_rain) / (p_temp | p_trav | p_pov) +
  plot_annotation(
    title = "District-level covariates across Ghana, Burkina Faso, and Cote d'Ivoire",
    subtitle = "Raw (non-standardised) values for Round 2. Heavy lines = country borders."
  )

ggsave("output/figures/three_country_covariates_all.png", p_all_covars,
       width = 14, height = 10, dpi = 300, bg = "white")
p_all_covars
```

```{r nightlights-vs-prevalence, fig.height=7, fig.width=14}
p_ntl_big <- ggplot() +
  geom_sf(data = covar_map_df, aes(fill = nightlights), colour = "grey50", linewidth = 0.05) +
  geom_sf(data = country_borders, fill = NA, colour = "black", linewidth = 0.8) +
  scale_fill_viridis_c(name = "Nightlights\n(nW/cm2/sr)", option = "magma",
                        trans = "log1p", na.value = "grey80") +
  labs(title = "Nightlights (VIIRS, log scale)") +
  theme_void(base_size = 12) +
  theme(legend.position = "right", plot.title = element_text(size = 13, face = "bold"))

pooled_prev_r2 <- fit_pooled$summary("prevalence", median = median) %>%
  mutate(row_id = as.integer(gsub("prevalence\\[|\\]", "", variable))) %>%
  left_join(panel_pooled %>% mutate(row_id = row_number()), by = "row_id") %>%
  filter(round_id == 2)

pooled_map_df <- all_sf %>%
  left_join(pooled_prev_r2, by = c("global_id" = "global_district_id"))

p_prev_big <- ggplot() +
  geom_sf(data = pooled_map_df, aes(fill = median), colour = "grey50", linewidth = 0.05) +
  geom_sf(data = country_borders, fill = NA, colour = "black", linewidth = 0.8) +
  scale_fill_viridis_c(name = "Prevalence",
                        labels = scales::percent_format(accuracy = 1),
                        option = "magma", direction = -1, na.value = "grey80") +
  labs(title = "Under-5 malaria prevalence (pooled model)") +
  theme_void(base_size = 12) +
  theme(legend.position = "right", plot.title = element_text(size = 13, face = "bold"))

p_ntl_vs_prev <- p_ntl_big | p_prev_big
ggsave("output/figures/nightlights_vs_prevalence.png", p_ntl_vs_prev,
       width = 14, height = 7, dpi = 300, bg = "white")
p_ntl_vs_prev
```

```{r pooled-prevalence-map, fig.height=8, fig.width=10}
p_pooled_map <- ggplot() +
  geom_sf(data = pooled_map_df, aes(fill = median), colour = "grey40", linewidth = 0.1) +
  geom_sf(data = country_borders, fill = NA, colour = "black", linewidth = 0.8) +
  scale_fill_viridis_c(name = "Prevalence",
                        labels = scales::percent_format(accuracy = 1),
                        option = "magma", direction = -1) +
  labs(title = "Unified West African Under-5 Malaria Prevalence (2021/2022)",
       subtitle = "Posterior medians from the 724-district pooled hierarchical model") +
  theme_void(base_size = 14) +
  theme(legend.position = "right")

ggsave("output/figures/pooled_prevalence_map.png", p_pooled_map,
       width = 10, height = 8, dpi = 300, bg = "white")
p_pooled_map
```

```{r three-country-adjacency, fig.height=8, fig.width=10}
centroids_all <- st_centroid(all_sf)
coords_all <- st_coordinates(centroids_all)

edges_all_df <- tibble(
  from = el$node1, to = el$node2,
  x_from = coords_all[el$node1, 1], y_from = coords_all[el$node1, 2],
  x_to   = coords_all[el$node2, 1], y_to   = coords_all[el$node2, 2],
  edge_type = ifelse(all_sf$country[el$node1] == all_sf$country[el$node2],
                     "Within-country", "Cross-border")
)

n_within <- sum(edges_all_df$edge_type == "Within-country")
n_cross  <- sum(edges_all_df$edge_type == "Cross-border")

p_adj_3country <- ggplot() +
  geom_sf(data = all_sf, fill = "grey95", colour = "grey80", linewidth = 0.05) +
  geom_segment(data = edges_all_df %>% filter(edge_type == "Within-country"),
               aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
               colour = "#93c5fd", alpha = 0.12, linewidth = 0.15) +
  geom_segment(data = edges_all_df %>% filter(edge_type == "Cross-border"),
               aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
               colour = "#dc2626", alpha = 0.85, linewidth = 0.6) +
  geom_sf(data = country_borders, fill = NA, colour = "black", linewidth = 1) +
  labs(title = paste0("Combined 724-district adjacency graph"),
       subtitle = paste0("Within-country edges (blue): ", n_within,
                          " | Cross-border edges (red): ", n_cross)) +
  theme_void(base_size = 12)

ggsave("output/figures/three_country_adjacency.png", p_adj_3country,
       width = 10, height = 8, dpi = 300, bg = "white")
p_adj_3country
```

## Pooled Model Reliability: CV Analysis

```{r pooled-cv-direct}
cv_df <- direct_est_all %>%
  filter(round_id == 2) %>%
  mutate(
    p_hat = plogis(y_i),
    se_p = psi_i * p_hat * (1 - p_hat),
    cv = se_p / p_hat,
    
    # UNIFIED THRESHOLDS
    reliability = case_when(
      cv < 0.16 ~ "Reliable (CV < 16%)",
      cv < 0.33 ~ "Caution (CV 16-33%)",
      TRUE      ~ "Unreliable (CV > 33%)"
    ),
    reliability = factor(reliability, levels = c(
      "Reliable (CV < 16%)", 
      "Caution (CV 16-33%)", 
      "Unreliable (CV > 33%)"
    ))
  )

# 2. Generate the faceted scatter plot
p_cv_direct <- ggplot(cv_df, aes(x = p_hat, y = cv, fill = reliability)) +
  geom_hline(yintercept = 0.16, linetype = "dashed", colour = "#16a34a", linewidth = 0.8) +
  geom_hline(yintercept = 0.33, linetype = "dashed", colour = "#dc2626", linewidth = 0.8) +
  
  geom_point(shape = 21, colour = "grey20", size = 2.5, alpha = 0.85) +
  facet_wrap(~country, ncol = 3) +
  
  scale_fill_manual(
    name = "Estimate Reliability",
    values = c(
      "Reliable (CV < 16%)" = "#16a34a",
      "Caution (CV 16-33%)" = "#f59e0b",
      "Unreliable (CV > 33%)" = "#dc2626"
    )
  ) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  # LOCKED Y-AXIS (0 to 1)
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  
  labs(
    title = "Direct Estimates: High Uncertainty (Survey Data Only)",
    subtitle = "Raw estimates before spatial smoothing.",
    x = "Direct Prevalence Estimate",
    y = "Coefficient of Variation"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold")
  )

ggsave("output/figures/direct_cv_scatter_fixed.png", p_cv_direct, width = 12, height = 5, dpi = 300, bg = "white")
```

```{r pooled-cv-posterior}



pooled_full <- fit_pooled$summary("prevalence",
  mean = mean, median = median, sd = sd) %>%
  mutate(row_id = as.integer(gsub("prevalence\\[|\\]", "", variable))) %>%
  left_join(panel_pooled %>% mutate(row_id = row_number()), by = "row_id") %>%
  filter(round_id == 2) %>%
  mutate(
    cv = sd / mean,
    
    # UNIFIED THRESHOLDS 
    reliability = case_when(
      cv < 0.16 ~ "Reliable (CV < 16%)",
      cv < 0.33 ~ "Caution (CV 16-33%)",
      TRUE      ~ "Unreliable (CV > 33%)"
    ),
    reliability = factor(reliability, levels = c(
      "Reliable (CV < 16%)",
      "Caution (CV 16-33%)",
      "Unreliable (CV > 33%)"
    ))
  )

# 2. Generate the faceted scatter plot
p_cv_pooled <- ggplot(pooled_full, aes(x = median, y = cv, fill = reliability)) +
  geom_hline(yintercept = 0.16, linetype = "dashed", colour = "#16a34a", linewidth = 0.8) +
  geom_hline(yintercept = 0.33, linetype = "dashed", colour = "#dc2626", linewidth = 0.8) +
  
  geom_point(shape = 21, colour = "grey20", size = 2.5, alpha = 0.85) +
  facet_wrap(~country, ncol = 3) +
  
  scale_fill_manual(
    name = "Estimate Reliability",
    values = c(
      "Reliable (CV < 16%)" = "#16a34a",
      "Caution (CV 16-33%)" = "#f59e0b",
      "Unreliable (CV > 33%)" = "#dc2626"
    )
  ) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  # LOCKED Y-AXIS (0 to 1)
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  
  labs(
    title = "Pooled Model: Dramatic Reduction in Uncertainty",
    subtitle = "Hierarchical smoothing rescues unobserved and noisy districts.",
    x = "Posterior Median Prevalence",
    y = "Coefficient of Variation",
    fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold")
  )

ggsave("output/figures/cv_reliability_scatter_pooled_fixed.png", p_cv_pooled, width = 12, height = 5, dpi = 300, bg = "white")
```

```{r pooled-cv-comparison}


# 1. Merge the direct and pooled dataframes using inner_join

cv_comparison <- cv_df %>%
  select(district_id, country, cv_direct = cv) %>% 
  inner_join(
    pooled_full %>% select(district_id, cv_pooled = cv), 
    by = c("district_id") 
  )

# 2. Generate the Variance Reduction Plot
p_cv_comparison <- ggplot(cv_comparison, aes(x = cv_direct, y = cv_pooled)) +
  
  # The 45-degree line: Anything below this line represents a reduction in variance
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey40", linewidth = 1) +
  
  # The 20% Policy Threshold lines
  geom_vline(xintercept = 0.20, colour = "#e3342f", linetype = "dotted", linewidth = 0.8) +
  geom_hline(yintercept = 0.20, colour = "#10b981", linetype = "dotted", linewidth = 0.8) +
  
  # The data points, colored by country
  geom_point(aes(fill = country), shape = 21, colour = "white", size = 3, alpha = 0.7) +
  
  scale_fill_brewer(palette = "Dark2", name = "Country") +
  
  # Force both axes to have the exact same scale and limits so the geometry is honest
  scale_x_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  
  # Annotations
  annotate("text", x = 0.75, y = 0.15, label = "Variance Reduced", 
           colour = "#10b981", fontface = "bold", size = 5) +
  
  labs(
    title = "Variance Reduction: Direct vs. Pooled CV",
    subtitle = "Points below the dashed line indicate precision gained by the hierarchical model.",
    x = "Direct Estimate CV (Survey Data)",
    y = "Pooled Model CV (Hierarchical Smoothing)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")

ggsave("output/figures/variance_reduction_45deg.png", p_cv_comparison, width = 8, height = 8, dpi = 300, bg = "white")
p_cv_comparison
```

```{r pooled-cv-rescue}
# 1. Categorize all districts based on strict StatCan guidelines
rescue_df <- pooled_full %>% 
  filter(round_id == 2) %>% 
  select(district_id, country, cv_pooled = cv) %>%
  left_join(
    cv_df %>% filter(round_id == 2) %>% select(district_id, cv_direct = cv), 
    by = "district_id"
  ) %>%
  mutate(
    # Categorize the "Before" state (Raw DHS Data)
    status_before = case_when(
      is.na(cv_direct)   ~ "1. No Data (Unobserved)",
      cv_direct <= 0.165 ~ "2. Acceptable (CV ≤ 16.5%)",
      cv_direct <= 0.333 ~ "3. Marginal (CV 16.6-33.3%)",
      TRUE               ~ "4. Unacceptable (CV > 33.3%)"
    ),
    # Categorize the "After" state (Pooled Model)
    status_after = case_when(
      cv_pooled <= 0.165 ~ "Acceptable (CV ≤ 16.5%)",
      cv_pooled <= 0.333 ~ "Marginal (CV 16.6-33.3%)",
      TRUE               ~ "Unacceptable (CV > 33.3%)"
    )
  )


transition_matrix <- rescue_df %>%
  count(status_before, status_after) %>%
  pivot_wider(names_from = status_after, values_from = n, values_fill = 0) %>%
  arrange(status_before)

print(transition_matrix)


rescued_districts <- rescue_df %>%
  filter(
    status_before %in% c("1. No Data (Unobserved)", "3. Marginal (CV 16.6-33.3%)", "4. Unacceptable (CV > 33.3%)") & 
    status_after == "Acceptable (CV ≤ 16.5%)"
  ) %>%
  nrow()

cat("\nTotal districts rescued into the Acceptable zone:", rescued_districts, "\n")
```

```{r cv-reliability-scatter, fig.height=6, fig.width=12}
# Extract prevalence and CV from the pooled model (Round 2)
pooled_full <- fit_pooled$summary("prevalence",
  mean = mean, median = median, sd = sd) %>%
  mutate(row_id = as.integer(gsub("prevalence\\[|\\]", "", variable))) %>%
  left_join(panel_pooled %>% mutate(row_id = row_number()), by = "row_id") %>%
  filter(round_id == 2) %>%
  mutate(
    cv = sd / mean,
    reliability = case_when(
      cv < 0.16 ~ "Reliable (CV < 16%)",
      cv < 0.33 ~ "Caution (CV 16-33%)",
      TRUE      ~ "Unreliable (CV > 33%)"
    ),
    reliability = factor(reliability, levels = c("Reliable (CV < 16%)",
                                                   "Caution (CV 16-33%)",
                                                   "Unreliable (CV > 33%)"))
  )

p_cv_reliability <- ggplot(pooled_full, aes(x = median, y = cv, colour = reliability)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_hline(yintercept = 0.16, linetype = "dashed", colour = "#16a34a", linewidth = 0.5) +
  geom_hline(yintercept = 0.33, linetype = "dashed", colour = "#dc2626", linewidth = 0.5) +
  scale_colour_manual(values = c("Reliable (CV < 16%)" = "#16a34a",
                                  "Caution (CV 16-33%)" = "#f59e0b",
                                  "Unreliable (CV > 33%)" = "#dc2626")) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~country) +
  labs(
    title = "Pooled model: prevalence vs coefficient of variation by reliability",
    subtitle = "Green dashed = 16% CV threshold. Red dashed = 33% threshold. Most districts now in the reliable region.",
    x = "Posterior median prevalence", y = "Coefficient of variation",
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 12, face = "bold"))

ggsave("output/figures/cv_reliability_scatter_pooled.png", p_cv_reliability,
       width = 12, height = 6, dpi = 300, bg = "white")
p_cv_reliability

# Print summary
pooled_full %>%
  count(country, reliability) %>%
  pivot_wider(names_from = reliability, values_from = n, values_fill = 0) %>%
  print()
```

```{r pooled-cv-map, fig.height=8, fig.width=10}
pooled_cv_df <- pooled_full %>%
  select(global_district_id, cv)

pooled_cv_map <- all_sf %>%
  left_join(pooled_cv_df, by = c("global_id" = "global_district_id"))

p_pooled_cv <- ggplot() +
  geom_sf(data = pooled_cv_map, aes(fill = cv), colour = "grey40", linewidth = 0.1) +
  geom_sf(data = country_borders, fill = NA, colour = "black", linewidth = 0.8) +
  scale_fill_viridis_c(name = "CV",
                        labels = scales::percent_format(accuracy = 1),
                        option = "cividis", direction = -1) +
  labs(title = "Coefficient of Variation: Pooled 3-Country Model (Round 2)",
       subtitle = "Low CV (dark) = high confidence. High CV (bright) = greater uncertainty.") +
  theme_void(base_size = 14) +
  theme(legend.position = "right")

ggsave("output/figures/pooled_cv_map.png", p_pooled_cv,
       width = 10, height = 8, dpi = 300, bg = "white")
p_pooled_cv
```

```{r pooled-prevalence-draws}
prev_draws <- fit_pooled$draws("prevalence", format = "df") %>%
  as_tibble() %>%
  select(starts_with("prevalence["), .draw) %>%
  pivot_longer(starts_with("prevalence["), names_to = "variable", values_to = "prev") %>%
  mutate(row_id = as.integer(gsub("prevalence\\[|\\]", "", variable)))


prev_full <- prev_draws %>%
  left_join(full_panel %>% mutate(row_id = row_number()), by = "row_id")


national_posteriors <- prev_full %>%
  group_by(.draw, country, time) %>%
  summarise(
    national_prev = weighted.mean(prev, w = census_pop, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Round = factor(ifelse(time == 1, "Round 1 (Baseline)", "Round 2 (Follow-up)"),
                   levels = c("Round 1 (Baseline)", "Round 2 (Follow-up)"))
  )


p_national_post <- ggplot(national_posteriors, aes(x = national_prev, y = country, fill = Round)) +
  
  
  geom_boxplot(
    alpha = 0.85, 
    position = position_dodge(width = 0.75), 
    outlier.alpha = 0.05, 
    outlier.size = 0.5,
    colour = "grey20"
  ) +
  
  scale_fill_manual(
    values = c("Round 1 (Baseline)" = "#e6b000", "Round 2 (Follow-up)" = "#3182bd")
  ) +
  
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  
  labs(
    title = "Posterior Uncertainty: National Malaria Prevalence",
    subtitle = "Population-weighted aggregation of 724 district-level posterior draws.",
    x = "Estimated Prevalence",
    y = NULL,
    fill = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

ggsave("output/figures/posterior_boxplot_prevalence.png", p_national_post, width = 10, height = 6, dpi = 300, bg = "white")
```

# Model 4: Round-2-Only Pooled Sensitivity Model {#sec:model4}

## Motivation

The full three-country pooled model joins two DHS rounds per country.
The Round 1 surveys, however, span four calendar years across countries
(Burkina Faso 2010, Côte d'Ivoire 2011, Ghana 2014), while the Round 2
surveys are tightly clustered within a single year (Burkina Faso 2021,
Côte d'Ivoire 2021, Ghana 2022). Because the BYM2 component smooths
spatial residuals across the combined cross-border adjacency graph, the
Round 1 calendar offset raises a legitimate concern that cross-border
edges may absorb temporal as well as spatial variation during a period
of substantial intervention scale-up across the region [@bhatt2015;
@weiss2020].

To isolate the cross-border spatial story from this temporal-misalignment
concern, we fit a Round-2-only pooled model. Restricting to the 2021–2022
window guarantees that cross-border smoothing operates on contemporaneous
prevalence surfaces; the published cross-border SAE precedent of
@saldarriaga2020 used a comparable two-year alignment window (2007–2009).
With a single time point, the Knorr-Held Type I interaction $\gamma_{it}$
and the country-specific time shifts $\delta_{c,t}$ become unnecessary,
and the model reduces to a pure spatial BYM2 with hierarchical
country-specific covariate effects. Findings shared between this model
and the full space-time pooled specification are the most robust results
of the analysis; differences localise the contribution of the Round 1
data and the cross-temporal smoothing.

## Specification

For each observed Round 2 district $i$, the design-based logit prevalence
direct estimate $y_i$ retains the same Fay-Herriot likelihood used
throughout [@wakefield2020]:
$$y_i \sim \mathcal{N}\!\left(\theta_i, \; \psi_i^2 \right),$$
with $\psi_i$ obtained by Taylor-series linearisation [@binder1983;
@lumley2010] and transformed to the logit scale via the delta method,
exactly as in the space-time models.

The latent logit prevalence decomposes as
$$\theta_i = \mu + \alpha_{c[i]} + b_i +
            \mathbf{x}_i^{\top} \bm{\beta}_{c[i]},$$
where $\mu$ is the regional logit intercept across the three countries,
$\alpha_{c[i]}$ is a country-specific deviation with the
sum-to-zero constraint $\sum_c \alpha_c = 0$ for identifiability with
$\mu$, $b_i$ is the BYM2 spatial random effect on the combined
724-district graph, and $\bm{\beta}_{c[i]}$ is the country-specific
vector of covariate effects on the seven standardised covariates.

The BYM2 spatial effect uses the same scaled, non-centred parameterisation
as the space-time models [@riebler2016]:
$$b_i = \sigma_b \left(
  \sqrt{1-\phi}\; v_i + \sqrt{\phi / s}\; u_i
\right),
\qquad
v_i \stackrel{\mathrm{iid}}{\sim} \mathcal{N}(0,1),
\qquad
u_i \sim \mathrm{ICAR}(W),$$
with $u_i$ sum-constrained, $W$ the combined adjacency matrix including
cross-border queen-contiguity edges, and $s$ the scaling factor of
@riebler2016 making $\phi \in (0,1)$ interpretable as the structured-
variance fraction. Disconnected components are reconnected by snap
tolerance to ensure a proper ICAR [@freni2018].

Country-specific covariate effects are partially pooled through a
hierarchical prior,
$$\beta_{ck} \;\sim\; \mathcal{N}\!\left(\bar{\beta}_k, \tau_k^2\right),
\qquad c \in \{1,2,3\}, \quad k \in \{1, \ldots, 7\},$$
where $\bar{\beta}_k$ is the regional consensus effect of covariate $k$
and $\tau_k$ the between-country standard deviation. Small $\tau_k$
indicates that countries share a common relationship between covariate
$k$ and malaria risk; large $\tau_k$ allows the slopes to diverge.
The slopes are implemented non-centred as
$\beta_{ck} = \bar{\beta}_k + \tau_k z_{ck}$ with
$z_{ck} \sim \mathcal{N}(0,1)$.

## Priors

Priors match those used in the country-specific and full pooled models
so that posterior comparisons are not driven by prior changes:
$\mu \sim \mathcal{N}(-1.85, 1)$, $\alpha_{\text{raw}} \sim \mathcal{N}(0, 1)$
with sum-to-zero, $\sigma_b \sim \mathrm{Exp}(4.6)$ (PC-style;
approximates $P(\sigma_b > 1) = 0.01$ [@simpson2017]),
$\phi \sim \mathrm{Beta}(3, 2)$, $\bar{\beta}_k \sim \mathcal{N}(0, 1)$,
$\tau_k \sim \mathrm{Exp}(2)$.

## Stan implementation

```{r r2-stan-write, eval = FALSE}
bym2_r2_pooled_code <- "
data {
  int<lower=1> N;
  int<lower=1> C;
  array[N] int<lower=1, upper=C> country;

  int<lower=1> N_obs;
  array[N_obs] int<lower=1, upper=N> obs_idx;
  vector[N_obs] y;
  vector<lower=0>[N_obs] psi;

  int<lower=1> N_edges;
  array[N_edges] int<lower=1, upper=N> node1;
  array[N_edges] int<lower=1, upper=N> node2;
  real<lower=0> scaling_factor;

  int<lower=1> K;
  matrix[N, K] X;
}
parameters {
  real mu;
  vector[C] alpha_raw;
  vector[N] v;
  vector[N] u_raw;
  real<lower=0> sigma_b;
  real<lower=0, upper=1> phi;
  vector[K] beta_bar;
  vector<lower=0>[K] tau;
  matrix[C, K] beta_z;
}
transformed parameters {
  vector[C] alpha = alpha_raw - mean(alpha_raw);
  vector[N] u = u_raw - mean(u_raw);
  vector[N] b = sigma_b * (sqrt(1 - phi) * v
                           + sqrt(phi / scaling_factor) * u);
  matrix[C, K] beta;
  for (c in 1:C)
    for (k in 1:K)
      beta[c, k] = beta_bar[k] + tau[k] * beta_z[c, k];

  vector[N] theta;
  for (i in 1:N) {
    int ci = country[i];
    theta[i] = mu + alpha[ci] + b[i] + dot_product(X[i], beta[ci]');
  }
}
model {
  mu        ~ normal(-1.85, 1.0);
  alpha_raw ~ normal(0, 1.0);
  sigma_b   ~ exponential(4.6);
  phi       ~ beta(3, 2);

  v         ~ std_normal();
  target   += -0.5 * dot_self(u_raw[node1] - u_raw[node2]);
  sum(u_raw) ~ normal(0, 0.001 * N);

  beta_bar  ~ normal(0, 1);
  tau       ~ exponential(2);
  to_vector(beta_z) ~ std_normal();

  y ~ normal(theta[obs_idx], psi);
}
generated quantities {
  vector[N_obs] log_lik;
  for (i in 1:N_obs)
    log_lik[i] = normal_lpdf(y[i] | theta[obs_idx[i]], psi[i]);

  vector[N] prevalence;
  for (i in 1:N)
    prevalence[i] = inv_logit(theta[i]);

  vector[C] p_national;
  for (c in 1:C) {
    real sum_p = 0;
    int  n_c   = 0;
    for (i in 1:N) {
      if (country[i] == c) {
        sum_p += prevalence[i];
        n_c   += 1;
      }
    }
    p_national[c] = sum_p / n_c;
  }
}
"
writeLines(bym2_r2_pooled_code, "fh_bym2_pooled_3country_round2.stan")
```

## Build Stan data for Round 2 and fit

The Round-2-only model reuses the same combined adjacency graph
(`el`, `scaling_factor_all`), the same country index vector
(`country_vec`), and the same district-id frame (`D_all = 724`) used by
the full pooled model. Only the panel and covariate matrix are subset to
$t = 2$.

```{r r2-stan-data, eval = FALSE}
# Subset the pooled panel to Round 2 only
panel_r2 <- panel_pooled %>% filter(round_id == 2L)

obs_mask_r2 <- panel_r2$observed == 1
obs_idx_r2  <- which(obs_mask_r2)
N_obs_r2    <- length(obs_idx_r2)

# Round-2 covariate matrix in the same district order
X_r2 <- panel_r2 %>%
  select(-any_of(cov_names)) %>%
  left_join(
    model_panel %>% filter(time == 2L) %>%
      select(country, district_id, all_of(cov_names)),
    by = c("country", "district_id" = "district_id")
  ) %>%
  select(all_of(cov_names)) %>%
  as.matrix()

stan_data_r2 <- list(
  N              = D_all,
  C              = 3L,
  country        = country_vec,
  N_obs          = N_obs_r2,
  obs_idx        = obs_idx_r2,
  y              = panel_r2$y_i[obs_idx_r2],
  psi            = panel_r2$psi_i[obs_idx_r2],
  N_edges        = el$N_edges,
  node1          = el$node1,
  node2          = el$node2,
  scaling_factor = scaling_factor_all,
  K              = length(cov_names),
  X              = X_r2
)

cat("Round-2 pooled Stan data assembled.\n")
cat("  N =", stan_data_r2$N, "\n")
cat("  N_obs =", stan_data_r2$N_obs, "\n")
cat("  K =", stan_data_r2$K, "\n")
cat("  N_edges =", stan_data_r2$N_edges, "\n")
```

```{r r2-stan-fit, eval = FALSE}
mod_r2 <- cmdstan_model("fh_bym2_pooled_3country_round2.stan")
mod_r2$check_syntax(pedantic = TRUE)

fit_r2 <- mod_r2$sample(
  data            = stan_data_r2,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 1500,
  iter_sampling   = 2000,
  adapt_delta     = 0.97,
  max_treedepth   = 13,
  seed            = 2026,
  refresh         = 200
)

fit_r2$summary(c("mu", "alpha", "sigma_b", "phi",
                 "beta_bar", "tau", "p_national"))
```

## Comparison to the full pooled model

Three diagnostics localise what the Round 1 data contribute to the full
pooled posterior, and whether the cross-border component depends on
cross-temporal smoothing.

```{r r2-vs-full-spatial, eval = FALSE}
# Spatial hyperparameters: do sigma_b and phi shift when Round 1 is removed?
spatial_compare <- bind_rows(
  fit_pooled$summary(c("sigma_b", "phi")) %>%
    mutate(model = "Full space-time pooled (R1 + R2)"),
  fit_r2$summary(c("sigma_b", "phi")) %>%
    mutate(model = "Round-2-only pooled")
)
spatial_compare
```

A stable $\phi$ across the two models supports the interpretation that
the high $\phi$ found in the full pooled model is driven by genuine
spatial structure rather than by cross-temporal smoothing of the Round 1
calendar offset. A substantial drop in $\phi$ between models would
suggest the converse.

```{r r2-vs-full-covariates, eval = FALSE}
# Regional consensus covariate effects: stable across models?
covar_labels  <- c("Elevation", "Rainfall", "Nighttime lights", "Temperature",
                   "Travel time", "Pop. density (log)", "Poverty")

bbar_full <- fit_pooled$summary("beta_bar",
              mean = mean,
              lower = ~ unname(quantile(.x, 0.05)),
              upper = ~ unname(quantile(.x, 0.95))) %>%
  mutate(covariate = covar_labels,
         model = "Full space-time pooled")

bbar_r2 <- fit_r2$summary("beta_bar",
              mean = mean,
              lower = ~ unname(quantile(.x, 0.05)),
              upper = ~ unname(quantile(.x, 0.95))) %>%
  mutate(covariate = covar_labels,
         model = "Round-2-only pooled")

bind_rows(bbar_full, bbar_r2)
```

Differences between the regional means $\bar{\beta}_k$ across the two
models identify covariate effects whose estimated value depends on the
Round 1 data — e.g., covariates whose effect plausibly changed over time
or whose Round 1 measurement quality differed (poverty in particular,
which was only available from the most recent census in Ghana).

```{r r2-vs-full-prevalence-r2, eval = FALSE}
# District-level Round 2 prevalences: how much does adding Round 1 move them?
prev_full_r2 <- fit_pooled$summary("prevalence", median = median) %>%
  mutate(row_id = as.integer(gsub("prevalence\\[|\\]", "", variable))) %>%
  left_join(panel_pooled %>% mutate(row_id = row_number()),
            by = "row_id") %>%
  filter(round_id == 2) %>%
  transmute(country, district_id, prev_full = median)

prev_r2_only <- fit_r2$summary("prevalence", median = median) %>%
  mutate(row_id = as.integer(gsub("prevalence\\[|\\]", "", variable))) %>%
  left_join(panel_r2 %>% mutate(row_id = row_number()),
            by = "row_id") %>%
  transmute(country, district_id, prev_r2 = median)

prev_compare <- prev_full_r2 %>%
  inner_join(prev_r2_only, by = c("country", "district_id")) %>%
  mutate(diff = prev_full - prev_r2)

summary(prev_compare$diff)
```

If the median absolute district-level prevalence difference is small
(say, under 1–2 percentage points), the Round 2 estimates are robust to
the inclusion of the temporally misaligned Round 1 data, and the full
pooled model's Round 2 inferences can be reported with confidence. If
the differences are large, the Round-2-only estimates should be treated
as the primary Round 2 result.

## Interpretation

The Round-2-only pooled model functions as one of three complementary
specifications in this analysis. The country-specific space-time models
provide the temporal-change story (national odds ratios between rounds,
district-level deviations $\gamma_{it}^{\,*}$) under the methodologically
conservative convention of fitting each country independently
[@wakefield2020]. The Round-2-only pooled model presented here provides
the cross-border borrowing story — the regional consensus covariate
effects $\bar{\beta}_k$, the spatial parameters $\sigma_b$ and $\phi$ on
the combined graph, and the hierarchical rescue of country-specific
signals — under temporal alignment guaranteed by construction. The full
space-time pooled model integrates both extensions, with the Round 1
calendar offset acknowledged as a limitation. Findings that are stable
across all three specifications are the most robust results of this
work.


## Round-2 Pooled Model: Full Posterior Summary

```{r m4-r2-pooled-summary, eval = FALSE}
# Full posterior summary of the Round-2-only pooled model
tryCatch(print(fit_r2$summary(), n = Inf),
         error = function(e) cat("[R2-only summary failed:", conditionMessage(e), "]
"))
```

```{r m4-r2-pooled-key-params, eval = FALSE}
if (!exists("safe_print")) {
  safe_print <- function(fit, params, label = NULL) {
    if (!is.null(label)) cat("
--- ", label, " ---
", sep = "")
    available <- fit$metadata()$stan_variables
    base_names <- sub("\\[.*$", "", params)
    keep <- params[base_names %in% available]
    if (length(keep) == 0L) {
      cat("  [none of the requested parameters exist]
"); return(invisible(NULL))
    }
    tryCatch(fit$print(keep),
             error = function(e) cat("  [print failed:", conditionMessage(e), "]
"))
  }
}
key_r2_params <- c(
  "mu", "sigma_b", "phi",
  paste0("alpha[", 1:3, "]"),
  paste0("tau[", 1:7, "]"),
  paste0("beta_bar[", 1:7, "]"),
  paste0("p_national[", 1:3, "]")
)

safe_print(fit_r2, key_r2_params, label = "R2-only pooled key parameters")

```

```{r m4-r2-country-betas, eval = FALSE}
if (!exists("safe_print")) {
  safe_print <- function(fit, params, label = NULL) {
    if (!is.null(label)) cat("
--- ", label, " ---
", sep = "")
    available <- fit$metadata()$stan_variables
    base_names <- sub("\\[.*$", "", params)
    keep <- params[base_names %in% available]
    if (length(keep) == 0L) {
      cat("  [none of the requested parameters exist]
"); return(invisible(NULL))
    }
    tryCatch(fit$print(keep),
             error = function(e) cat("  [print failed:", conditionMessage(e), "]
"))
  }
}
cov_names_short <- c("elev", "rain", "nightlights", "temp",
                     "travel", "popdens", "poverty")
country_labels  <- c("GHA", "CIV", "BFA")
for (cc in 1:3) {
  params <- sprintf("beta[%d,%d]", cc, 1:7)
  safe_print(fit_r2, params,
             label = paste0("Country ", cc, " (", country_labels[cc],
                            ") covariate effects (R2-only)"))
}
```



